---
title: "01_QualityControl"
author: "Chloe J. McGovern"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    css: "hpcstyles.css"
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = TRUE,
  fig.path   = "~/MIX-March2025/Figures/16S/01_QualityControl/",
  fig.width  = 12,
  fig.height = 8,
  dev        = "png",
  dpi        = 300
)
```

#Set Up Working Environment

## Load Packages
```{r libraries}
library(dada2)
library(tidyverse)
library(ShortRead)
library(Biostrings)
```


## Load Data
```{r paths}
project_dir <- "~/MIX-March2025"
raw_dir     <- file.path(project_dir, "Data/16S/raw_fastq")
trim_dir    <- file.path(project_dir, "Data/16S/trimmed")

stopifnot(dir.exists(raw_dir))
stopifnot(dir.exists(trim_dir))

raw_files    <- sort(list.files(raw_dir, pattern = "\\.fastq\\.gz$", full.names = TRUE))
sample_names <- sub("\\.fastq\\.gz$", "", basename(raw_files))

cat("Number of samples:", length(sample_names), "\n")
cat("Sample names:\n")
print(sample_names)
```

Yay! All of the files are here. 

# Quality Analysis

## Raw Quality Plots

```{r raw_quality_h2o, fig.width=12, fig.height=8}
# H2O negative controls
plotQualityProfile(raw_files[1:4])
```

```{r raw_quality_mix, fig.width=12, fig.height=8}
# MIX biological samples
plotQualityProfile(raw_files[5:9])   # MIX-1
plotQualityProfile(raw_files[10:14]) # MIX-2
plotQualityProfile(raw_files[15:19]) # MIX-3
```

## Primer Set Up

27F/1492R Kinnex full-length 16S primer set. Primers and their 
reverse complements must all be removed to prevent interference 
with DADA2 denoising (Callahan et al., 2016, *Nature Methods*).

```{r primers}
FWD <- "AGAGTTTGATCMTGGCTCAG"   # 27F
REV <- "TACGGYTACCTTGTTACGACTT" # 1492R

allOrients <- function(primer) {
  dna <- DNAString(primer)
  c(
    Forward    = toString(dna),
    Complement = toString(complement(dna)),
    Reverse    = toString(reverse(dna)),
    RevComp    = toString(reverseComplement(dna))
  )
}

FWD.orients <- allOrients(FWD)
REV.orients <- allOrients(REV)

cat("FWD primer orientations:\n"); print(FWD.orients)
cat("\nREV primer orientations:\n"); print(REV.orients)
```

## Primer Check

```{r primer_check}
primerHits <- function(primer, fn) {
  nhits <- vcountPattern(primer, sread(readFastq(fn)), fixed = FALSE)
  return(sum(nhits > 0))
}

# Check first biological sample (MIX-1-1)
cat("Primer hits in", sample_names[5], ":\n")
rbind(
  FWD.ForwardReads = primerHits(FWD.orients[["Forward"]], raw_files[5]),
  FWD.ReverseReads = primerHits(FWD.orients[["RevComp"]], raw_files[5]),
  REV.ForwardReads = primerHits(REV.orients[["Forward"]], raw_files[5]),
  REV.ReverseReads = primerHits(REV.orients[["RevComp"]], raw_files[5])
)
```

## Primer Trim

```{r cutadapt_trim}
cutadapt <- "/home/cjm423/.local/bin/cutadapt"
system2(cutadapt, args = "--version")

trim_files <- file.path(trim_dir, basename(raw_files))

trim_results <- vector("list", length(raw_files))

for (i in seq_along(raw_files)) {
  cat("Trimming:", sample_names[i], "\n")
  
  trim_results[[i]] <- system2(
    cutadapt,
    args = c(
      "-g", FWD,
      "-a", REV,
      "--rc",
      "--discard-untrimmed",
      "--minimum-length", "1000",
      "--maximum-length", "2000",
      "-o", trim_files[i],
      raw_files[i],
      "--quiet"
    ),
    stdout = TRUE,
    stderr = TRUE
  )
}

```

# Post Trim Analysis

## Post Trim Quality

```{r post_trim_quality, fig.width=12, fig.height=8}
# Verify trimmed files exist and are non-empty
trim_sizes <- file.info(trim_files)$size
trim_check <- data.frame(
  sample       = sample_names,
  raw_size_MB  = round(file.info(raw_files)$size / 1e6, 2),
  trim_size_MB = round(trim_sizes / 1e6, 2)
)
print(trim_check)

# Quality profiles post-trimming
plotQualityProfile(trim_files[1:4])   # H2O
plotQualityProfile(trim_files[5:9])   # MIX-1
plotQualityProfile(trim_files[10:14]) # MIX-2
plotQualityProfile(trim_files[15:19]) # MIX-3
```


## Post Trim Read Counts and Retention

```{r read_counts}
count_reads <- function(filepath) {
  length(readLines(gzcon(file(filepath, "rb")))) / 4
}

read_counts <- data.frame(
  sample     = sample_names,
  raw_reads  = sapply(raw_files, count_reads),
  trim_reads = sapply(trim_files, count_reads)
) %>%
  mutate(
    pct_retained = round(trim_reads / raw_reads * 100, 1),
    group = case_when(
      grepl("H2O",   sample, ignore.case = TRUE) ~ "H2O",
      grepl("MIX-1", sample, ignore.case = TRUE) ~ "MIX1",
      grepl("MIX-2", sample, ignore.case = TRUE) ~ "MIX2",
      grepl("MIX-3", sample, ignore.case = TRUE) ~ "MIX3"
    )
  )

print(read_counts)
```

```{r mix3_diagnostic}
library(ShortRead)

# Read length distribution for MIX-3-1 (worst retention) vs MIX-1-1 (normal)
mix3_reads <- readFastq(raw_files[15]) # MIX-3-1
mix1_reads <- readFastq(raw_files[5])  # MIX-1-1

mix3_lengths <- width(sread(mix3_reads))
mix1_lengths <- width(sread(mix1_reads))

cat("MIX-3-1 read length summary:\n")
print(summary(mix3_lengths))

cat("\nMIX-1-1 read length summary:\n")
print(summary(mix1_lengths))

# Primer hits in MIX-3-1
cat("\nPrimer hits in MIX-3-1:\n")
rbind(
  FWD.ForwardReads = primerHits(FWD.orients[["Forward"]], raw_files[15]),
  FWD.ReverseReads = primerHits(FWD.orients[["RevComp"]], raw_files[15]),
  REV.ForwardReads = primerHits(REV.orients[["Forward"]], raw_files[15]),
  REV.ReverseReads = primerHits(REV.orients[["RevComp"]], raw_files[15])
)
```

The read length distributions are essentially identical between MIX-3-1 and MIX-1-1 (median ~1526 bp both), so the reads themselves are fine. The problem is the primer hits. 74 primer hits across 139,000(ish) reads is marginal. Since *--discard-untrimmed* drops any read where primers aren't found, this directly explains the low retention.

Let's compare them to Mix 1

```{r mix3_primer_compare}
# Direct comparison — primer hits MIX-1-1 vs all MIX-3 samples
cat("Primer hits in MIX-1-1 (reference):\n")
rbind(
  FWD.ForwardReads = primerHits(FWD.orients[["Forward"]], raw_files[5]),
  FWD.ReverseReads = primerHits(FWD.orients[["RevComp"]], raw_files[5]),
  REV.ForwardReads = primerHits(REV.orients[["Forward"]], raw_files[5]),
  REV.ReverseReads = primerHits(REV.orients[["RevComp"]], raw_files[5])
)

# Check all MIX-3 samples
for (i in 15:19) {
  cat("\nPrimer hits in", sample_names[i], ":\n")
  print(rbind(
    FWD.ForwardReads = primerHits(FWD.orients[["Forward"]], raw_files[i]),
    FWD.ReverseReads = primerHits(FWD.orients[["RevComp"]], raw_files[i]),
    REV.ForwardReads = primerHits(REV.orients[["Forward"]], raw_files[i]),
    REV.ReverseReads = primerHits(REV.orients[["RevComp"]], raw_files[i])
  ))
}
```

***QC Decision Log***
  **Decision**
    1. H2O-5 retained through decontam
    2. MIX-3 low post-trim retention (32.7-47.2%)",
    3. MIX-3-4 low raw depth",
    4. Proceed with all 19 samples to DADA2

  **Rationale**
    1. Anomalously high raw reads (294,246) consistent with reagent contamination in non-sterile environment; retain for contaminant ASV identification (Davis et al., 2018, Microbiome)
    2. Read length distributions identical to MIX-1/2 (median ~1526 bp); pattern consistent across all 5 MIX-3 samples suggesting library prep differences; reads are high quality HiFi CCS
    3. 28,015 raw reads -> 9,728 post-trim; lowest biological sample in dataset; flag for rarefaction threshold assessment in Notebook 05
    4. No samples meet exclusion criteria at QC stage; all groups retain minimum n=5 replicates
  
  **Impact**
    1. H2O-5 excluded from diversity and DA analyses post-decontam
    2. MIX-3 samples monitored closely through DADA2 filtering and rarefaction; wet lab team notified of observation
    3. MIX-3-4 may fall below rarefaction threshold; assess in Notebook 05
    4. 19 samples forward to 02_dada2_processing.Rmd

# Session Information 

```{r session_info}
sessionInfo()
```


