---
title: "02_DADA2"
author: "Chloe J McGovern"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    css: "hpcstyles.css"
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  cache      = TRUE,
  fig.path   = "/home/cjm423/MIX-March2025/Figures/16S/02_DADA2/",
  fig.width  = 12,
  fig.height = 8,
  dev        = "png",
  dpi        = 300
)
```

## Overview

DADA2 amplicon sequence variant (ASV) inference from PacBio HiFi 
full-length 16S rRNA gene sequences (27F/1492R). Primer-trimmed reads 
from `01_QualityControl.Rmd` are processed through the DADA2 pipeline 
(Callahan et al., 2016, *Nature Methods*) including quality filtering, 
error modeling, denoising, and chimera removal.

## Libraries
```{r libraries}
library(dada2)
library(tidyverse)
```

## File Paths
```{r paths}
project_dir <- "/home/cjm423/MIX-March2025"

# Input — trimmed FASTQs from 01_QualityControl
trim_dir <- file.path(project_dir, "Data/16S/trimmed")

# Output — processed data objects
data_dir <- file.path(project_dir, "Data/16S")

# Figures
fig_dir <- file.path(project_dir, "Figures/16S/02_DADA2")
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)

# SILVA reference — symlinked from LPHY project
silva_train <- "/home/cjm423/LPHY-March2025/data/16S/reference/silva_nr99_v138.1_train_set.fa.gz"

# Confirm reference exists
stopifnot(file.exists(silva_train))

# List trimmed files
trim_files <- sort(list.files(trim_dir,
                               pattern = "\\.fastq\\.gz$",
                               full.names = TRUE))

# Sample names
sample_names <- sub("\\.fastq\\.gz$", "", basename(trim_files))

cat("Samples detected:", length(sample_names), "\n")
print(sample_names)
```

## Sample Metadata
```{r metadata}
# Construct metadata for MIX caecal samples
metadata <- data.frame(
  sample_id  = sample_names,
  file       = basename(trim_files),
  group      = case_when(
    grepl("H2O",   sample_names, ignore.case = TRUE) ~ "H2O",
    grepl("MIX-1", sample_names, ignore.case = TRUE) ~ "MIX1",
    grepl("MIX-2", sample_names, ignore.case = TRUE) ~ "MIX2",
    grepl("MIX-3", sample_names, ignore.case = TRUE) ~ "MIX3"
  ),
  is_control = grepl("H2O", sample_names, ignore.case = TRUE),
  tissue     = "caecum",
  sequencer  = "PacBio_HiFi",
  primers    = "27F_1492R_Kinnex"
) %>%
  mutate(
    replicate = as.integer(str_extract(sample_id, "(?<=-)[0-9](?=\\.)")),
  )

# Verify no NA groups
if (any(is.na(metadata$group))) {
  warning("NA group assignments detected for: ",
          paste(metadata$sample_id[is.na(metadata$group)], collapse = ", "))
} else {
  cat("All samples assigned to groups successfully\n")
}

# Save metadata
write.csv(metadata,
          file.path(data_dir, "MIX_metadata.csv"),
          row.names = FALSE)

print(metadata)
```

## Quality Filtering

Filter and trim reads using DADA2. For PacBio HiFi full-length 16S,
truncation is not applied since reads are already high quality and
uniform length. We filter on `maxEE` (maximum expected errors) which
is the recommended approach for HiFi data (Callahan et al., 2016,
*Nature Methods*).

```{r filter_trim}
# Output paths for filtered reads
filt_dir <- file.path(project_dir, "Data/16S/filtered")
if (!dir.exists(filt_dir)) dir.create(filt_dir, recursive = TRUE)

filt_files <- file.path(filt_dir, basename(trim_files))
names(filt_files) <- sample_names

# Filter parameters for PacBio HiFi full-length 16S — mirrors LPHY-March2025
out <- filterAndTrim(
  trim_files, filt_files,
  minLen      = 1000,
  maxLen      = 1600,
  maxEE       = 2,
  maxN        = 0,
  compress    = TRUE,
  multithread = TRUE,
  verbose     = TRUE
)

# View filtering results
out_df <- as.data.frame(out) %>%
  mutate(
    sample       = sample_names,
    pct_retained = round(reads.out / reads.in * 100, 1)
  ) %>%
  select(sample, reads.in, reads.out, pct_retained)

print(out_df)
```

## Error Learning

DADA2 learns the error model from the data. For PacBio HiFi, error
rates are very low and consistent, so error learning converges quickly
(Callahan et al., 2016, *Nature Methods*).
```{r learn_errors, fig.width=10, fig.height=6}
# Learn error rates — use all available cores
set.seed(100)
err <- learnErrors(
  filt_files,
  multithread = TRUE,
  errorEstimationFunction = PacBioErrfun,  # PacBio-specific error function
  BAND_SIZE = 32,                          
  verbose = TRUE
)

# Visualize error model — points should follow fitted line
plotErrors(err, nominalQ = TRUE)
```


Black fitted curve follows the observed points (grey dots) reasonably well, with the characteristic U-shape at high Q-scores typical of HiFi CCS data. Red and black line divergence expected with PacBio sequencing. *PacBioErrfun* accounts for nonlinear error structure rather than assuming Illumina style monotonic decrease. 

## Sample Inference — DADA2 Denoising
```{r dada2_denoise}
# Denoise each sample independently
dada_results <- dada(
  filt_files,
  err         = err,
  multithread = TRUE,
  BAND_SIZE   = 32,
  pool        = FALSE
)

# Summary of denoising
dada_summary <- sapply(dada_results, function(x)
  c(ASVs = x$denoised, reads = sum(x$denoised)))

cat("Denoising summary:\n")
print(dada_summary)
```

## Construct ASV Table
```{r asv_table}
# Make sequence table
seqtab <- makeSequenceTable(dada_results)

# Inspect ASV length distribution
seq_lengths <- table(nchar(getSequences(seqtab)))
cat("ASV length distribution:\n")
print(seq_lengths)

# Plot length distribution
seq_len_df <- data.frame(
  length = as.integer(names(seq_lengths)),
  count  = as.integer(seq_lengths)
)

ggplot(seq_len_df, aes(x = length, y = count)) +
  geom_col(fill = "violetred") +
  labs(
    title = "ASV Length Distribution — MIX-March2025",
    x     = "Sequence Length (bp)",
    y     = "Number of ASVs"
  ) +
  theme_bw()
```

## Remove Chimeras
```{r chimera_removal}
# Remove bimeric sequences
seqtab.nochim <- removeBimeraDenovo(
  seqtab,
  method      = "consensus",
  multithread = TRUE,
  verbose     = TRUE
)

# Chimera summary
cat("ASVs before chimera removal:", ncol(seqtab), "\n")
cat("ASVs after chimera removal:", ncol(seqtab.nochim), "\n")
cat("Reads retained:",
    round(sum(seqtab.nochim) / sum(seqtab) * 100, 1), "%\n")

# Save seqtab
saveRDS(seqtab.nochim,
        file.path(data_dir, "seqtab_nochim.rds"))

# Confirm final ASV table dimensions
dim(seqtab.nochim)  # should be 19 samples x 118 ASVs

# Per-sample read depth going into decontam — this is what matters most right now
rowSums(seqtab.nochim)

# Sequence length distribution — should cluster tightly around expected full-length 16S
table(nchar(getSequences(seqtab.nochim)))
```

## Taxonomic Assignment — SILVA r138.1

Taxonomic classification using the SILVA r138.1 reference database
(Quast et al., 2013, *Nucleic Acids Research*) with DADA2's naive
Bayesian classifier (Wang et al., 2007, *Applied and Environmental
Microbiology*).

```{r taxonomy}
# Assign taxonomy to genus level
taxa <- assignTaxonomy(
  seqtab.nochim,
  silva_train,
  multithread = TRUE,
  verbose     = TRUE
)

# Save taxonomy table
saveRDS(taxa, file.path(data_dir, "taxa.rds"))

# Preview taxonomy — mask sequence rownames for readability
taxa_preview <- taxa
rownames(taxa_preview) <- NULL
head(taxa_preview, 20)
```

## Read Tracking Table

Track reads through every step of the pipeline — critical for QC
and reporting (Callahan et al., 2016, *Nature Methods*).

```{r read_tracking}
source("/home/cjm423/MIX-March2025/Code/colors.R")

getN <- function(x) sum(getUniques(x))

track <- data.frame(
  sample   = sample_names,
  raw      = out_df$reads.in,
  filtered = out_df$reads.out,
  denoised = sapply(dada_results, getN),
  nonchim  = rowSums(seqtab.nochim)
) %>%
  mutate(
    pct_final = round(nonchim / raw * 100, 1),
    group     = metadata$group
  )

print(track)

# Save tracking table
write.csv(track,
          file.path(data_dir, "read_tracking.csv"),
          row.names = FALSE)

# Visualize read retention
track_long <- track %>%
  select(sample, group, raw, filtered, denoised, nonchim) %>%
  pivot_longer(cols = c(raw, filtered, denoised, nonchim),
               names_to  = "step",
               values_to = "reads") %>%
  mutate(step = factor(step,
                       levels = c("raw", "filtered",
                                  "denoised", "nonchim")))

ggplot(track_long, aes(x = step, y = reads,
                        group = sample, color = group)) +
  geom_line(alpha = 0.7) +
  geom_point(size = 2) +
  scale_color_manual(values = c(mix_colors, "H2O" = "grey60")) +
  labs(
    title = "Read Retention Through DADA2 Pipeline — MIX",
    x     = "Pipeline Step",
    y     = "Read Count",
    color = "Group"
  ) +
  theme_bw()
```

MIX-3-2: 36.4% final retention despite being the highest raw depth sample (437k raw → 206k trimmed → 75k nonchim). The drop from denoised (113,834) to nonchim (75,337) is substantial — 33% of reads lost at chimera removal for this one sample alone. 

MIX-3-4: 9,728 trimmed → 5,009 nonchim. Very low absolute depth — almost certainly below rarefaction threshold. Flag for notebook 05

H2O-1: Only 32 nonchim reads — essentially empty, as expected for a true negative control

H2O-5: 47,673 nonchim reads — confirms it carries real biological signal, reinforcing the decision to retain it through decontam

## Session Info
```{r session_info}
sessionInfo()
```
