---
title: "03_Taxonomy"
author: "Chloe J McGovern"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    css: "hpcstyles.css"
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  cache      = TRUE,
  fig.path   = "/home/cjm423/MIX-March2025/Figures/16S/03_Taxonomy/",
  fig.width  = 12,
  fig.height = 8,
  dev        = "png",
  dpi        = 300
)
```

## Overview

Taxonomic composition analysis of PacBio HiFi full-length 16S rRNA 
gene amplicon data from MIX caecal samples. ASVs were assigned to 
genus level using the SILVA r138.1 reference database (Quast et al., 
2013, *Nucleic Acids Research*). Composition is visualized at phylum 
and genus level prior to decontamination.

## Libraries
```{r libraries}
library(dada2)
library(phyloseq)
library(tidyverse)
library(patchwork)

# Source project color scheme
source("~/MIX-March2025/Code/colors.R")
```

## Load Data
```{r load_data}
project_dir <- "~/MIX-March2025"
data_dir    <- file.path(project_dir, "Data/16S")

# Load DADA2 outputs from 02_DADA2.Rmd
seqtab <- readRDS(file.path(data_dir, "seqtab_nochim.rds"))
taxa   <- readRDS(file.path(data_dir, "taxa.rds"))

# Load metadata
metadata <- read.csv(file.path(data_dir, "MIX_metadata.csv"),
                     row.names = 1)

cat("Samples:", nrow(seqtab), "\n")
cat("ASVs:", ncol(seqtab), "\n")
cat("Taxa rows:", nrow(taxa), "\n")
cat("Metadata sample_ids:\n")
print(metadata$sample_id)
```

## Build Phyloseq Object
```{r build_phyloseq}
# Align sample names between seqtab and metadata
rownames(seqtab) <- sub("\\.fastq\\.gz$", "", rownames(seqtab))

# Check rownames
cat("seqtab rownames:\n")
print(rownames(seqtab))

cat("\nmetadata colnames:\n")
print(colnames(metadata))

cat("\nmetadata head:\n")
print(head(metadata))

# Add sample_id as proper column if needed
metadata <- metadata %>%
  rownames_to_column(var = "sample_id")

# Verify
print(head(metadata))
print(colnames(metadata))

# Overwrite CSV with sample_id as proper column
write.csv(metadata,
          file.path(data_dir, "MIX_metadata.csv"),
          row.names = FALSE)

# Confirm rownames match
stopifnot(all(rownames(seqtab) %in% metadata$sample_id))

# Set metadata rownames for phyloseq
meta_ps <- metadata
rownames(meta_ps) <- meta_ps$sample_id

# Build phyloseq object
# McMurdie & Holmes 2013, PLOS ONE
ps <- phyloseq(
  otu_table(seqtab, taxa_are_rows = FALSE),
  tax_table(taxa),
  sample_data(meta_ps)
)

# Replace ASV sequences with short IDs
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))

cat("Phyloseq object:\n")
print(ps)
```

## Overview of Taxonomic Assignment
```{r taxa_overview}
tax_df <- as.data.frame(tax_table(ps))

classification_rates <- data.frame(
  rank    = c("Kingdom", "Phylum", "Class", "Order",
              "Family", "Genus"),
  assigned = c(
    sum(!is.na(tax_df$Kingdom)),
    sum(!is.na(tax_df$Phylum)),
    sum(!is.na(tax_df$Class)),
    sum(!is.na(tax_df$Order)),
    sum(!is.na(tax_df$Family)),
    sum(!is.na(tax_df$Genus))
  ),
  total = nrow(tax_df)
) %>%
  mutate(pct_assigned = round(assigned / total * 100, 1))

print(classification_rates)
```

## Phylum-Level Composition
```{r phylum_composition}
# Aggregate to phylum level
ps_phylum <- tax_glom(ps, taxrank = "Phylum", NArm = FALSE)

# Convert to relative abundance
ps_phylum_rel <- transform_sample_counts(ps_phylum,
                                          function(x) x / sum(x))

# Melt to long format
phylum_long <- psmelt(ps_phylum_rel) %>%
  mutate(
    Phylum = ifelse(is.na(Phylum), "Unclassified", Phylum),
    group  = factor(group, levels = c("H2O", "MIX1", "MIX2", "MIX3"))
  )

# Keep top 10 phyla, collapse rest to "Other"
top_phyla <- phylum_long %>%
  group_by(Phylum) %>%
  summarise(mean_abund = mean(Abundance)) %>%
  arrange(desc(mean_abund)) %>%
  slice_head(n = 10) %>%
  pull(Phylum)

phylum_long <- phylum_long %>%
  mutate(Phylum = ifelse(Phylum %in% top_phyla, Phylum, "Other"))

ggplot(phylum_long, aes(x = Sample, y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  facet_grid(~ group, scales = "free_x", space = "free") +
  scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Phylum-Level Relative Abundance",
    x     = "Sample",
    y     = "Relative Abundance",
    fill  = "Phylum"
  ) +
  theme_bw() +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 7),
    legend.position = "right"
  )
```

## Genus-Level Composition
```{r genus_composition, fig.width=8, fig.height=6, dpi=300}
# Aggregate to genus level
ps_genus <- tax_glom(ps, taxrank = "Genus", NArm = FALSE)

# Convert to relative abundance
ps_genus_rel <- transform_sample_counts(ps_genus,
                                         function(x) x / sum(x))

# Melt to long format
genus_long <- psmelt(ps_genus_rel) %>%
  mutate(
    Genus = ifelse(is.na(Genus), "Unclassified", Genus),
    group = factor(group, levels = c("H2O", "MIX1", "MIX2", "MIX3"))
  )

# Keep top 15 genera
top_genera <- genus_long %>%
  group_by(Genus) %>%
  summarise(mean_abund = mean(Abundance)) %>%
  arrange(desc(mean_abund)) %>%
  slice_head(n = 15) %>%
  pull(Genus)

genus_long <- genus_long %>%
  mutate(Genus = ifelse(Genus %in% top_genera, Genus, "Other"))

ggplot(genus_long, aes(x = Sample, y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity") +
  facet_grid(~ group, scales = "free_x", space = "free") +
  scale_fill_manual(values = colorRampPalette(
    RColorBrewer::brewer.pal(12, "Paired"))(16)) +
  labs(
    title = "Genus-Level Relative Abundance",
    x     = "Sample",
    y     = "Relative Abundance",
    fill  = "Genus"
  ) +
  theme_bw() +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 10),
    legend.position = "right",
    text            = element_text(family = "Times New Roman"),
    plot.title      = element_text(family = "Times New Roman"),
    axis.title      = element_text(family = "Times New Roman"),
    legend.text     = element_text(family = "Times New Roman"),
    legend.title    = element_text(family = "Times New Roman"),
    strip.text      = element_text(family = "Times New Roman")
  )
```

## Mean Relative Abundance by Treatment Group
```{r group_composition, fig.height=10}
# Summarize genus abundance by treatment group
# Excludes H2O controls
genus_group <- genus_long %>%
  filter(group != "H2O") %>%
  group_by(group, Genus) %>%
  summarise(mean_abund = mean(Abundance), .groups = "drop")

ggplot(genus_group, aes(x = group, y = mean_abund, fill = group)) +
  geom_col() +
  facet_wrap(~ Genus, scales = "free_y", ncol = 4) +
  scale_fill_manual(values = mix_colors) +
  labs(
    title = "Mean Genus Abundance by MIX Treatment Group",
    x     = "Treatment",
    y     = "Mean Relative Abundance",
    fill  = "Group"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Save Phyloseq Object
```{r save_phyloseq}
saveRDS(ps, file.path(data_dir, "ps_predecontam.rds"))
cat("Phyloseq object saved to Data/16S/ps_predecontam.rds\n")
```

###Check same ASV for lactiplanti before decontam of water control and treatments

```{r check-before-decon}
lacto_asvs <- taxa_names(ps)[!is.na(tax_table(ps)[, "Genus"]) & tax_table(ps)[, "Genus"] == "Lactiplantibacillus"]
length(lacto_asvs)  # should be 15

otu_table(ps)[, lacto_asvs] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("sample") %>%
  pivot_longer(-sample, names_to = "ASV", values_to = "reads") %>%
  left_join(as(sample_data(ps), "data.frame") %>% tibble::rownames_to_column("sample"),
             by = "sample") %>%
  group_by(ASV, group) %>%
  summarise(mean_reads = mean(reads), .groups = "drop") %>%
  filter(mean_reads > 0) %>%
  pivot_wider(names_from = group, values_from = mean_reads, values_fill = 0)
```

13 of 15 Lactiplantibacillus ASVs (ASV1–ASV14, excluding the two singleton-driven ASV75/ASV77) show up in every group, including H2O

## Session Info
```{r session_info}
sessionInfo()
```
