---
title: "04_Decontam"
author: "Chloe J McGovern"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    css: "hpcstyles.css"
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  cache      = TRUE,
  fig.path   = "/home/cjm423/MIX-March2025/Figures/16S/04_Decontam/",
  fig.width  = 12,
  fig.height = 8,
  dev        = "png",
  dpi        = 300
)
```

## Overview

Identification and removal of contaminant ASVs using the `decontam` 
package (Davis et al., 2018, *Microbiome*). Contaminants are identified 
using the prevalence-based method, which flags ASVs that occur more 
frequently in negative controls (H2O) than in biological samples. 
H2O-5 is retained for contaminant identification but excluded from 
all downstream analyses per QC decisions documented in notebook 01.

## Libraries
```{r libraries}
library(phyloseq)
#if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install("decontam")
library(decontam)
library(tidyverse)
library(patchwork)

source("/home/cjm423/MIX-March2025/Code/colors.R")
```

## Load Data
```{r load_data}
project_dir <- "~/MIX-March2025"
data_dir    <- file.path(project_dir, "Data/16S")

# Load pre-decontam phyloseq from 03_Taxonomy.Rmd
ps <- readRDS(file.path(data_dir, "ps_predecontam.rds"))

cat("Pre-decontam phyloseq:\n")
print(ps)

# Confirm negative controls are flagged correctly
cat("\nNegative control samples:\n")
print(sample_data(ps)$is_control)
```

## Inspect Library Sizes

Library size is used by decontam's prevalence method to identify
contaminants — contaminants tend to be more prevalent in low-biomass
negative controls (Davis et al., 2018, *Microbiome*).
```{r library_sizes}
# Add library size to sample data
sample_data(ps)$lib_size <- sample_sums(ps)

lib_df <- data.frame(sample_data(ps)) %>%
  mutate(
    sample     = rownames(.),
    is_control = as.logical(is_control)
  ) %>%
  arrange(lib_size)

ggplot(lib_df, aes(x = reorder(sample, lib_size),
                    y = lib_size,
                    fill = is_control)) +
  geom_col() +
  scale_fill_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "tomato"),
    labels = c("Biological", "Negative Control")
  ) +
  labs(
    title = "Library Size per Sample — MIX-March2025",
    x     = "Sample",
    y     = "Read Count",
    fill  = "Sample Type"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    text        = element_text(family = "Times New Roman")
  )
```

## Identify Contaminants — Prevalence Method

The prevalence method compares the prevalence of each ASV in 
negative controls vs biological samples. ASVs with a decontam 
score > 0.5 are flagged as contaminants (Davis et al., 2018, 
*Microbiome*).
```{r identify_contaminants}
# Run decontam prevalence method
contamdf <- isContaminant(
  ps,
  method    = "prevalence",
  neg       = "is_control",
  threshold = 0.5
)

# Summary
cat("Total ASVs:", nrow(contamdf), "\n")
cat("Contaminant ASVs identified:", sum(contamdf$contaminant), "\n")
cat("Non-contaminant ASVs:", sum(!contamdf$contaminant), "\n")

# View contaminant ASVs with taxonomy
contam_asvs <- rownames(contamdf)[contamdf$contaminant]
cat("\nContaminant ASV taxonomy:\n")
print(tax_table(ps)[contam_asvs, ])
```

## Visualize Contaminant Prevalence
```{r contaminant_plot}
# Validate phyloseq object
head(taxa_names(ps))
head(rownames(tax_table(ps)))
head(colnames(otu_table(ps)))
class(sample_data(ps)$is_control)
validObject(ps)

# Presence/absence transform
ps_pa <- transform_sample_counts(ps, function(x) 1*(x > 0))

# Split into controls and biological samples
ps_pa_neg <- prune_samples(
  sample_data(ps_pa)$is_control == TRUE, ps_pa)
ps_pa_pos <- prune_samples(
  sample_data(ps_pa)$is_control == FALSE, ps_pa)

# Build prevalence dataframe
prev_df <- data.frame(
  pa_neg      = taxa_sums(ps_pa_neg),
  pa_pos      = taxa_sums(ps_pa_pos),
  contaminant = contamdf$contaminant
)

ggplot(prev_df, aes(x = pa_neg, y = pa_pos, color = contaminant)) +
  geom_point(alpha = 0.7, size = 3) +
  scale_color_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "tomato"),
    labels = c("Non-contaminant", "Contaminant")
  ) +
  labs(
    title = "Decontam Prevalence Plot",
    x     = "Prevalence in Negative Controls",
    y     = "Prevalence in Biological Samples",
    color = "Classification"
  ) +
  theme_bw() +
  theme(text = element_text(family = "Times New Roman"))
```

## Remove Contaminants and Negative Controls
```{r remove_contaminants}
# Remove contaminant ASVs
ps_decontam <- prune_taxa(!contamdf$contaminant, ps)

cat("ASVs before decontam:", ntaxa(ps), "\n")
cat("ASVs after decontam:", ntaxa(ps_decontam), "\n")

# Remove all H2O negative control samples
ps_clean <- prune_samples(
  sample_data(ps_decontam)$is_control == FALSE,
  ps_decontam)

# Remove ASVs now zero across all biological samples
ps_clean <- prune_taxa(taxa_sums(ps_clean) > 0, ps_clean)

cat("\nPost-decontam phyloseq (biological samples only):\n")
print(ps_clean)

sort(sample_sums(ps_clean)) #MIX-3-4 at 5,009 sits well below everything else, investigate rarefaction curves to see plateau before deciding to drop a replicate


```

## Compare Composition Pre vs Post Decontam
```{r pre_post_comparison, fig.height=10}
# Biological samples only, pre-decontam
ps_bio <- prune_samples(
  sample_data(ps)$is_control == FALSE, ps)

ps_bio_rel   <- transform_sample_counts(ps_bio,
                                         function(x) x / sum(x))
ps_clean_rel <- transform_sample_counts(ps_clean,
                                         function(x) x / sum(x))

pre_long <- psmelt(ps_bio_rel) %>%
  mutate(status = "Pre-decontam",
         Phylum = ifelse(is.na(Phylum), "Unclassified", Phylum))

post_long <- psmelt(ps_clean_rel) %>%
  mutate(status = "Post-decontam",
         Phylum = ifelse(is.na(Phylum), "Unclassified", Phylum))

combined <- bind_rows(pre_long, post_long) %>%
  mutate(
    status = factor(status, levels = c("Pre-decontam", "Post-decontam")),
    group  = factor(group, levels = c("MIX1", "MIX2", "MIX3"))
  )

top_phyla <- combined %>%
  group_by(Phylum) %>%
  summarise(mean_abund = mean(Abundance)) %>%
  arrange(desc(mean_abund)) %>%
  slice_head(n = 10) %>%
  pull(Phylum)

combined <- combined %>%
  mutate(Phylum = ifelse(Phylum %in% top_phyla, Phylum, "Other"))

ggplot(combined, aes(x = Sample, y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  facet_grid(status ~ group, scales = "free_x", space = "free") +
  scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Phylum Composition Pre vs Post Decontam",
    x     = "Sample",
    y     = "Relative Abundance",
    fill  = "Phylum"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
    text        = element_text(family = "Times New Roman")
  )
```

## Save Clean Phyloseq Object
```{r save_clean}
saveRDS(ps_clean, file.path(data_dir, "ps_clean.rds"))

cat("Clean phyloseq saved to Data/16S/ps_clean.rds\n")
cat("Samples:", nsamples(ps_clean), "\n")
cat("ASVs:", ntaxa(ps_clean), "\n")

# Save contaminant ASV list for methods reporting
write.csv(
  as.data.frame(tax_table(ps)[contam_asvs, ]),
  file.path(data_dir, "contaminant_asvs.csv"),
  row.names = TRUE
)
cat("Contaminant ASV list saved to Data/16S/contaminant_asvs.csv\n")
```

## Session Info
```{r session_info}
sessionInfo()
```