---
title: "05_Phyloseq"
author: "Chloe J McGovern"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    css: "hpcstyles.css"
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  cache      = TRUE,
  fig.path   = "/home/cjm423/MIX-March2025/Figures/16S/05_Phyloseq/",
  fig.width  = 12,
  fig.height = 8,
  dev        = "png",
  dpi        = 300
)
```

## Overview

Alpha and beta diversity analysis of MIX caecal microbiome data 
following decontamination. Diversity metrics are calculated on 
rarefied data to account for uneven sequencing depth (McMurdie & 
Holmes, 2014, *PLOS Computational Biology*). Comparisons are made 
between MIX treatment groups (MIX1, MIX2, MIX3).

Note: MIX-3-4 (9,728 post-trim reads) is flagged as potentially 
below rarefaction threshold — rarefaction depth will be determined 
empirically from rarefaction curves.

## Libraries

```{r libraries}
library(phyloseq)
library(vegan)
library(tidyverse)
library(patchwork)

source("/home/cjm423/MIX-March2025/Code/colors.R")

# MIX group colors
mix_colors
```

## Load Data
```{r load_data}
project_dir <- "/home/cjm423/MIX-March2025"
data_dir    <- file.path(project_dir, "Data/16S")

# Load clean phyloseq from 04_Decontam.Rmd
ps_clean <- readRDS(file.path(data_dir, "ps_clean.rds"))

cat("Clean phyloseq:\n")
print(ps_clean)

# Check sample depths
cat("\nSample read depths:\n")
sort(sample_sums(ps_clean))
```

## Rarefaction Curves

Rarefaction curves confirm whether sequencing depth is sufficient 
to capture community diversity (Gotelli & Colwell, 2001, 
*Ecology Letters*). Rarefaction depth will be set at the point 
where all retained samples reach asymptote.
```{r rarefaction_curves}
# Build rarefaction curve data
rare_df <- rarecurve(
  as.data.frame(otu_table(ps_clean)),
  step    = 500,
  sample  = min(sample_sums(ps_clean)),
  label   = FALSE,
  tidy    = TRUE
)

# Add group metadata
sample_meta <- data.frame(sample_data(ps_clean)) %>%
  rownames_to_column("Site")

rare_df <- rare_df %>%
  left_join(sample_meta, by = "Site")

# Plot
ggplot(rare_df, aes(x = Sample, y = Species,
                     group = Site, color = group)) +
  geom_line(alpha = 0.8) +
  scale_color_manual(values = mix_colors) +
  labs(
    title = "Rarefaction Curves",
    x     = "Reads Sampled",
    y     = "ASVs Observed",
    color = "Treatment Group"
  ) +
  theme_bw() +
  theme(text = element_text(family = "Times New Roman", size = 14))

# Check exact sample depths to inform rarefaction threshold
sort(sample_sums(ps_clean))
```

## Rarefy
```{r rarefy}
# Rarefaction depth is 5009 because every curve is already flat well before 5,009 reads
set.seed(123)

ps_rare <- rarefy_even_depth(
  ps_clean,
  sample.size = 5009,
  rngseed     = 123,
  replace     = FALSE,
  trimOTUs    = TRUE,
  verbose     = TRUE
)

cat("\nPost-rarefaction phyloseq:\n")
print(ps_rare)
cat("\nSamples retained:", nsamples(ps_rare), "\n")
cat("ASVs retained:", ntaxa(ps_rare), "\n")
```

## Alpha Diversity

Alpha diversity measures within-sample diversity. Shannon entropy 
accounts for both richness and evenness, while Observed ASVs 
measures richness only (Hill, 1973, *Ecology*).
```{r alpha_diversity, fig.height=10}
# Calculate alpha diversity metrics
rich <- estimate_richness(
  ps_rare,
  measures = c("Observed", "Shannon", "Simpson")
)

# Map back to original sample names via position
meta_df <- data.frame(sample_data(ps_rare))

alpha_df <- rich %>%
  mutate(
    sample_id = meta_df$sample_id,
    group     = meta_df$group,
    replicate = meta_df$replicate
  ) %>%
  mutate(group = factor(group, levels = c("MIX1", "MIX2", "MIX3")))

print(alpha_df)

# Save alpha diversity table
write.csv(alpha_df,
          file.path(data_dir, "alpha_diversity.csv"),
          row.names = FALSE)

# Plot observed ASVs
p1 <- ggplot(alpha_df, aes(x = group, y = Observed, fill = group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.8) +
  scale_fill_manual(values = mix_colors) +
  labs(title = "Richness", x = "Treatment", y = "ASVs") +
  theme_bw() +
  theme(
    legend.position = "none",
    text = element_text(family = "Times New Roman", size = 16)
  )

# Plot Shannon diversity
p2 <- ggplot(alpha_df, aes(x = group, y = Shannon, fill = group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.8) +
  scale_fill_manual(values = mix_colors) +
  labs(title = "Shannon Diversity", x = "Treatment",
       y = "Shannon Index") +
  theme_bw() +
  theme(
    legend.position = "none",
    text = element_text(family = "Times New Roman", size = 16)
  )

# Plot Simpson diversity
p3 <- ggplot(alpha_df, aes(x = group, y = Simpson, fill = group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.8) +
  scale_fill_manual(values = mix_colors) +
  labs(title = "Simpson Diversity", x = "Treatment",
       y = "Simpson Index") +
  theme_bw() +
  theme(
    legend.position = "none",
    text = element_text(family = "Times New Roman", size = 16)
  )

p1 + p2 + p3
```

## Alpha Diversity Statistics

Kruskal-Wallis test for differences in alpha diversity between 
treatment groups, followed by pairwise Wilcoxon tests with 
Benjamini-Hochberg correction (Benjamini & Hochberg, 1995, 
*Journal of the Royal Statistical Society*).
```{r alpha_stats}
metrics <- c("Observed", "Shannon", "Simpson")

alpha_stats <- lapply(metrics, function(m) {
  kw <- kruskal.test(reformulate("group", response = m),
                     data = alpha_df)
  data.frame(
    metric  = m,
    H       = round(kw$statistic, 3),
    df      = kw$parameter,
    p_value = round(kw$p.value, 4)
  )
}) %>% bind_rows()

print(alpha_stats)

# Pairwise Wilcoxon with BH correction for each metric
for (m in metrics) {
  cat("\nPairwise Wilcoxon —", m, ":\n")
  print(pairwise.wilcox.test(
    alpha_df[[m]],
    alpha_df$group,
    p.adjust.method = "BH"
  ))
}
```

## Beta Diversity — Bray-Curtis Ordination

Beta diversity measures between-sample diversity. Bray-Curtis 
dissimilarity is visualized using NMDS ordination (Clarke, 1993, 
*Australian Journal of Ecology*).
```{r beta_diversity_braycurtis}
# Calculate Bray-Curtis dissimilarity
bc_dist <- phyloseq::distance(ps_rare, method = "bray")

# NMDS ordination
set.seed(123)
nmds_bc <- ordinate(ps_rare, method = "NMDS", distance = bc_dist)

cat("NMDS stress:", nmds_bc$stress, "\n")

# Plot NMDS
plot_ordination(ps_rare, nmds_bc, color = "group") +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = mix_colors) +
  stat_ellipse(aes(group = group), level = 0.95) +
  labs(
    title = paste0("Bray-Curtis NMDS (stress = ",
                   round(nmds_bc$stress, 3), ")"),
    color = "Treatment Group"
  ) +
  theme_bw() +
  theme(
    legend.position = "right",
    text = element_text(family = "Times New Roman", size = 16)
  )

summary(bc_dist)
 hist(bc_dist, breaks = 20)
# Or as a matrix, to see the actual range
as.matrix(bc_dist) %>% round(3)
 
bc_mat <- as.matrix(bc_dist)

# Pull out all pairs with distance > 0.85
high_dist_pairs <- which(bc_mat > 0.85 & upper.tri(bc_mat), arr.ind = TRUE)
data.frame(
  sample1 = rownames(bc_mat)[high_dist_pairs[, 1]],
  sample2 = colnames(bc_mat)[high_dist_pairs[, 2]],
  distance = bc_mat[high_dist_pairs]
) %>% arrange(desc(distance))

#see if removing two outliers (n=13) changes NMDS 
ps_no_outliers <- subset_samples(ps_rare, !(sample_names(ps_rare) %in% c("MIX-2-5.4_43", "MIX-3-2.5_37")))
nsamples(ps_no_outliers)  # should be 13

bc_dist_v2 <- phyloseq::distance(ps_no_outliers, method = "bray")
set.seed(123)
nmds_bc_v2 <- ordinate(ps_no_outliers, method = "NMDS", distance = bc_dist_v2)
cat("NMDS stress (outliers excluded):", nmds_bc_v2$stress, "\n")

summary(bc_dist_v2)
hist(as.numeric(bc_dist_v2), breaks = 20)

```

## PERMANOVA — Bray-Curtis

PERMANOVA tests whether microbial community composition differs 
significantly between treatment groups (Anderson, 2001, 
*Australian Ecology*).
```{r permanova_bc}
otu_mat <- as.data.frame(otu_table(ps_rare))
meta_df <- data.frame(sample_data(ps_rare))

set.seed(123)
permanova_bc <- adonis2(
  bc_dist ~ group,
  data         = meta_df,
  permutations = 999
)

print(permanova_bc)

# Beta dispersion test
betadisp <- betadisper(bc_dist, meta_df$group)
permutest(betadisp, permutations = 999)


meta_df_v2 <- as(sample_data(ps_no_outliers), "data.frame")

set.seed(123)
permanova_bc_v2 <- adonis2(bc_dist_v2 ~ group, data = meta_df_v2, permutations = 999)
print(permanova_bc_v2)

betadisp_v2 <- betadisper(bc_dist_v2, meta_df_v2$group)
permutest(betadisp_v2, permutations = 999)
```

## Beta Diversity — PCoA
```{r pcoa}
pcoa_bc <- ordinate(ps_rare, method = "PCoA", distance = bc_dist)

eig <- pcoa_bc$values$Eigenvalues
var_explained <- round(eig / sum(eig) * 100, 1)

plot_ordination(ps_rare, pcoa_bc, color = "group") +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = mix_colors) +
  stat_ellipse(aes(group = group), level = 0.95) +
  labs(
    title = "Bray-Curtis PCoA",
    x     = paste0("PC1 (", var_explained[1], "%)"),
    y     = paste0("PC2 (", var_explained[2], "%)"),
    color = "Treatment Group"
  ) +
  theme_bw() +
  theme(
    legend.position = "right",
    text = element_text(family = "Times New Roman", size = 16)
  )

#n13 pcoa plot

pcoa_bc_v2 <- ordinate(ps_no_outliers, method = "PCoA", distance = bc_dist_v2)
eig_v2 <- pcoa_bc_v2$values$Eigenvalues
var_explained_v2 <- round(eig_v2 / sum(eig_v2) * 100, 1)

plot_ordination(ps_no_outliers, pcoa_bc_v2, color = "group") +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = mix_colors) +
  stat_ellipse(aes(group = group), level = 0.95) +
  labs(
    title = "Bray-Curtis PCoA (outliers excluded)",
    x     = paste0("PC1 (", var_explained_v2[1], "%)"),
    y     = paste0("PC2 (", var_explained_v2[2], "%)"),
    color = "Treatment Group"
  ) +
  theme_bw() +
  theme(
    legend.position = "right",
    text = element_text(family = "Times New Roman", size = 16)
  )
```

## Save Rarefied Phyloseq
```{r save_rare}
saveRDS(ps_rare, file.path(data_dir, "ps_rare.rds"))
cat("Rarefied phyloseq saved to Data/16S/ps_rare.rds\n")
```

## Session Info
```{r session_info}
sessionInfo()
```

