---
title: "06_DifferentialAbundance"
author: "Chloe J McGovern"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    css: "hpcstyles.css"
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  cache      = TRUE,
  fig.path   = "/home/cjm423/MIX-March2025/Figures/16S/06_DifferentialAbundance/",
  fig.width  = 12,
  fig.height = 8,
  dev        = "png",
  dpi        = 300
)
```

## Overview

Differential abundance analysis of MIX caecal microbiome at genus 
level using DESeq2 (Love et al., 2014, *Genome Biology*). Pairwise 
comparisons between MIX treatment groups are performed alongside a 
likelihood ratio test (LRT) to detect dose-dependent trends across 
MIX1 → MIX2 → MIX3. Raw (non-rarefied) counts are used as DESeq2 
implements its own size factor normalization (McMurdie & Holmes, 2014, 
*PLOS Computational Biology*).

## Libraries
```{r libraries}
library(phyloseq)
library(DESeq2)
library(tidyverse)
library(patchwork)
library(pheatmap)

source("/home/cjm423/MIX-March2025/Code/colors.R")

# MIX group colors for 16S analysis
mix_colors
```

## Load Data
```{r load_data}
project_dir <- "/home/cjm423/MIX-March2025"
data_dir    <- file.path(project_dir, "Data/16S")

# Load clean NON-rarefied phyloseq — DESeq2 uses raw counts
ps_clean <- readRDS(file.path(data_dir, "ps_clean.rds"))

cat("Clean phyloseq (non-rarefied):\n")
print(ps_clean)
```

## Aggregate to Genus Level
```{r genus_aggregate}
# Aggregate ASVs to genus level
# McMurdie & Holmes 2013, PLOS ONE
ps_genus <- tax_glom(ps_clean, taxrank = "Genus", NArm = FALSE)

# Remove unclassified genera
ps_genus <- subset_taxa(ps_genus, !is.na(Genus))

cat("Genera after aggregation:", ntaxa(ps_genus), "\n")
cat("Samples:", nsamples(ps_genus), "\n")

# Set group as factor with explicit ordering for dose-response
sample_data(ps_genus)$group <- factor(
  sample_data(ps_genus)$group,
  levels = c("MIX1", "MIX2", "MIX3")
)
```

## Convert to DESeq2 Object
```{r deseq2_setup}
# Convert phyloseq to DESeq2
dds <- phyloseq_to_deseq2(ps_genus, ~ group)

# Estimate size factors using poscounts method
dds <- estimateSizeFactors(dds, type = "poscounts")

cat("DESeq2 object created\n")
cat("Genera:", nrow(dds), "\n")
cat("Samples:", ncol(dds), "\n")
```

## Pairwise Comparisons — DESeq2 Wald Test
```{r pairwise_deseq2}
# Run DESeq2 with Wald test for pairwise comparisons
dds_wald <- DESeq(dds, test = "Wald", fitType = "parametric")

# Function to extract and format results
get_results <- function(dds, contrast, alpha = 0.05) {
  res <- results(
    dds,
    contrast      = contrast,
    alpha         = alpha,
    pAdjustMethod = "BH"
  )

  res_df <- as.data.frame(res) %>%
    rownames_to_column("taxon") %>%
    left_join(
      as.data.frame(tax_table(ps_genus)) %>%
        rownames_to_column("taxon") %>%
        select(taxon, Genus, Family, Phylum),
      by = "taxon"
    ) %>%
    arrange(padj) %>%
    filter(!is.na(padj))

  return(res_df)
}

# MIX1 vs MIX2
res_1v2 <- get_results(dds_wald, c("group", "MIX2", "MIX1"))
cat("MIX1 vs MIX2 — significant genera (padj < 0.05):",
    sum(res_1v2$padj < 0.05, na.rm = TRUE), "\n")

# MIX1 vs MIX3
res_1v3 <- get_results(dds_wald, c("group", "MIX3", "MIX1"))
cat("MIX1 vs MIX3 — significant genera (padj < 0.05):",
    sum(res_1v3$padj < 0.05, na.rm = TRUE), "\n")

# MIX2 vs MIX3
res_2v3 <- get_results(dds_wald, c("group", "MIX3", "MIX2"))
cat("MIX2 vs MIX3 — significant genera (padj < 0.05):",
    sum(res_2v3$padj < 0.05, na.rm = TRUE), "\n")

# Save all pairwise results
write.csv(res_1v2, file.path(data_dir, "DA_MIX1vsMIX2.csv"),
          row.names = FALSE)
write.csv(res_1v3, file.path(data_dir, "DA_MIX1vsMIX3.csv"),
          row.names = FALSE)
write.csv(res_2v3, file.path(data_dir, "DA_MIX2vsMIX3.csv"),
          row.names = FALSE)
```

No significant difference

## Dose-Response Trend Test — DESeq2 LRT

The likelihood ratio test identifies genera whose abundance changes
systematically across the ordered dose gradient MIX1 → MIX2 → MIX3
(Love et al., 2014, *Genome Biology*).
```{r lrt_dose_response}
# Encode dose as numeric for trend testing
sample_data(ps_genus)$dose_numeric <- as.numeric(
  factor(sample_data(ps_genus)$group,
         levels = c("MIX1", "MIX2", "MIX3"))
)

# Rebuild DESeq2 object with numeric dose
dds_lrt <- phyloseq_to_deseq2(ps_genus, ~ dose_numeric)
dds_lrt <- estimateSizeFactors(dds_lrt, type = "poscounts")

# LRT tests full model (~ dose_numeric) vs reduced (~ 1)
dds_lrt <- DESeq(
  dds_lrt,
  test        = "LRT",
  reduced     = ~ 1,
  fitType     = "parametric"
)

# Extract LRT results
lrt_res <- results(dds_lrt, alpha = 0.05) %>%
  as.data.frame() %>%
  rownames_to_column("taxon") %>%
  left_join(
    as.data.frame(tax_table(ps_genus)) %>%
      rownames_to_column("taxon") %>%
      select(taxon, Genus, Family, Phylum),
    by = "taxon"
  ) %>%
  arrange(padj) %>%
  filter(!is.na(padj))

cat("Genera with significant dose-response trend (padj < 0.05):",
    sum(lrt_res$padj < 0.05, na.rm = TRUE), "\n")

write.csv(lrt_res,
          file.path(data_dir, "DA_dose_response_LRT.csv"),
          row.names = FALSE)

head(lrt_res %>% filter(padj < 0.05), 20)
```

No significant effect dose response.

## Heatmap — Top Genera
```{r heatmap_top_genera, fig.height=10, fig.width=12, dpi=300}
ps_genus_rel <- transform_sample_counts(ps_genus,
                                         function(x) x / sum(x))

top20_genera <- psmelt(ps_genus_rel) %>%
  mutate(Genus = ifelse(is.na(Genus), "Unclassified", Genus)) %>%
  group_by(Genus) %>%
  summarise(mean_abund = mean(Abundance)) %>%
  arrange(desc(mean_abund)) %>%
  slice_head(n = 20) %>%
  pull(Genus)

heatmap_mat <- psmelt(ps_genus_rel) %>%
  mutate(Genus = ifelse(is.na(Genus), "Unclassified", Genus)) %>%
  filter(Genus %in% top20_genera) %>%
  select(Genus, Sample, Abundance) %>%
  pivot_wider(names_from  = Sample,
              values_from = Abundance,
              values_fn   = mean) %>%
  column_to_rownames("Genus") %>%
  as.matrix()

heatmap_mat_log <- log10(heatmap_mat + 1e-6)

# Use full group names to prevent legend truncation
sample_annot <- data.frame(sample_data(ps_genus)) %>%
  select(group) %>%
  mutate(group = recode(group,
    "MIX1" = "MIX1",
    "MIX2" = "MIX2",
    "MIX3" = "MIX3"
  )) %>%
  arrange(group)

heatmap_mat_log <- heatmap_mat_log[, rownames(sample_annot)]

# Match colors to renamed levels
heatmap_colors <- mix_colors
names(heatmap_colors) <- c("MIX1", "MIX2", "MIX3")

pheatmap::pheatmap(
  heatmap_mat_log,
  annotation_col    = sample_annot,
  annotation_colors = list(group = heatmap_colors),
  cluster_rows      = TRUE,
  cluster_cols      = FALSE,
  show_colnames     = TRUE,
  fontsize_row      = 12,
  fontsize_col      = 9,
  color             = colorRampPalette(
                        c("#2166AC", "#F7F7F7", "#B2182B"))(50),
  main              = "Top Genera — Log10 Relative Abundance",
  fontsize          = 10,
  fontfamily        = "Times New Roman",
  annotation_legend = TRUE,
  legend            = TRUE
)
```

## Session Info
```{r session_info}
sessionInfo()
```
