---
title: "07_PhylogeneticTree"
author: "Chloe J McGovern"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    css: "hpcstyles.css"
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  cache      = TRUE,
  fig.path   = "/home/cjm423/MIX-March2025/Figures/16S/07_PhylogeneticTree/",
  fig.width  = 12,
  fig.height = 8,
  dev        = "png",
  dpi        = 300
)
```

## Overview

Construction of a maximum likelihood phylogenetic tree from MIX 
caecal ASV sequences. Full-length 16S rRNA sequences (~1465bp) 
are aligned using DECIPHER (Wright, 2016, *Journal of Statistical 
Software*) and a phylogenetic tree is inferred using phangorn 
(Schliep, 2011, *Bioinformatics*). The tree is integrated into 
the phyloseq object for UniFrac diversity analysis and PICRUSt2 
functional prediction.

## Libraries
```{r libraries}
library(phyloseq)
library(DECIPHER)
library(phangorn)
library(tidyverse)

source("/home/cjm423/MIX-March2025/Code/colors.R")

mix_colors
```

## Load Data
```{r load_data}
project_dir <- "/home/cjm423/MIX-March2025"
data_dir    <- file.path(project_dir, "Data/16S")

# Load phyloseq objects
ps_clean <- readRDS(file.path(data_dir, "ps_clean.rds"))
ps_pre   <- readRDS(file.path(data_dir, "ps_predecontam.rds"))

# Load seqtab to recover sequences
seqtab <- readRDS(file.path(data_dir, "seqtab_nochim.rds"))

cat("Clean phyloseq:\n")
print(ps_clean)

# Build sequence lookup using pre-decontam phyloseq as bridge
seq_lookup <- data.frame(
  asv_label = taxa_names(ps_pre),
  sequence  = colnames(seqtab),
  stringsAsFactors = FALSE
)

# Extract sequences for ps_clean ASVs only
seqs_clean <- seq_lookup %>%
  filter(asv_label %in% taxa_names(ps_clean)) %>%
  arrange(match(asv_label, taxa_names(ps_clean)))

# Build named DNAStringSet
dna_seqs <- DNAStringSet(seqs_clean$sequence)
names(dna_seqs) <- seqs_clean$asv_label

cat("\nASV sequences extracted:", length(dna_seqs), "\n")
cat("NA names:", sum(is.na(names(dna_seqs))), "\n")
cat("Example sequence length:", nchar(seqs_clean$sequence[1]), "bp\n")
```

## Multiple Sequence Alignment

Full-length 16S sequences are aligned using DECIPHER's 
`AlignSeqs()` function which implements a profile-based 
progressive alignment optimized for ribosomal RNA sequences
(Wright, 2016, *Journal of Statistical Software*).

```{r alignment}

# Align sequences using DECIPHER
alignment <- AlignSeqs(
  dna_seqs,
  anchor  = NA,
  verbose = TRUE
)

# Ensure names are preserved after alignment
names(alignment) <- names(dna_seqs)

cat("\nAlignment complete\n")
cat("Alignment length:", width(alignment)[1], "bp\n")
cat("NA names:", sum(is.na(names(alignment))), "\n")

# Save alignment
writeXStringSet(
  alignment,
  file.path(data_dir, "ASV_alignment.fasta")
)
cat("Alignment saved to Data/16S/ASV_alignment.fasta\n")
```

## Convert to phyDat Format
```{r phydat}
phang_align <- phyDat(
  as(alignment, "matrix"),
  type = "DNA"
)

cat("phyDat object created\n")
cat("Sequences:", length(phang_align), "\n")
```

## Build Distance Matrix and Initial Tree
```{r initial_tree}
# Calculate pairwise distances — JC69 for NJ starting tree
dm <- dist.ml(phang_align, model = "JC69")

# Neighbor-joining tree as starting point for ML optimization
nj_tree <- NJ(dm)

cat("Neighbor-joining tree built\n")
cat("Tips:", length(nj_tree$tip.label), "\n")
```

## Maximum Likelihood Tree Optimization
```{r ml_tree}
# GTR+G+I model — standard for 16S rRNA phylogenetics
# GTR accounts for differential substitution rates between bases
# Gamma distribution models rate variation across sites

fit_initial <- pml(nj_tree, data = phang_align)

set.seed(123)
fit_ml <- optim.pml(
  fit_initial,
  model         = "GTR",
  optInv        = TRUE,
  optGamma      = TRUE,
  rearrangement = "stochastic",
  control       = pml.control(trace = 0)
)

cat("\nML tree optimization complete\n")
cat("Log-likelihood:", fit_ml$logLik, "\n")

saveRDS(fit_ml, file.path(data_dir, "ml_tree_fit.rds"))
```

## Root and Visualize Tree
```{r tree_visualization}
# Midpoint rooting — standard when no outgroup is available
ml_tree <- midpoint(fit_ml$tree)

cat("Tree is rooted:", is.rooted(ml_tree), "\n")
cat("Tree tips:", length(ml_tree$tip.label), "\n")

plot(ml_tree,
     type = "fan",
     cex  = 0.5,
     main = "ML Phylogenetic Tree — MIX Caecal ASVs")
```

## Integrate Tree into Phyloseq Objects
```{r add_tree_clean}
# Add tree to clean phyloseq
ps_tree <- merge_phyloseq(ps_clean, phy_tree(ml_tree))

cat("Clean phyloseq with tree:\n")
print(ps_tree)

saveRDS(ps_tree, file.path(data_dir, "ps_tree.rds"))
cat("Clean phyloseq with tree saved to Data/16S/ps_tree.rds\n")
```

```{r add_tree_rare}
# Load rarefied phyloseq and add tree
ps_rare <- readRDS(file.path(data_dir, "ps_rare.rds"))

ps_rare_tree <- merge_phyloseq(ps_rare, phy_tree(ml_tree))

cat("Rarefied phyloseq with tree:\n")
print(ps_rare_tree)

saveRDS(ps_rare_tree, file.path(data_dir, "ps_rare_tree.rds"))
cat("Rarefied phyloseq with tree saved to Data/16S/ps_rare_tree.rds\n")
```

## Export Files for PICRUSt2

PICRUSt2 requires the tree in Newick format and ASV sequences
in FASTA format (Douglas et al., 2020, *Nature Biotechnology*).
```{r export_picrust2}
picrust2_dir <- file.path(project_dir, "Data/16S/picrust2_input")
if (!dir.exists(picrust2_dir)) dir.create(picrust2_dir)

# Write tree in Newick format
write.tree(
  ml_tree,
  file.path(picrust2_dir, "ASV_tree.nwk")
)

# Write ASV sequences
writeXStringSet(
  dna_seqs,
  file.path(picrust2_dir, "ASV_seqs.fasta")
)

# Write ASV count table (samples x ASVs) for PICRUSt2
asv_table <- as.data.frame(t(otu_table(ps_clean)))
write.table(
  asv_table,
  file.path(picrust2_dir, "ASV_table.tsv"),
  sep       = "\t",
  quote     = FALSE,
  col.names = NA
)

cat("PICRUSt2 input files saved to Data/16S/picrust2_input/\n")
cat("Files:\n")
cat("  ASV_tree.nwk\n")
cat("  ASV_seqs.fasta\n")
cat("  ASV_table.tsv\n")
```

## Session Info
```{r session_info}
sessionInfo()
```


