---
title: "08_UnifracDiversity"
author: "Chloe J McGovern"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    css: "hpcstyles.css"
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  cache      = TRUE,
  fig.path   = "/home/cjm423/MIX-March2025/Figures/16S/08_UniFrac/",
  fig.width  = 12,
  fig.height = 8,
  dev        = "png",
  dpi        = 300
)
```

# Overview

Phylogeny-aware beta diversity analysis using UniFrac distance metrics
(Lozupone & Knight, 2005, *Applied and Environmental Microbiology*).
Unweighted UniFrac captures presence/absence of lineages while weighted
UniFrac incorporates relative abundance. Both metrics are compared to
Bray-Curtis dissimilarity from `05_Phyloseq.Rmd`. PERMANOVA tests
whether treatment group significantly explains community variation
(Anderson, 2001, *Australian Ecology*).

## Libraries and Set Up Environment
```{r libraries}
library(phyloseq)
library(vegan)
library(tidyverse)
library(patchwork)

source("/home/cjm423/MIX-March2025/Code/colors.R")

mix_colors
```

## Load Data
```{r load_data}
project_dir <- "/home/cjm423/MIX-March2025"
data_dir    <- file.path(project_dir, "Data/16S")

# Load rarefied phyloseq with tree from 07_PhylogeneticTree.Rmd
ps_rare_tree <- readRDS(file.path(data_dir, "ps_rare_tree.rds"))

cat("Rarefied phyloseq with tree:\n")
print(ps_rare_tree)

cat("\nTree present:", !is.null(phy_tree(ps_rare_tree)), "\n")
cat("Tree tips:", length(phy_tree(ps_rare_tree)$tip.label), "\n")

# Extract metadata
meta_df <- data.frame(sample_data(ps_rare_tree)) %>%
  mutate(group = factor(group, levels = c("MIX1", "MIX2", "MIX3")))
```

## Calculate UniFrac Distances
```{r unifrac_distances}
# Unweighted UniFrac — presence/absence with phylogeny
# Lozupone & Knight 2005, Applied and Environmental Microbiology
cat("Calculating unweighted UniFrac...\n")
uw_unifrac <- phyloseq::distance(ps_rare_tree, method = "unifrac",
                                  weighted = FALSE)

# Weighted UniFrac — abundance-weighted with phylogeny
cat("Calculating weighted UniFrac...\n")
w_unifrac <- phyloseq::distance(ps_rare_tree, method = "unifrac",
                                 weighted = TRUE)

# Bray-Curtis for comparison
cat("Calculating Bray-Curtis...\n")
bray_dist <- phyloseq::distance(ps_rare_tree, method = "bray")

cat("\nAll distance matrices calculated\n")
cat("Samples:", nsamples(ps_rare_tree), "\n")
```

## NMDS Ordination — Unweighted UniFrac
```{r nmds_unwunifrac}
set.seed(123)
nmds_uw <- ordinate(ps_rare_tree, method = "NMDS",
                     distance = uw_unifrac)

cat("Unweighted UniFrac NMDS stress:", nmds_uw$stress, "\n")

p_uw <- plot_ordination(ps_rare_tree, nmds_uw, color = "group") +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = mix_colors) +
  stat_ellipse(aes(group = group), level = 0.95) +
  labs(
    title = paste0("Unweighted UniFrac NMDS (stress = ",
                   round(nmds_uw$stress, 3), ")"),
    color = "Treatment Group"
  ) +
  theme_bw() +
  theme(
    legend.position = "right",
    text = element_text(family = "Times New Roman", size = 16)
  )

print(p_uw)
```

## NMDS Ordination — Weighted UniFrac
```{r nmds_wunifrac}
set.seed(123)
nmds_w <- ordinate(ps_rare_tree, method = "NMDS",
                    distance = w_unifrac)

cat("Weighted UniFrac NMDS stress:", nmds_w$stress, "\n")

p_w <- plot_ordination(ps_rare_tree, nmds_w, color = "group") +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = mix_colors) +
  stat_ellipse(aes(group = group), level = 0.95) +
  labs(
    title = paste0("Weighted UniFrac NMDS (stress = ",
                   round(nmds_w$stress, 3), ")"),
    color = "Treatment Group"
  ) +
  theme_bw() +
  theme(
    legend.position = "right",
    text = element_text(family = "Times New Roman", size = 16)
  )

print(p_w)
```

## PCoA Ordination — All Three Metrics
```{r pcoa_comparison, fig.height=12}
pcoa_uw <- ordinate(ps_rare_tree, method = "PCoA",
                     distance = uw_unifrac)
pcoa_w  <- ordinate(ps_rare_tree, method = "PCoA",
                     distance = w_unifrac)
pcoa_bc <- ordinate(ps_rare_tree, method = "PCoA",
                     distance = bray_dist)

get_var <- function(pcoa) {
  eig <- pcoa$values$Eigenvalues
  round(eig / sum(eig) * 100, 1)
}

var_uw <- get_var(pcoa_uw)
var_w  <- get_var(pcoa_w)
var_bc <- get_var(pcoa_bc)

p_pcoa_uw <- plot_ordination(ps_rare_tree, pcoa_uw,
                              color = "group") +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = mix_colors) +
  stat_ellipse(aes(group = group), level = 0.95) +
  labs(
    title = "Unweighted UniFrac PCoA",
    x     = paste0("PC1 (", var_uw[1], "%)"),
    y     = paste0("PC2 (", var_uw[2], "%)"),
    color = "Treatment Group"
  ) +
  theme_bw() +
  theme(text = element_text(family = "Times New Roman", size = 14))

p_pcoa_w <- plot_ordination(ps_rare_tree, pcoa_w,
                             color = "group") +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = mix_colors) +
  stat_ellipse(aes(group = group), level = 0.95) +
  labs(
    title = "Weighted UniFrac PCoA",
    x     = paste0("PC1 (", var_w[1], "%)"),
    y     = paste0("PC2 (", var_w[2], "%)"),
    color = "Treatment Group"
  ) +
  theme_bw() +
  theme(text = element_text(family = "Times New Roman", size = 14))

p_pcoa_bc <- plot_ordination(ps_rare_tree, pcoa_bc,
                              color = "group") +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = mix_colors) +
  stat_ellipse(aes(group = group), level = 0.95) +
  labs(
    title = "Bray-Curtis PCoA",
    x     = paste0("PC1 (", var_bc[1], "%)"),
    y     = paste0("PC2 (", var_bc[2], "%)"),
    color = "Treatment Group"
  ) +
  theme_bw() +
  theme(text = element_text(family = "Times New Roman", size = 14))

p_pcoa_uw / p_pcoa_w / p_pcoa_bc
```

## PERMANOVA — All Three Metrics
```{r permanova_all}
set.seed(123)

permanova_uw <- adonis2(
  uw_unifrac ~ group,
  data         = meta_df,
  permutations = 999
)

permanova_w <- adonis2(
  w_unifrac ~ group,
  data         = meta_df,
  permutations = 999
)

permanova_bc <- adonis2(
  bray_dist ~ group,
  data         = meta_df,
  permutations = 999
)

cat("=== Unweighted UniFrac PERMANOVA ===\n")
print(permanova_uw)

cat("\n=== Weighted UniFrac PERMANOVA ===\n")
print(permanova_w)

cat("\n=== Bray-Curtis PERMANOVA ===\n")
print(permanova_bc)

# Summary table
permanova_summary <- data.frame(
  metric  = c("Unweighted UniFrac", "Weighted UniFrac",
               "Bray-Curtis"),
  R2      = c(permanova_uw$R2[1], permanova_w$R2[1],
               permanova_bc$R2[1]),
  F_stat  = c(permanova_uw$F[1], permanova_w$F[1],
               permanova_bc$F[1]),
  p_value = c(permanova_uw$`Pr(>F)`[1], permanova_w$`Pr(>F)`[1],
               permanova_bc$`Pr(>F)`[1])
) %>%
  mutate(
    R2      = round(R2, 3),
    F_stat  = round(F_stat, 3),
    p_value = round(p_value, 3),
    sig     = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      p_value < 0.1   ~ ".",
      TRUE            ~ "NS"
    )
  )

cat("\n=== PERMANOVA Summary ===\n")
print(permanova_summary)

write.csv(permanova_summary,
          file.path(data_dir, "permanova_summary.csv"),
          row.names = FALSE)
```

## Betadispersion — All Three Metrics
```{r betadisper_all}
set.seed(123)

betadisp_uw <- betadisper(uw_unifrac, meta_df$group)
betadisp_w  <- betadisper(w_unifrac, meta_df$group)
betadisp_bc <- betadisper(bray_dist, meta_df$group)

cat("=== Unweighted UniFrac Betadispersion ===\n")
print(permutest(betadisp_uw, permutations = 999))

cat("\n=== Weighted UniFrac Betadispersion ===\n")
print(permutest(betadisp_w, permutations = 999))

cat("\n=== Bray-Curtis Betadispersion ===\n")
print(permutest(betadisp_bc, permutations = 999))
```

## Distance Metric Correlation

How correlated are the three distance matrices? High correlation
confirms results are robust across metrics (Legendre & Legendre,
2012, *Numerical Ecology*).
```{r distance_correlation}
mantel_uw_w  <- vegan::mantel(uw_unifrac, w_unifrac,
                               method = "pearson", permutations = 999)
mantel_uw_bc <- vegan::mantel(uw_unifrac, bray_dist,
                               method = "pearson", permutations = 999)
mantel_w_bc  <- vegan::mantel(w_unifrac, bray_dist,
                               method = "pearson", permutations = 999)

cat("Mantel test correlations between distance matrices:\n")
cat("Unweighted UniFrac vs Weighted UniFrac: r =",
    round(mantel_uw_w$statistic, 3),
    "p =", mantel_uw_w$signif, "\n")
cat("Unweighted UniFrac vs Bray-Curtis: r =",
    round(mantel_uw_bc$statistic, 3),
    "p =", mantel_uw_bc$signif, "\n")
cat("Weighted UniFrac vs Bray-Curtis: r =",
    round(mantel_w_bc$statistic, 3),
    "p =", mantel_w_bc$signif, "\n")
```

## Session Info
```{r session_info}
sessionInfo()
```
