---
title: "09_Picrust2_KEGG"
author: "Chloe J McGovern"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    css: "hpcstyles.css"
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  cache      = TRUE,
  fig.path   = "/home/cjm423/MIX-March2025/Figures/16S/09_PICRUSt2/",
  fig.width  = 12,
  fig.height = 8,
  dev        = "png",
  dpi        = 300
)
```

# Overview

Functional prediction analysis using PICRUSt2 v2.6.2 (Douglas et al., 
2020, *Nature Biotechnology*). KEGG Ortholog (KO) and MetaCyc pathway 
abundances were predicted from 16S rRNA ASV sequences using phylogenetic 
placement onto reference trees via EPA-NG. Differential functional 
abundance between MIX treatment groups is tested using DESeq2 (Love et 
al., 2014, *Genome Biology*).

## Libraries and Set Up
```{r libraries}
library(ggpicrust2)
library(tidyverse)
library(DESeq2)
library(pheatmap)
library(patchwork)
library(readr)

source("/home/cjm423/MIX-March2025/Code/colors.R")

mix_colors

# Load pathway descriptions from PICRUSt2 mapfiles
path_desc <- read_tsv(
  "/home/cjm423/workdir/picrust2-2.6.2/picrust2/default_files/pathway_mapfiles/metacyc_pathways_structured_filtered_v24.txt",
  col_names = FALSE
)

head(path_desc)
```

## File Paths and Data
```{r paths}
project_dir  <- "/home/cjm423/MIX-March2025"
data_dir     <- file.path(project_dir, "Data/16S")
picrust_dir  <- file.path(data_dir, "picrust2_output")
fig_dir      <- file.path(project_dir, "Figures/16S/09_PICRUSt2")

if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)

ko_file      <- file.path(picrust_dir, "ko.txt_metagenome_out/pred_metagenome_unstrat.tsv.gz")
ec_file      <- file.path(picrust_dir, "ec.txt_metagenome_out/pred_metagenome_unstrat.tsv.gz")
pathway_file <- file.path(picrust_dir, "pathways_out/path_abun_unstrat.tsv.gz")
nsti_file    <- file.path(picrust_dir, "combined_marker_predicted_and_nsti.tsv.gz")

stopifnot(file.exists(ko_file))
stopifnot(file.exists(ec_file))
stopifnot(file.exists(pathway_file))
stopifnot(file.exists(nsti_file))

cat("All PICRUSt2 output files found\n")
```

## Load Metadata
```{r metadata}
metadata <- read.csv(file.path(data_dir, "MIX_metadata.csv")) %>%
  filter(!is_control) %>%  # exclude H2O controls
  mutate(group = factor(group, levels = c("MIX1", "MIX2", "MIX3")))

cat("Samples:", nrow(metadata), "\n")
print(metadata$sample_id)
```

## NSTI Quality Check

NSTI values indicate how well each ASV is represented in the PICRUSt2
reference database. Lower values indicate higher confidence predictions
(Douglas et al., 2020, *Nature Biotechnology*).
```{r nsti_check}
nsti_raw <- read_tsv(nsti_file)
nsti_df  <- nsti_raw %>%
  select(asv_id = 1, nsti = 3)

cat("NSTI summary:\n")
print(summary(nsti_df$nsti))
cat("\nASVs above 0.05 threshold:", sum(nsti_df$nsti > 0.05), "\n")
cat("ASVs above 0.1 threshold:", sum(nsti_df$nsti > 0.1), "\n")

ggplot(nsti_df, aes(x = reorder(asv_id, nsti), y = nsti)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  geom_hline(yintercept = 0.1,  linetype = "dashed", color = "orange") +
  labs(
    title    = "NSTI per ASV — MIX-March2025",
    subtitle = "Red = 0.05 threshold, Orange = 0.1 threshold",
    x        = "ASV",
    y        = "NSTI"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 5),
    text        = element_text(family = "Times New Roman", size = 16)
  )
```

## Load KO Abundances
```{r load_ko}
ko_raw <- read_tsv(ko_file)
colnames(ko_raw)[1] <- "ko_id"

ko_abun <- ko_raw %>%
  column_to_rownames("ko_id")

cat("KO features:", nrow(ko_abun), "\n")
cat("Samples:", ncol(ko_abun), "\n")

ko_abun <- ko_abun[, metadata$sample_id]
cat("After filtering to biological samples:", ncol(ko_abun), "\n")
```

## Top KO Abundances
```{r top_ko, fig.height=10}
ko_means <- rowMeans(ko_abun) %>%
  sort(decreasing = TRUE) %>%
  head(30)

ko_top_df <- data.frame(
  ko_id      = names(ko_means),
  mean_abund = ko_means
)

ggplot(ko_top_df, aes(x = reorder(ko_id, mean_abund),
                       y = mean_abund)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 30 KEGG Orthologs by Mean Predicted Abundance",
    x     = "KO ID",
    y     = "Mean Predicted Abundance"
  ) +
  theme_bw() +
  theme(
    text = element_text(family = "Times New Roman", size = 16)
  )
```

## KO Heatmap — Top 50
```{r ko_heatmap, fig.height=14}
top50_ko <- names(sort(rowMeans(ko_abun), decreasing = TRUE))[1:50]
ko_mat   <- log10(ko_abun[top50_ko, ] + 1)

sample_annot <- data.frame(
  group     = metadata$group,
  row.names = metadata$sample_id
)

pheatmap(
  as.matrix(ko_mat),
  annotation_col    = sample_annot,
  annotation_colors = list(group = mix_colors),
  cluster_rows      = TRUE,
  cluster_cols      = FALSE,
  show_rownames     = TRUE,
  show_colnames     = TRUE,
  fontsize_row      = 10,
  fontsize_col      = 10,
  fontfamily        = "serif",
  color             = colorRampPalette(
                        c("white", "steelblue", "darkblue"))(50),
  main              = "Top 50 KOs — log10 Predicted Abundance"
)
```

## Load MetaCyc Pathway Abundances
```{r load_pathways}
path_raw <- read_tsv(pathway_file)
colnames(path_raw)[1] <- "pathway"

path_abun <- path_raw %>%
  column_to_rownames("pathway")

cat("Pathways:", nrow(path_abun), "\n")
cat("Samples:", ncol(path_abun), "\n")

path_abun <- path_abun[, metadata$sample_id]
cat("After filtering to biological samples:", ncol(path_abun), "\n")
```

## Top MetaCyc Pathways
```{r top_pathways, fig.height=10}
path_means <- rowMeans(path_abun) %>%
  sort(decreasing = TRUE) %>%
  head(30)

path_top_df <- data.frame(
  pathway    = names(path_means),
  mean_abund = path_means
)

ggplot(path_top_df, aes(x = reorder(pathway, mean_abund),
                         y = mean_abund)) +
  geom_col(fill = "cadetblue3") +
  coord_flip() +
  labs(
    title = "Top 30 MetaCyc Pathways by Mean Predicted Abundance — MIX-March2025",
    x     = "Pathway",
    y     = "Mean Predicted Abundance"
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_text(size = 7),
    text        = element_text(family = "Times New Roman", size = 16)
  )
```

## Differential Pathway Abundance — DESeq2
```{r deseq2_pathways}
path_counts <- round(path_abun)
path_counts <- path_counts[rowSums(path_counts) > 0, ]

meta_deseq <- metadata %>%
  column_to_rownames("sample_id")

dds_path <- DESeqDataSetFromMatrix(
  countData = path_counts,
  colData   = meta_deseq,
  design    = ~ group
)

dds_path <- estimateSizeFactors(dds_path, type = "poscounts")
dds_path <- DESeq(dds_path, test = "Wald", fitType = "parametric")

cat("DESeq2 pathway analysis complete\n")

get_path_results <- function(dds, contrast, alpha = 0.05) {
  res <- results(dds, contrast = contrast,
                 alpha = alpha, pAdjustMethod = "BH")
  as.data.frame(res) %>%
    rownames_to_column("pathway") %>%
    arrange(padj) %>%
    filter(!is.na(padj))
}

path_1v2 <- get_path_results(dds_path, c("group", "MIX2", "MIX1"))
path_1v3 <- get_path_results(dds_path, c("group", "MIX3", "MIX1"))
path_2v3 <- get_path_results(dds_path, c("group", "MIX3", "MIX2"))

cat("MIX1 vs MIX2 significant pathways:",
    sum(path_1v2$padj < 0.05, na.rm = TRUE), "\n")
cat("MIX1 vs MIX3 significant pathways:",
    sum(path_1v3$padj < 0.05, na.rm = TRUE), "\n")
cat("MIX2 vs MIX3 significant pathways:",
    sum(path_2v3$padj < 0.05, na.rm = TRUE), "\n")

write.csv(path_1v2, file.path(data_dir, "pathway_DA_MIX1vsMIX2.csv"),
          row.names = FALSE)
write.csv(path_1v3, file.path(data_dir, "pathway_DA_MIX1vsMIX3.csv"),
          row.names = FALSE)
write.csv(path_2v3, file.path(data_dir, "pathway_DA_MIX2vsMIX3.csv"),
          row.names = FALSE)
```

## Pathway Dose-Response LRT
```{r lrt_pathways}
meta_deseq$dose_numeric <- as.numeric(
  factor(meta_deseq$group, levels = c("MIX1", "MIX2", "MIX3"))
)

dds_path_lrt <- DESeqDataSetFromMatrix(
  countData = path_counts,
  colData   = meta_deseq,
  design    = ~ dose_numeric
)

dds_path_lrt <- estimateSizeFactors(dds_path_lrt, type = "poscounts")
dds_path_lrt <- DESeq(dds_path_lrt, test = "LRT",
                       reduced = ~ 1, fitType = "parametric")

lrt_path <- results(dds_path_lrt, alpha = 0.05) %>%
  as.data.frame() %>%
  rownames_to_column("pathway") %>%
  arrange(padj) %>%
  filter(!is.na(padj))

cat("Pathways with significant dose-response trend (padj < 0.05):",
    sum(lrt_path$padj < 0.05, na.rm = TRUE), "\n")

write.csv(lrt_path,
          file.path(data_dir, "pathway_DA_dose_response_LRT.csv"),
          row.names = FALSE)

head(lrt_path, 20)

print(lrt_path)
```

## Pathway Names

Pathway names added manually from MetaCyc PWY identifiers for
figure interpretability. Update this lookup table as significant
pathways are identified in MIX results.

```{r pathway_names}
# NOTE: This lookup table will need to be updated based on
# which pathways are significant in MIX results.
# Populated from LPHY as a starting reference — add/remove as needed.
pathway_names  <- data.frame(
  pathway = c(
    "PWY0-1586", "PWY-8178", "PWY-5973", "PWY-6317", "PENTOSE-P-PWY",
    "COMPLETE-ARO-PWY", "PWY-7220", "PWY-7222", "PWY-6163", "RIBOSYN2-PWY",
    "PWY-6168", "DAPLYSINESYN-PWY", "THRESYN-PWY", "PWY-7208", "PWY-7228",
    "PWY-6122", "PWY-6277", "ARO-PWY", "PWY-2942", "P461-PWY",
    "PWY-6385", "PWY-7953", "PWY-6125", "PWY-5667", "PWY0-1319",
    "PWY-7790", "PWY-7791", "PWY-5686", "PWY-6387", "POLYISOPRENSYN-PWY",
    "THISYNARA-PWY", "PWY-6121", "PWY-2941", "PWY-7221", "PHOSLIPSYN-PWY",
    "GLYCOGENSYNTH-PWY", "PWY-6151", "PWY-6147", "PWY-7663", "NONOXIPENT-PWY",
    # --- new: dose-responsive pathway heatmap ---
    "PWY-1042", "PWY-6897", "PWY-7357", "ARGSYN-PWY", "SER-GLYSYN-PWY",
    "GLUTORN-PWY", "PWY-3841", "PWY-6859", "UDPNAGSYN-PWY", "COA-PWY-1",
    "PWY-6124", "COA-PWY", "PWY-7851", "GLYCOLYSIS", "PWY-6126",
    "PWY4FS-7", "PWY4FS-8", "PWY-7229", "PWY-6123", "PWY-7356",
    "ARGSYNBSUB-PWY", "PWY-6386", "PEPTIDOGLYCANSYN-PWY", "PWY-841",
    "SALVADEHYPOX-PWY", "PWY-6922",
    "PWY-5100", "OANTIGEN-PWY", "PWY-5659", "PWY-6293", "PWY4LZ-257",
    "CATECHOL-ORTHO-CLEAVAGE-PWY", "PWY-5910", "PWY-561", "PWY-7376",
    "PWY-3661", "PWY-5130", "PWY-8011", "DENITRIFICATION-PWY", "PWY-7900",
    "PWY-1622", "PWY-6313", "PWY-7431", "PWY-6383", "P105-PWY",
    "REDCITCYC", "ANAEROFRUCAT-PWY", "PWY-5861", "FERMENTATION-PWY",
    "PWY-5838", "PWY-5345", "CITRULBIO-PWY", "PWY-821", "HEXITOLDEGSUPER-PWY",
    "PWY-622", "SULFATE-CYS-PWY", "PWY-5850", "SO4ASSIM-PWY", "PWY-5840",
    "GLUCOSE1PMETAB-PWY", "PWY-7237", "PWY-5918"
  ),
  name = c(
    "Peptidoglycan maturation (meso-diaminopimelate containing)",
    "Pentose phosphate pathway (non-oxidative branch) II",
    "cis-vaccenate biosynthesis",
    "Galactose degradation I (Leloir pathway)",
    "Pentose phosphate pathway",
    "Superpathway of aromatic amino acid biosynthesis",
    "Adenosine deoxyribonucleotides de novo biosynthesis II",
    "Guanosine deoxyribonucleotides de novo biosynthesis II",
    "Chorismate biosynthesis from 3-dehydroquinate",
    "Flavin biosynthesis I (bacteria and plants)",
    "Flavin biosynthesis III (fungi)",
    "L-lysine biosynthesis I",
    "Superpathway of L-threonine biosynthesis",
    "Superpathway of pyrimidine nucleobases salvage",
    "Superpathway of guanosine nucleotides de novo biosynthesis I",
    "5-aminoimidazole ribonucleotide biosynthesis II",
    "Superpathway of 5-aminoimidazole ribonucleotide biosynthesis",
    "Chorismate biosynthesis I",
    "L-lysine biosynthesis III",
    "Hexitol fermentation to lactate, formate, ethanol and acetate",
    "Peptidoglycan biosynthesis III (mycobacteria)",
    "UDP-N-acetylmuramoyl-pentapeptide biosynthesis III", 
    "Superpathway of guanosine nucleotides de novo biosynthesis II",
    "CDP-diacylglycerol biosynthesis I",
    "CDP-diacylglycerol biosynthesis II",
    "UMP biosynthesis II",  
    "UMP biosynthesis III",
    "UMP biosynthesis",
    "UDP-N-acetylmuramoyl-pentapeptide biosynthesis I (meso-DAP)",
    "Polyisoprenoid biosynthesis (E. coli)",
    "Superpathway of thiamin diphosphate biosynthesis III (eukaryotes)",
    "5-aminoimidazole ribonucleotide biosynthesis I",
    "L-lysine biosynthesis II",
    "Guanosine ribonucleotides de novo biosynthesis",
    "Superpathway of phospholipid biosynthesis I (bacteria)",
    "Glycogen biosynthesis I (from ADP-D-glucose)",
    "S-adenosyl-L-methionine cycle I",
    "6-hydroxymethyl-dihydropterin diphosphate biosynthesis I",
    "Gondoate biosynthesis (anaerobic)",
    "Pentose phosphate pathway (non-oxidative branch)",
    "Glycolysis IV", 
    "Thiamin salvage II",
    "Thiamine diphosphate formation from pyrithiamine and oxythiamine", 
    "L-arginine biosynthesis I (via L-ornithine)",
    "Superpathway of L-serine and glycine biosynthesis I",
    "L-ornithine biosynthesis",
    "Folate transformations II",
    "All-trans-farnesol biosynthesis",  
    "UDP-N-acetyl-D-glucosamine biosynthesis I",
    "Superpathway of coenzyme A biosynthesis III",  
    "Inosine-5'-phosphate biosynthesis II",  
    "Coenzyme A biosynthesis I",
    "Coenzyme A biosynthesis II", 
    "Glycolysis I (from glucose 6-phosphate)",
    "Superpathway of adenosine nucleotides de novo biosynthesis II",
    "Phosphatidylglycerol biosynthesis I (plastidic)",
    "Phosphatidylglycerol biosynthesis II (non-plastidic)",
    "Superpathway of adenosine nucleotides de novo biosynthesis I",
    "Inosine-5'-phosphate biosynthesis I",
    "Thiamine diphosphate salvage IV",  
    "L-arginine biosynthesis II (acetyl cycle)",
    "UDP-N-acetylmuramoyl-pentapeptide biosynthesis II (lysine-containing)",
    "Peptidoglycan biosynthesis I (meso-diaminopimelate containing)",
    "Superpathway of purine nucleotides de novo biosynthesis I",
    "Adenosine nucleotides degradation II",
    "L-ornithine degradation I (L-proline biosynthesis)",
    "Pyruvate fermentation to acetate and lactate II",
    "O-antigen building blocks biosynthesis (E. coli)",
    "GDP-mannose biosynthesis",
    "Superpathway of L-cysteine biosynthesis",
    "Superpathway of fermentation (Chlamydomonas reinhardtii)",
    "Catechol degradation to beta-ketoadipate",
    "Superpathway of geranylgeranyldiphosphate biosynthesis I (via mevalonate)",
    "Superpathway of glyoxylate cycle and fatty acid degradation",
    "cob(II)yrinate a,c-diamide biosynthesis II (late cobalt incorporation)",
    "Glycine betaine degradation I",
    "2-oxobutanoate degradation I", 
    "L-serine biosynthesis II",  
    "Nitrate reduction I (denitrification)",
    "Glycogen biosynthesis III (from α-maltose 1-phosphate)", 
    "Formaldehyde assimilation I (serine pathway)",
    "Serotonin degradation",
    "Aromatic biogenic amine degradation (bacteria)",
    "Mono-trans,poly-cis decaprenyl phosphate biosynthesis",
    "TCA cycle IV (2-oxoglutarate decarboxylase)",
    "TCA cycle VIII (helicobacter)",
    "Homolactic fermentation",
    "Superpathway of demethylmenaquinol-8 biosynthesis",
    "Mixed acid fermentation",
    "Superpathway of menaquinol-8 biosynthesis I",
    "Superpathway of L-methionine biosynthesis (by sulfhydrylation)",
    "Citrulline biosynthesis",
    "Superpathway of sulfur amino acid biosynthesis (S. cerevisiae)",
    "Superpathway of hexitol degradation (bacteria)",
    "Starch biosynthesis",
    "Superpathway of sulfate assimilation and cysteine biosynthesis",
    "Superpathway of menaquinol-6 biosynthesis I",
    "Sulfate reduction I (assimilatory)",
    "Superpathway of menaquinol-7 biosynthesis",
    "Glucose and glucose-1-phosphate degradation",
    "Myo-, chiro- and scillo-inositol degradation",
    "Superpathway of heme b biosynthesis from glutamate"
  ),
  stringsAsFactors = FALSE
)

lrt_path_named <- lrt_path %>%
  left_join(pathway_names, by = "pathway") %>%
  mutate(label = ifelse(!is.na(name), name, pathway))

write.csv(lrt_path_named,
          file.path(data_dir, "pathway_DA_dose_response_LRT_named.csv"),
          row.names = FALSE)

cat("Total pathways in lookup table:", nrow(pathway_names), "\n")
print(pathway_names)
```

## Pathway Heatmap — Top 40
```{r pathway_heatmap, fig.height=12}
# Normalize to relative abundance (or CPM) per sample BEFORE ranking/log-transforming
path_relabund <- sweep(path_abun, 2, colSums(path_abun), FUN = "/") * 1e6  # CPM

# Rank top 40 on normalized values, not raw
top40_path <- names(sort(rowMeans(path_relabund),
                          decreasing = TRUE))[1:40]

path_mat <- log10(path_relabund[top40_path, ] + 1)

rownames(path_mat) <- ifelse(
  rownames(path_mat) %in% pathway_names$pathway,
  pathway_names$name[match(rownames(path_mat), pathway_names$pathway)],
  rownames(path_mat)
)

pheatmap(
  as.matrix(path_mat),
  annotation_col    = sample_annot,
  annotation_colors = list(group = mix_colors),
  cluster_rows      = TRUE,
  cluster_cols      = FALSE,
  show_rownames     = TRUE,
  show_colnames     = TRUE,
  fontsize_row      = 12,
  fontsize_col      = 12,
  fontfamily        = "serif",
  color             = colorRampPalette(
                        c("white", "cadetblue3", "darkblue"))(50),
  main              = "Top 40 MetaCyc Pathways — log10 CPM"
)
```

## Pairwise Heatmap — Significant Pathways
```{r pairwise_pathway_heatmap, fig.height=12}
sig_1v2 <- path_1v2 %>% filter(padj < 0.05) %>% pull(pathway)
sig_1v3 <- path_1v3 %>% filter(padj < 0.05) %>% pull(pathway)
sig_all <- union(sig_1v2, sig_1v3)

cat("Significant pathways MIX1 vs MIX2:", length(sig_1v2), "\n")
cat("Significant pathways MIX1 vs MIX3:", length(sig_1v3), "\n")
cat("Union of significant pathways:", length(sig_all), "\n")

if (length(sig_all) > 0) {
  sig_all_mat <- log10(path_abun[sig_all, ] + 1)
  rownames(sig_all_mat) <- ifelse(
    rownames(sig_all_mat) %in% pathway_names$pathway,
    pathway_names$name[match(rownames(sig_all_mat),
                              pathway_names$pathway)],
    rownames(sig_all_mat)
  )

  pheatmap(
    as.matrix(sig_all_mat),
    annotation_col    = sample_annot,
    annotation_colors = list(group = mix_colors),
    cluster_rows      = TRUE,
    cluster_cols      = FALSE,
    show_rownames     = TRUE,
    show_colnames     = TRUE,
    fontsize_row      = 7,
    fontsize_col      = 8,
    fontfamily        = "serif",
    color             = colorRampPalette(
                          c("white", "cadetblue3", "darkblue"))(50),
    main = "Significant Pathways — MIX1 vs MIX2 and MIX1 vs MIX3 (padj < 0.05)"
  )
} else {
  cat("No significant pathways to plot\n")
}
```

## Volcano Plots — Pairwise Comparisons
```{r volcano_pathways_1v2, fig.height=8}
path_1v2 %>%
  mutate(
    significance = case_when(
      padj < 0.05 & log2FoldChange > 1  ~ "Up in MIX2",
      padj < 0.05 & log2FoldChange < -1 ~ "Down in MIX2",
      TRUE                               ~ "NS"
    ),
    label = ifelse(padj < 0.05,
                   ifelse(pathway %in% pathway_names$pathway,
                          pathway_names$name[match(pathway,
                                                    pathway_names$pathway)],
                          pathway),
                   NA)
  ) %>%
  ggplot(aes(x = log2FoldChange, y = -log10(padj),
             color = significance, label = label)) +
  geom_point(alpha = 0.7, size = 3) +
  ggrepel::geom_text_repel(size = 2.5, max.overlaps = 20,
                            family = "serif") +
  scale_color_manual(values = c(
    "Up in MIX2"   = "cyan3",
    "Down in MIX2" = "coral2",
    "NS"           = "grey70"
  )) +
  geom_vline(xintercept = c(-1, 1),
             linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed", color = "grey40") +
  labs(
    title = "Differential Pathway Abundance — MIX2 vs MIX1",
    x     = "Log2 Fold Change",
    y     = "-log10(adjusted p-value)",
    color = "Direction"
  ) +
  theme_bw() +
  theme(text = element_text(family = "Times New Roman", size = 14))
```

```{r volcano_pathways_1v3, fig.height=8}
path_1v3 %>%
  mutate(
    significance = case_when(
      padj < 0.05 & log2FoldChange > 1  ~ "Up in MIX3",
      padj < 0.05 & log2FoldChange < -1 ~ "Down in MIX3",
      TRUE                               ~ "NS"
    ),
    label = ifelse(padj < 0.05,
                   ifelse(pathway %in% pathway_names$pathway,
                          pathway_names$name[match(pathway,
                                                    pathway_names$pathway)],
                          pathway),
                   NA)
  ) %>%
  ggplot(aes(x = log2FoldChange, y = -log10(padj),
             color = significance, label = label)) +
  geom_point(alpha = 0.7, size = 3) +
  ggrepel::geom_text_repel(size = 2.5, max.overlaps = 20,
                            family = "serif") +
  scale_color_manual(values = c(
    "Up in MIX3"   = "mediumseagreen",
    "Down in MIX3" = "coral2",
    "NS"           = "grey70"
  )) +
  geom_vline(xintercept = c(-1, 1),
             linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed", color = "grey40") +
  labs(
    title = "Differential Pathway Abundance — MIX3 vs MIX1",
    x     = "Log2 Fold Change",
    y     = "-log10(adjusted p-value)",
    color = "Direction"
  ) +
  theme_bw() +
  theme(text = element_text(family = "Times New Roman", size = 14))
```

## Effect Size Plots — Pairwise
```{r pairwise_foldchange_1v2, fig.height=8}
path_1v2 %>%
  filter(!is.na(log2FoldChange)) %>%
  arrange(desc(abs(log2FoldChange))) %>%
  slice_head(n = 20) %>%
  left_join(pathway_names, by = "pathway") %>%
  mutate(
    label       = ifelse(!is.na(name), name, pathway),
    label       = reorder(label, log2FoldChange),
    direction   = ifelse(log2FoldChange > 0, "Higher in MIX2",
                                             "Higher in MIX1"),
    significant = padj < 0.05
  ) %>%
  ggplot(aes(x = log2FoldChange, y = label, fill = direction,
              alpha = significant)) +
  geom_col() +
  geom_errorbarh(
    aes(xmin = log2FoldChange - lfcSE,
        xmax = log2FoldChange + lfcSE),
    height = 0.3
  ) +
  scale_fill_manual(values = c(
    "Higher in MIX2" = "cyan3",
    "Higher in MIX1" = "coral2"
  )) +
  scale_alpha_manual(values = c("TRUE" = 1.0, "FALSE" = 0.4),
                     guide = "none") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title    = "Top 20 Pathways by Effect Size — MIX2 vs MIX1",
    subtitle = "Solid bars = padj < 0.05, transparent = NS",
    x        = "Log2 Fold Change",
    y        = "MetaCyc Pathway",
    fill     = ""
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_text(size = 7, family = "serif"),
    text        = element_text(family = "Times New Roman", size = 14)
  )
```

```{r pairwise_foldchange_1v3, fig.height=8}
path_1v3 %>%
  filter(!is.na(log2FoldChange)) %>%
  arrange(desc(abs(log2FoldChange))) %>%
  slice_head(n = 20) %>%
  left_join(pathway_names, by = "pathway") %>%
  mutate(
    label       = ifelse(!is.na(name), name, pathway),
    label       = reorder(label, log2FoldChange),
    direction   = ifelse(log2FoldChange > 0, "Higher in MIX3",
                                             "Higher in MIX1"),
    significant = padj < 0.05
  ) %>%
  ggplot(aes(x = log2FoldChange, y = label, fill = direction,
              alpha = significant)) +
  geom_col() +
  geom_errorbarh(
    aes(xmin = log2FoldChange - lfcSE,
        xmax = log2FoldChange + lfcSE),
    height = 0.3
  ) +
  scale_fill_manual(values = c(
    "Higher in MIX3" = "mediumseagreen",
    "Higher in MIX1" = "coral2"
  )) +
  scale_alpha_manual(values = c("TRUE" = 1.0, "FALSE" = 0.4),
                     guide = "none") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title    = "Top 20 Pathways by Effect Size — MIX3 vs MIX1",
    subtitle = "Solid bars = padj < 0.05, transparent = NS",
    x        = "Log2 Fold Change",
    y        = "MetaCyc Pathway",
    fill     = ""
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_text(size = 7, family = "serif"),
    text        = element_text(family = "Times New Roman", size = 14)
  )
```

## Dose-Response Pathway Visualization
```{r significant_pathways_MetaCyc, fig.height=10}
sig_paths <- lrt_path %>%
  filter(padj < 0.05) %>%
  pull(pathway)

cat("Significant dose-responsive pathways:\n")
print(sig_paths)

if (length(sig_paths) > 0) {
  sig_path_df <- as.data.frame(path_abun[sig_paths, ]) %>%
    rownames_to_column("pathway") %>%
    pivot_longer(
      cols      = -pathway,
      names_to  = "sample_id",
      values_to = "abundance"
    ) %>%
    left_join(metadata %>% select(sample_id, group),
              by = "sample_id") %>%
    left_join(pathway_names, by = "pathway") %>%
    mutate(
      group = factor(group, levels = c("MIX1", "MIX2", "MIX3")),
      label = ifelse(!is.na(name), name, pathway)
    )

  ggplot(sig_path_df, aes(x = group, y = log10(abundance + 1),
                           color = group)) +
    geom_point(size = 2, alpha = 0.6) +
    stat_summary(fun = mean, geom = "line",
                 aes(group = 1), color = "black", linewidth = 1) +
    stat_summary(fun = mean, geom = "point",
                 size = 3, color = "black") +
    scale_color_manual(values = mix_colors) +
    facet_wrap(~ label, scales = "free_y", ncol = 3) +
    labs(
      title = "Dose-Responsive MetaCyc Pathways (DESeq2 LRT, padj < 0.05)",
      x     = "Treatment Group",
      y     = "log10(Predicted Abundance + 1)",
      color = "Group"
    ) +
    theme_bw() +
    theme(
      legend.position = "none",
      axis.text.x     = element_text(angle = 45, hjust = 1),
      strip.text      = element_text(size = 7, family = "serif"),
      text            = element_text(family = "Times New Roman", size = 14)
    )
} else {
  cat("No significant dose-responsive pathways detected\n")
}
```


### Sig Pathways with classifications

```{r class-sig-path-vol}
pwy_classes <- tibble::tribble(
  ~pathway,              ~pwy_class,
  "ARGSYN-PWY",          "Amino Acid Biosynthesis",
  "THRESYN-PWY",         "Amino Acid Biosynthesis",
  "PWY-2941",            "Amino Acid Biosynthesis",
  "PWY-2942",            "Amino Acid Biosynthesis",
  "DAPLYSINESYN-PWY",    "Amino Acid Biosynthesis",
  "PWY-6163",             "Aromatic Compound Biosynthesis",

  "PWY-7953",            "Nucleotide Biosynthesis (Purine)",
  "PWY-6124",            "Nucleotide Biosynthesis (Purine)",
  "PWY-6126",            "Nucleotide Biosynthesis (Purine)",
  "PWY-7229",            "Nucleotide Biosynthesis (Purine)",
  "PWY-7220",            "Nucleotide Biosynthesis (Purine)",
  "PWY-7222",            "Nucleotide Biosynthesis (Purine)",
  "PWY-7228",            "Nucleotide Biosynthesis (Purine)",
  "PWY-6121",            "Nucleotide Biosynthesis (Purine)",
  "PWY-6151",            "Nucleotide Biosynthesis (Purine)",
  "PWY-7221",            "Nucleotide Biosynthesis (Purine)",
  "PWY-6122",            "Nucleotide Biosynthesis (Purine)",
  "PWY-6277",            "Nucleotide Biosynthesis (Purine)",

  "PWY-6387",            "Nucleotide Biosynthesis (Pyrimidine)",
  "PWY-7790",            "Nucleotide Biosynthesis (Pyrimidine)",
  "PWY-7791",            "Nucleotide Biosynthesis (Pyrimidine)",
  "PWY-7208",            "Nucleotide Biosynthesis (Pyrimidine)",
  "PWY-5686",            "Nucleotide Biosynthesis (Pyrimidine)",

  "THISYNARA-PWY",       "Cofactor Biosynthesis",
  "PWY-6897",            "Cofactor Biosynthesis",
  "PWY-7357",            "Cofactor Biosynthesis",
  "COA-PWY-1",           "Cofactor Biosynthesis",
  "COA-PWY",             "Cofactor Biosynthesis",
  "PWY-5918",            "Cofactor Biosynthesis",

  "PWY4FS-7",            "Lipid Biosynthesis",
  "PWY4FS-8",            "Lipid Biosynthesis",
  "PHOSLIPSYN-PWY",      "Lipid Biosynthesis",
  "PWY-5667",            "Lipid Biosynthesis",
  "PWY0-1319",           "Lipid Biosynthesis",

  "PWY0-1586",           "Cell Envelope Biosynthesis",
  "UDPNAGSYN-PWY",       "Cell Envelope Biosynthesis",

  "PWY-7851",            "Carbohydrate Metabolism",
  "GLYCOLYSIS",          "Carbohydrate Metabolism",
  "PWY-1042",            "Carbohydrate Metabolism",
  "PWY-622",             "Carbohydrate Metabolism",
  "PWY-5659",            "Carbohydrate Metabolism",

  "POLYISOPRENSYN-PWY",  "Secondary Metabolite Biosynthesis"
)


unclassified <- setdiff(sig_paths, pwy_classes$pathway)
if (length(unclassified) > 0) {
  cat("WARNING — pathways with no class assigned:\n")
  print(unclassified)
} else {
  cat("All", length(sig_paths), "significant pathways have a class assignment\n")
}

lrt_path_classed <- lrt_path %>%
  left_join(pwy_classes, by = "pathway") %>%
  left_join(pathway_names, by = "pathway") %>%
  mutate(
    label     = ifelse(!is.na(name), name, pathway),
    direction = ifelse(log2FoldChange > 0, "Up with dose", "Down with dose")
  )

write.csv(lrt_path_classed,
          file.path(data_dir, "pathway_DA_dose_response_LRT_classed.csv"),
          row.names = FALSE)

class_summary <- lrt_path_classed %>%
  filter(padj < 0.05) %>%
  group_by(pwy_class) %>%
  summarise(
    n_pathways  = n(),
    n_up        = sum(log2FoldChange > 0),
    n_down      = sum(log2FoldChange < 0),
    mean_lfc    = mean(log2FoldChange),
    median_padj = median(padj),
    .groups = "drop"
  ) %>%
  arrange(median_padj)

cat("\nDose-response direction by MetaCyc class (significant pathways only):\n")
print(class_summary)

# --- ADDED: forest/dot plot, ordered by class, showing per-pathway trend
# with SE and significance, so the class-level coherence is visible without
# reading 42 individual rows ---
sig_classed <- lrt_path_classed %>% filter(padj < 0.05)

ggplot(sig_classed, aes(x = log2FoldChange,
                         y = reorder(label, log2FoldChange))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = log2FoldChange - lfcSE,
                       xmax = log2FoldChange + lfcSE),
                   color = "violetred4", size = 0.35) +
  facet_grid(pwy_class ~ ., scales = "free_y", space = "free_y") +
  labs(
    x = "log2FC per log10(dose)",
    y = NULL,
    title = "Dose-response trend by MetaCyc pathway class",
    subtitle = "DESeq2 LRT, padj < 0.05"
  ) +
  theme_bw(base_size = 10) +
  theme(
    strip.text.y = element_text(angle = 0, hjust = 0, size = 7),
    axis.text.y  = element_text(size = 6, family = "serif"),
    text         = element_text(family = "Times New Roman", size = 12)
  )
```


## Significant Pathway Heatmap
```{r significant_pathway_heatmap, fig.height=8}
if (length(sig_paths) > 0) {
  sig_path_mat <- log10(path_abun[sig_paths, ] + 1)

  rownames(sig_path_mat) <- ifelse(
    rownames(sig_path_mat) %in% pathway_names$pathway,
    pathway_names$name[match(rownames(sig_path_mat),
                              pathway_names$pathway)],
    rownames(sig_path_mat)
  )

  pheatmap(
    as.matrix(sig_path_mat),
    annotation_col    = sample_annot,
    annotation_colors = list(group = mix_colors),
    cluster_rows      = TRUE,
    cluster_cols      = FALSE,
    show_rownames     = TRUE,
    show_colnames     = TRUE,
    fontsize_row      = 10,
    fontsize_col      = 10,
    fontfamily        = "serif",
    color             = colorRampPalette(
                          c("white", "cadetblue3", "darkblue"))(50),
    main              = "Significant Dose-Responsive Pathways (padj < 0.05)"
  )
} else {
  cat("No significant pathways to plot\n")
}
```

## Session Info
```{r session_info}
sessionInfo()
```
