---
title: "Lipid Profile Blood Analysis"
author: "Chloe J McGovern"
date: "`r Sys.Date()`"
output:
  html_document:
    css: "styles.css"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "Figures/02_BloodLipid")
```

# Set up the environment!

```{r load_packages-font}

#Load packages
#install.packages("patchwork")
pacman::p_load(tidyverse, patchwork, install = FALSE)

#load times new roman font
windowsFonts("Times New Roman" = windowsFont("Times New Roman"))

theme(
    text = element_text(family = "Times New Roman")
)

```

# Lipid Profile in Blood

Blood was taken from the heart after euthanasia. Due to the size of the birds, two birds blood was pooled together into one tube. So all individual data points is a pooling of two birds blood. 

## Load in the data

```{r load-data}
mix_lipid_raw <- read.csv("Data/CSVs/lipid-blood-mix.csv")

glimpse(mix_lipid_raw)

source("C:/Users/Chloe/OneDrive - Cornell University/study-1-march25/MIX/Code/colors.R")
str(treatment_colors)

```

## Let's run some statistics

```{r anova-tukey}
# ANOVA for HDL
anova_hdl <- aov(hdl ~ treatment, data = mix_lipid_raw)
summary(anova_hdl)

# Tukey for HDL
TukeyHSD(anova_hdl)

# extract p-value for HDL
hdl_p <- summary(anova_hdl)[[1]]$'Pr(>F)'[1]
hdl_p_text <- paste0("ANOVA p = ", signif(hdl_p, 3))

# ANOVA for LDL
anova_ldl <- aov(ldl ~ treatment, data = mix_lipid_raw)
summary(anova_ldl)

# Tukey for LDL
TukeyHSD(anova_ldl)

# extract p-value for LDL
ldl_p <- summary(anova_ldl)[[1]]$'Pr(>F)'[1]
ldl_p_text <- paste0("ANOVA p = ", signif(ldl_p, 3))

# ANOVA for Cholesterol
anova_chol <- aov(chol ~ treatment, data = mix_lipid_raw)
summary(anova_chol)

# Tukey for Cholesterol
TukeyHSD(anova_chol)

# extract p-value for HDL
chol_p <- summary(anova_chol)[[1]]$'Pr(>F)'[1]
chol_p_text <- paste0("ANOVA p = ", signif(chol_p, 3))

```

There was no significant difference in HDL, LDL, or cholesterol values. 

## Time to make a figure

```{r treatment-HDL}

# make treatment factor with custom labels and order
mix_lipid_labs <- mix_lipid_raw %>%
  mutate(treatment = factor(treatment,
    levels = c("no_inj", "water", "mix_1", "mix_2", "mix_3"),
    labels = c("No Injection", "Water", "Mix Dose 1", "Mix Dose 2", "Mix Dose 3")
  ))

# Check treatment levels
levels(mix_lipid_labs$treatment)

# plot it
f1 <- ggplot(mix_lipid_labs, aes(x = treatment, y = hdl)) +
    geom_boxplot(aes(fill = treatment), outlier.shape = NA, color = "black", alpha = 0.5, width = 0.5) +  # Lighter box plot with treatment colors
    geom_jitter(aes(color = treatment), width = 0.15, alpha = 0.7, size = 2) +
    scale_color_manual(values = treatment_colors) +  # Use the custom treatment colors
    scale_fill_manual(values = treatment_colors) +  # Use the custom treatment colors for fill
    theme_minimal(base_size = 14) +
    expand_limits(y = max(mix_lipid_labs$hdl, na.rm = TRUE) * 1.08) +
    annotate("text", x = 1, y = max(mix_lipid_labs$hdl, na.rm = TRUE) * 1.05,
             label = hdl_p_text, hjust = 0, size = 4) +
    labs(
        title = "Figure A: HDL Levels by Treatment",
        y = "HDL (mg/dL)",
        x = NULL
    ) +
    theme(
        text = element_text(family = "Times New Roman"),
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.title = element_text(face = "bold"),
        axis.text = element_text(size = 12, color = "black"),
        legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)
    )

f1
```


```{r treatment-LDL}

# plot it
f2 <- ggplot(mix_lipid_labs, aes(x = treatment, y = ldl)) +
    geom_boxplot(aes(fill = treatment), outlier.shape = NA, color = "black", alpha = 0.5, width = 0.5) +  # Lighter box plot with treatment colors
    geom_jitter(aes(color = treatment), width = 0.15, alpha = 0.7, size = 2) +
    scale_color_manual(values = treatment_colors) +  # Use the custom treatment colors
    scale_fill_manual(values = treatment_colors) +  # Use the custom treatment colors for fill
    theme_minimal(base_size = 14) +
    expand_limits(y = max(mix_lipid_labs$ldl, na.rm = TRUE) * 1.08) +
    annotate("text", x = 1, y = max(mix_lipid_labs$ldl, na.rm = TRUE) * 1.05,
             label = ldl_p_text, hjust = 0, size = 4) +
    labs(
        title = "Figure B: LDL Levels by Treatment",
        y = "LDL (mg/dL)",
        x = NULL
    ) +
    theme(
        text = element_text(family = "Times New Roman"),
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.title = element_text(face = "bold"),
        axis.text = element_text(size = 12, color = "black"),
        legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)
    )

f2
```


```{r treatment-chol}
# plot it
f3 <- ggplot(mix_lipid_labs, aes(x = treatment, y = chol)) +
    geom_boxplot(aes(fill = treatment), outlier.shape = NA, color = "black", alpha = 0.5, width = 0.5) +  # Lighter box plot with treatment colors
    geom_jitter(aes(color = treatment), width = 0.15, alpha = 0.7, size = 2) +
    scale_color_manual(values = treatment_colors) +  # Use the custom treatment colors
    scale_fill_manual(values = treatment_colors) +  # Use the custom treatment colors for fill
    theme_minimal(base_size = 14) +
    expand_limits(y = max(mix_lipid_labs$chol, na.rm = TRUE) * 1.08) +
    annotate("text", x = 1, y = max(mix_lipid_labs$chol, na.rm = TRUE) * 1.05,
             label = hdl_p_text, hjust = 0, size = 4) +
    labs(
        title = "Figure C: Cholesterol Levels by Treatment",
        y = "Cholesterol (mg/dL)",
        x = NULL
    ) +
    theme(
        text = element_text(family = "Times New Roman"),
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.title = element_text(face = "bold"),
        axis.text = element_text(size = 12, color = "black"),
        legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)
    )


f3
```

### Final Figure

Let's put it into one three panel graph 

```{r}
# f1 = HDL plot
# f2 = LDL plot
# f3 = cholesterol plot

# Combine horizontally into a 3-panel figure
combined_plot <- (f1 | f2 | f3) + 
  plot_annotation(title = "Lipid Profile by Treatment")

combined_plot

ggsave("Figures/02_BLoodLipid/lipid_panel.png", plot = combined_plot, width = 14, height = 5, dpi = 600)
```

