---
title: "Lipid Profile Blood Analysis"
author: "Chloe J McGovern"
date: "`r Sys.Date()`"
output:
  html_document:
    css: "styles.css"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "../Figures/02_BloodLipid")
```

# Set up the environment!

```{r load_packages-font}

#Load packages
#install.packages("patchwork")
pacman::p_load(tidyverse, patchwork, install = FALSE)

#load times new roman font
windowsFonts("Times New Roman" = windowsFont("Times New Roman"))

theme(
    text = element_text(family = "Times New Roman")
)

```

# Lipid Profile in Blood

Blood was taken from the heart after euthanasia. Due to the size of the birds, two birds blood was pooled together into one tube. So all individual data points is a pooling of two birds blood. 

## Load in the data

```{r load-data}
mix_lipid_raw <- read.csv("Data/CSVs/lipid-blood-mix.csv")

glimpse(mix_lipid_raw)

```

## Let's run some statistics

```{r anova-tukey}
# ANOVA for HDL
anova_hdl <- aov(hdl ~ treatment, data = mix_lipid_raw)
summary(anova_hdl)

# Tukey for HDL
TukeyHSD(anova_hdl)

# extract p-value for HDL
hdl_p <- summary(anova_hdl)[[1]]$'Pr(>F)'[1]
hdl_p_text <- paste0("ANOVA p = ", signif(hdl_p, 3))

# ANOVA for LDL
anova_ldl <- aov(ldl ~ treatment, data = mix_lipid_raw)
summary(anova_ldl)

# Tukey for LDL
TukeyHSD(anova_ldl)

# extract p-value for LDL
ldl_p <- summary(anova_ldl)[[1]]$'Pr(>F)'[1]
ldl_p_text <- paste0("ANOVA p = ", signif(ldl_p, 3))

# ANOVA for Cholesterol
anova_chol <- aov(chol ~ treatment, data = mix_lipid_raw)
summary(anova_chol)

# Tukey for Cholesterol
TukeyHSD(anova_chol)

# extract p-value for HDL
chol_p <- summary(anova_chol)[[1]]$'Pr(>F)'[1]
chol_p_text <- paste0("ANOVA p = ", signif(chol_p, 3))

```

There was no significant difference in HDL, LDL, or cholesterol values. 

## Time to make a figure

```{r treatment-HDL}

# make treatment factor with custom labels and order
mix_lipid_labs <- mix_lipid_raw %>%
  mutate(treatment = factor(treatment,
    levels = c("no_inj", "water", "mix_1", "mix_2", "mix_3"),
    labels = c("No Injection", "Water", "Mix Dose 1", "Mix Dose 2", "Mix Dose 3")
  ))

# plot it
f1 <- ggplot(mix_lipid_labs, aes(x = treatment, y = hdl)) +
    geom_boxplot(outlier.shape = NA, fill = "gray90") +
    geom_jitter(width = 0.15, alpha = 0.7, size = 2, aes(color = treatment)) +
    scale_color_brewer(palette = "Set2") +
    theme_minimal(base_size = 14) +
    expand_limits(y = max(mix_lipid_labs$hdl, na.rm = TRUE) * 1.08) + # gives room for p-value
    annotate("text", x = 1, y = max(mix_lipid_labs$hdl, na.rm = TRUE) * 1.05,
             label = hdl_p_text, hjust = 0, size = 4) +
    labs(
        title = "Figure A: HDL Levels by Treatment",
        y = "HDL (mg/dL)",
        x = NULL
    ) +
    theme(
        text = element_text(family = "Times New Roman"),
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.title = element_text(face = "bold"),
        axis.text = element_text(size = 12, color = "black"),
        legend.position = "none"
    )

f1
```


```{r treatment-LDL}

# plot it
f2 <- ggplot(mix_lipid_labs, aes(x = treatment, y = ldl)) +
    geom_boxplot(outlier.shape = NA, fill = "gray90") +
    geom_jitter(width = 0.15, alpha = 0.7, size = 2, aes(color = treatment)) +
    scale_color_brewer(palette = "Set2") +
    theme_minimal(base_size = 14) +
    expand_limits(y = max(mix_lipid_labs$ldl, na.rm = TRUE) * 1.08) + # gives room for p-value
    annotate("text", x = 1, y = max(mix_lipid_labs$ldl, na.rm = TRUE) * 1.05,
             label = ldl_p_text, hjust = 0, size = 4) +
    labs(
        title = "Figure B: LDL Levels by Treatment",
        y = "LDL (mg/dL)",
        x = NULL
    ) +
    theme(
        text = element_text(family = "Times New Roman"),
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.title = element_text(face = "bold"),
        axis.text = element_text(size = 12, color = "black"),
        legend.position = "none"
    )

f2
```


```{r treatment-chol}
# plot it
f3 <- ggplot(mix_lipid_labs, aes(x = treatment, y = chol)) +
    geom_boxplot(outlier.shape = NA, fill = "gray90") +
    geom_jitter(width = 0.15, alpha = 0.7, size = 2, aes(color = treatment)) +
    scale_color_brewer(palette = "Set2") +
    theme_minimal(base_size = 14) +
    expand_limits(y = max(mix_lipid_labs$chol, na.rm = TRUE) * 1.08) + # gives room for p-value
    annotate("text", x = 1, y = max(mix_lipid_labs$chol, na.rm = TRUE) * 1.05,
             label = chol_p_text, hjust = 0, size = 4) +
    labs(
        title = "Figure C: Cholesterol Levels by Treatment",
        y = "Cholesterol (mg/dL)",
        x = NULL
    ) +
    theme(
        text = element_text(family = "Times New Roman"),
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.title = element_text(face = "bold"),
        axis.text = element_text(size = 12, color = "black"),
        legend.position = "none"
    )

f3
```

### Final Figure

Let's put it into one three panel graph 

```{r}
# f1 = HDL plot
# f2 = LDL plot
# f3 = cholesterol plot

# Combine horizontally into a 3-panel figure
(f1 | f2 | f3) + 
  plot_annotation(title = "Lipid Profile by Treatment")
```

