---
title: "PCR Analysis for MIX Manuscript of Duodenum, Colon, Pectoral, and Liver"
author: "Chloe J McGovern"
date: "`r Sys.Date()`"
output:
  html_document:
    css: "styles.css"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                        fig.path = "Figures/04_PCRAnalysis/")
```

# PCR Analysis

Specific genes were analyzed for gene expression due to probiotics. 

# Load in data and any other necessities

```{r load-packages}

#install.packages("rtpcr", dependencies = TRUE, type = "source")
#library(rtpcr)
#install.packages("broom")
#install.packages("FSA")

pacman::p_load(tidyverse, ggplot2, dplyr, patchwork, 
               rtpcr, ggsignif, car, broom, FSA, rstatix, install = FALSE)

#load times new roman font
windowsFonts("Times New Roman" = windowsFont("Times New Roman"))

theme(
    text = element_text(family = "Times New Roman")
)

setwd("C:/Users/Chloe/OneDrive - Cornell University/study-1-march25/MIX")

mix_pcr <- read.csv("Data/CSVs/pcr-data-mix.csv")

glimpse(mix_pcr)

source("Code/colors.R")
```

## Method of analysis 

Raw Ct values were calculated to the Pfaffl method which accounts for efficency curves created during optimization. Link to the original manuscript here: https://pubmed.ncbi.nlm.nih.gov/11328886/. 

# Tissue

## Duodenum

Let's get to work...

### Pfaffl Analysis

```{r duo-pfaffl}

# STEP 1: Preprocess and average replicates
duo_avg <- mix_pcr %>%
  filter(Tissue == "Duodenum") %>%
  rowwise() %>%
  mutate(Ct = mean(c_across(Rep1:Rep2), na.rm = TRUE)) %>%
  ungroup() %>%
  select(Sample, Treatment, Gene, Ct)

# STEP 2: Join in efficiency values
duo_eff <- tibble(
  Gene = c("GAPDH", "OCLN", "MUC2", "TNF-alpha"),
  Efficiency = c(2.02, 1.93, 1.90, 1.90)
)

duo_data <- left_join(duo_avg, duo_eff, by = "Gene")

glimpse(duo_data)

# STEP 3: Split reference and target genes
# (We'll loop over target genes vs GAPDH reference)

reference_gene <- "GAPDH"
control_group <- "Water"

genes_to_test <- c("MUC2", "OCLN", "TNF-alpha")

# STEP 4: Create fold change calculator function using Pfaffl formula
pfaffl_fc <- function(df, target_gene, ref_gene, control_group) {
  # Subset target and reference data
  target_df <- df %>% filter(Gene == target_gene)
  ref_df    <- df %>% filter(Gene == ref_gene)

  # Join on sample and treatment
  joined <- inner_join(target_df, ref_df, by = c("Sample", "Treatment"), suffix = c("_target", "_ref"))

  # Calculate ΔCt
  joined <- joined %>%
    mutate(delta_Ct_target = Ct_target - Ct_ref)

  # Pre-calculate mean ΔCt of control group
  control_delta_mean <- joined %>%
    filter(Treatment == control_group) %>%
    summarise(mean_delta_Ct = mean(delta_Ct_target, na.rm = TRUE)) %>%
    pull(mean_delta_Ct)

  # Calculate ΔΔCt and fold change
  joined <- joined %>%
    mutate(
      delta_delta_Ct = delta_Ct_target - control_delta_mean,
      fold_change = (Efficiency_target ^ -delta_delta_Ct) / (Efficiency_ref ^ -delta_delta_Ct)
    )

  return(joined)
}


# STEP 5: Apply function for each target gene
# STEP 5: Apply function for each target gene (with explicit arguments)
duo_results <- map_dfr(genes_to_test, ~pfaffl_fc(duo_data, .x, ref_gene = "GAPDH", control_group = "Water"), .id = "GeneIndex") %>%
  mutate(
    log2_fc = log2(fold_change),
    Gene = factor(Gene_target, levels = c("MUC2", "OCLN", "TNF-alpha"))  # Adjust if needed
  )


# STEP 6: Plot fold change

duo_results <- duo_results %>%
  select(-Gene) %>%                 # remove the existing Gene column
  rename(Gene = Gene_target)       # now safely rename Gene_target to Gene

duo_nosig <- ggplot(duo_results, aes(x = Treatment, y = fold_change, fill = Treatment)) +
  geom_boxplot(alpha = 0.5) +
  geom_jitter(width = 0.2, size = 2) +
  facet_wrap(~ Gene, scales = "free_y") +
  labs(title = "Duodenum: Pfaffl-Calculated Fold Change",
       y = "Relative Expression (Fold Change)",
       x = "Treatment") +
  theme_minimal(base_family = "Times New Roman")

duo_nosig

```

YAY! We calculated the Pfaffl Fold Change, now time to do some statistics and make this figure look manuscript ready like the other analyses!

### Statistics

```{r stat-assumptions-duo}

# Check ANOVA assumptions

# STEP 1: Shapiro-Wilk test for normality (per gene)
shapiro_results <- duo_results %>%
  group_by(Gene, Treatment) %>%
  summarise(
    shapiro_p = shapiro.test(fold_change)$p.value,
    .groups = "drop"
  )

shapiro_results #p > 0.05 normality assumption met for tnf 


# STEP 2: Levene's Test for equal variances (per gene)
levene_results <- duo_results %>%
  group_by(Gene) %>%
  group_map(~car::leveneTest(fold_change ~ Treatment, data = .x) %>% 
              tidy() %>%
              mutate(Gene = .y$Gene),
            .keep = TRUE) %>%
  bind_rows()

levene_results # p >0.05 for all

```

Since only MUC2 passed ANOVA assumptions, we will run...
  1. ANOVA for TNFa
  2. Kruskal Wallis for MUC2 and OCLN
  
```{r stats-final-duo}

# MUC2 and KW
# Filter

muc2_data <- filter(duo_results, Gene == "MUC2")
kw_muc2 <- kruskal.test(log2_fc ~ Treatment, data = muc2_data)
kw_muc2
muc2_p <- kw_muc2$p.value
muc2_p

# Dunn's Test
dunn_muc2 <- dunnTest(log2_fc ~ Treatment, data = muc2_data, method = "holm")
muc2_pairs <- dunn_muc2$res
muc2_sig <- muc2_pairs %>%
  filter(P.adj < 0.05) %>%
  separate(Comparison, into = c("group1", "group2"), sep = " - ") %>%
  mutate(
    label = ifelse(P.adj < 0.01, "**", "*"),
    y.position = max(muc2_data$log2_fc, na.rm = TRUE) + 0.2,
    Gene = "MUC2"
  )

dunn_muc2
muc2_sig


# OCLN KW
ocln_data <- filter(duo_results, Gene == "OCLN")
kw_ocln <- kruskal.test(log2_fc ~ Treatment, data = ocln_data)
kw_ocln
ocln_p <- kw_ocln$p.value
ocln_p

# Dunn's Test
dunn_ocln <- dunnTest(log2_fc ~ Treatment, data = ocln_data, method = "holm")
ocln_pairs <- dunn_ocln$res
ocln_sig <- ocln_pairs %>%
  filter(P.adj < 0.05) %>%
  separate(Comparison, into = c("group1", "group2"), sep = " - ") %>%
  mutate(
    label = ifelse(P.adj < 0.01, "**", "*"),
    y.position = max(ocln_data$log2_fc, na.rm = TRUE) + 0.2,
    Gene = "OCLN"
  )

dunn_ocln
ocln_sig

# TNFA ANOVA

tnf_data <- filter(duo_results, Gene == "TNF-alpha")

# ANOVA
tnf_aov <- aov(log2_fc ~ Treatment, data = tnf_data)
tnf_p <- summary(tnf_aov)[[1]][["Pr(>F)"]][1]
tnf_p

# Tukey HSD
tukey_tnf <- TukeyHSD(tnf_aov)
tnf_pairs <- as.data.frame(tukey_tnf$Treatment)
tnf_pairs$comparison <- rownames(tnf_pairs)
tnf_sig <- tnf_pairs %>%
  filter(`p adj` < 0.05) %>%
  separate(comparison, into = c("group1", "group2"), sep = "-") %>%
  mutate(
    label = ifelse(`p adj` < 0.01, "**", "*"),
    y.position = max(tnf_data$log2_fc, na.rm = TRUE) + 0.2,
    Gene = "TNF-alpha"
  )
tnf_sig

```
  
### Manuscript Ready Duodenum Figure

#### MUC2

```{r muc-2-fig-duo}

# Plot for MUC2

# Ensure treatment factor order
muc2_data <- muc2_data %>%
  mutate(Treatment = factor(Treatment, levels = c("No Injection", "Water", "Mix Dose 1", "Mix Dose 2", "Mix Dose 3")))

# Define comparison pairs (order matters!)
muc2_comparisons <- list(
  c("Water", "Mix Dose 2"),
  c("Mix Dose 2" , "Water")
)

muc2_sig <- muc2_sig %>%
  mutate(
    group1 = factor(group1, levels = levels(muc2_data$Treatment)),
    group2 = factor(group2, levels = levels(muc2_data$Treatment))
  )

max_y <- max(muc2_data$log2_fc, na.rm = TRUE)
muc2_sig <- muc2_sig %>%
  mutate(y.position = max_y + 0.05 + (row_number() * 0.03))  # stagger brackets

levels(muc2_data$Treatment)
levels(muc2_sig$group1)

muc2_sig <- muc2_sig %>%
  mutate(across(c(group1, group2), ~factor(.x, levels = levels(muc2_data$Treatment))))

muc2_sig <- muc2_sig %>%
  mutate(
    group1 = factor(group1, levels = levels(muc2_data$Treatment)),
    group2 = factor(group2, levels = levels(muc2_data$Treatment)),
    y.position = max(muc2_data$log2_fc, na.rm = TRUE) + 0.05 + (row_number() * 0.03)
  )

muc2_sig <- muc2_sig %>%
  mutate(
    y.position = c(0.22, 0.25)  # lower than before
  )

# Plot using your working style
p_muc2 <- ggplot(muc2_data, aes(x = Treatment, y = log2_fc, fill = Treatment)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA, color = "black") +
  geom_jitter(aes(color = Treatment), width = 0.15, size = 1.8, alpha = 0.8) +
  scale_fill_manual(values = treatment_colors) +
  scale_color_manual(values = treatment_colors) +
  theme_minimal(base_size = 14) +
  theme(
    text = element_text(family = "Times New Roman"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    legend.position = "none"
  ) +
  labs(
    title = "MUC2",
    x = NULL,
    y = expression(log[2]*"(Fold Change)")
  ) +
  ggsignif::geom_signif(
    data = muc2_sig,
    aes(
      xmin = group1,
      xmax = group2,
      annotations = label,
      y_position = y.position
    ),
    inherit.aes = FALSE,
    manual = TRUE,
    tip_length = 0.01,
    textsize = 4,
    vjust = 0.3
  ) +
  annotate("text", x = 3, y = 0.3,
           label = "KW~italic(p) == 0.0009", parse = TRUE, size = 4.5) +
  coord_cartesian(ylim = c(-0.1, 0.3))

# Show the plot

p_muc2

```

#### OCLN

```{r ocln-fig-duo}

# Plot for OCLN

# Ensure treatment factor levels
ocln_data <- duo_results %>%
  filter(Gene == "OCLN") %>%
  mutate(Treatment = factor(Treatment, levels = c("No Injection", "Water", "Mix Dose 1", "Mix Dose 2", "Mix Dose 3")))

# Define comparison pairs (order matters!)
ocln_comparisons <- list(
  c("Mix Dose 2", "Mix Dose 3")
)

ocln_sig <- ocln_sig %>%
  mutate(
    group1 = factor(group1, levels = levels(ocln_data$Treatment)),
    group2 = factor(group2, levels = levels(ocln_data$Treatment))
  )

max_y <- max(ocln_data$log2_fc, na.rm = TRUE)
ocln_sig <- ocln_sig %>%
  mutate(y.position = max_y + 0.05 + (row_number() * 0.03))  # stagger brackets

levels(ocln_data$Treatment)
levels(ocln_sig$group1)

ocln_sig <- ocln_sig %>%
  mutate(across(c(group1, group2), ~factor(.x, levels = levels(ocln_data$Treatment))))

ocln_sig <- ocln_sig %>%
  mutate(
    group1 = factor(group1, levels = levels(ocln_data$Treatment)),
    group2 = factor(group2, levels = levels(ocln_data$Treatment)),
    y.position = max(ocln_data$log2_fc, na.rm = TRUE) + 0.05 + (row_number() * 0.03)
  )

ocln_sig <- ocln_sig %>%
  mutate(
    y.position = c(0.1)  # lower than before
  )


# Plot
p_ocln <- ggplot(ocln_data, aes(x = Treatment, y = log2_fc, fill = Treatment)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA, color = "black") +
  geom_jitter(aes(color = Treatment), width = 0.15, size = 1.8, alpha = 0.8) +
  scale_fill_manual(values = treatment_colors) +
  scale_color_manual(values = treatment_colors) +
  theme_minimal(base_size = 14) +
  theme(
    text = element_text(family = "Times New Roman"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    legend.position = "none"
  ) +
  labs(
    title = "OCLN",
    x = NULL,
    y = expression(log[2]*"(Fold Change)")
  ) +
  ggsignif::geom_signif(
    comparisons = ocln_comparisons,
    annotations = "*",        # or "**", depending on significance
    y_position = 0.12,
    tip_length = 0.01,
    textsize = 4,
    vjust = 0.3
  ) +
  annotate("text", x = 3, y = 0.15,
           label = "KW~italic(p) == 0.015", parse = TRUE, size = 4.5) +
  coord_cartesian(ylim = c(-0.15, 0.15))

# Show plot
p_ocln

```


#### TNFa

```{r tnfa-fig-duo}

# Plot for TNFa

# Filter and factor levels
tnf_data <- duo_results %>%
  filter(Gene == "TNF-alpha") %>%
  mutate(Treatment = factor(Treatment, levels = c("No Injection", "Water", "Mix Dose 1", "Mix Dose 2", "Mix Dose 3")))

tnf_comparisons <- list(
  c("Mix Dose 3", "Mix Dose 2"),
  c("No Injection", "Mix Dose 2")
)


# Plot
p_tnf <- ggplot(tnf_data, aes(x = Treatment, y = log2_fc, fill = Treatment)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA, color = "black") +
  geom_jitter(aes(color = Treatment), width = 0.15, size = 1.8, alpha = 0.8) +
  scale_fill_manual(values = treatment_colors) +
  scale_color_manual(values = treatment_colors) +
  theme_minimal(base_size = 14) +
  theme(
    text = element_text(family = "Times New Roman"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    legend.position = "none"
  ) +
  labs(
    title = "TNF-α",
    x = NULL,
    y = expression(log[2]*"(Fold Change)")
  ) +
  ggsignif::geom_signif(
    comparisons = tnf_comparisons,
    annotations = "*",        # or "**", depending on significance
    y_position = seq(0.3, 0.35, by = 0.03),
    tip_length = 0.01,
    textsize = 4,
    vjust = 0.3
  ) +
  annotate("text", x = 3, y = 0.4,
           label = "ANOVA~italic(p) == 0.015", parse = TRUE, size = 4.5) +
  coord_cartesian(ylim = c(-0.2, 0.4))

# Show the plot
p_tnf

```


#### All Final Figures

```{r final-manuscript-fig-duo, fig.width=6.5, fig.height=6, dpi=300}

duo_p <- (p_muc2 | p_ocln | p_tnf) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Duodenum Gene Expression",
    theme = theme(
      plot.title = element_text(family = "Times New Roman", size = 16, face = "bold", hjust = 0.5)
    )
  )

duo_p

ggsave("final_plot_duo_all_mix.png", plot = duo_p, 
       width = 10, height = 6, units = "in", dpi = 600)

```

YAY! 

## Colon

### Pfaffl Analysis

```{r col-pfaffl}

# STEP 1: Preprocess and average replicates
col_avg <- mix_pcr %>%
  filter(Tissue == "Colon") %>%
  rowwise() %>%
  mutate(Ct = mean(c_across(Rep1:Rep2), na.rm = TRUE)) %>%
  ungroup() %>%
  select(Sample, Treatment, Gene, Ct)

# STEP 2: Join in efficiency values
col_eff <- tibble(
  Gene = c("GAPDH", "OCLN", "MUC2", "TNF-alpha"),
  Efficiency = c(1.93, 1.9, 1.96, 2.1)
)

col_data <- left_join(col_avg, col_eff, by = "Gene")

glimpse(col_data)

# STEP 3: Split reference and target genes
# (We'll loop over target genes vs GAPDH reference)

reference_gene <- "GAPDH"
control_group <- "Water"

genes_to_test <- c("MUC2", "OCLN", "TNF-alpha")

# STEP 4: Create fold change calculator function using Pfaffl formula
pfaffl_fc <- function(df, target_gene, ref_gene, control_group) {
  # Subset target and reference data
  target_df <- df %>% filter(Gene == target_gene)
  ref_df    <- df %>% filter(Gene == ref_gene)

  # Join on sample and treatment
  joined <- inner_join(target_df, ref_df, by = c("Sample", "Treatment"), suffix = c("_target", "_ref"))

  # Calculate ΔCt
  joined <- joined %>%
    mutate(delta_Ct_target = Ct_target - Ct_ref)

  # Pre-calculate mean ΔCt of control group
  control_delta_mean <- joined %>%
    filter(Treatment == control_group) %>%
    summarise(mean_delta_Ct = mean(delta_Ct_target, na.rm = TRUE)) %>%
    pull(mean_delta_Ct)

  # Calculate ΔΔCt and fold change
  joined <- joined %>%
    mutate(
      delta_delta_Ct = delta_Ct_target - control_delta_mean,
      fold_change = (Efficiency_target ^ -delta_delta_Ct) / (Efficiency_ref ^ -delta_delta_Ct)
    )

  return(joined)
}


# STEP 5: Apply function for each target gene
# STEP 5: Apply function for each target gene (with explicit arguments)
col_results <- map_dfr(genes_to_test, ~pfaffl_fc(col_data, .x, ref_gene = "GAPDH", control_group = "Water"), .id = "GeneIndex") %>%
  mutate(
    log2_fc = log2(fold_change),
    Gene = factor(Gene_target, levels = c("MUC2", "OCLN", "TNF-alpha"))  # Adjust if needed
  )


# STEP 6: Plot fold change

col_results <- col_results %>%
  select(-Gene) %>%                 # remove the existing Gene column
  rename(Gene = Gene_target)       # now safely rename Gene_target to Gene

col_nosig <- ggplot(col_results, aes(x = Treatment, y = fold_change, fill = Treatment)) +
  geom_boxplot(alpha = 0.5) +
  geom_jitter(width = 0.2, size = 2) +
  facet_wrap(~ Gene, scales = "free_y") +
  labs(title = "Colon: Pfaffl-Calculated Fold Change",
       y = "Relative Expression (Fold Change)",
       x = "Treatment") +
  theme_minimal(base_family = "Times New Roman")

col_nosig

```

YAY! We calculated the Pfaffl Fold Change, now time to do some statistics and make this figure look manuscript ready like the other analyses!

### Statistics

```{r stat-assumptions-col}

# Check ANOVA assumptions

# STEP 1: Shapiro-Wilks test for normality (per gene)
shapiro_results_col <- col_results %>%
  group_by(Gene, Treatment) %>%
  summarise(
    shapiro_p = shapiro.test(fold_change)$p.value,
    .groups = "drop"
  )

shapiro_results_col #p > 0.05 normality assumption met


# STEP 2: Levene's Test for equal variances (per gene)
levene_results_col <- col_results %>%
  group_by(Gene) %>%
  group_map(~car::leveneTest(fold_change ~ Treatment, data = .x) %>% 
              tidy() %>%
              mutate(Gene = .y$Gene),
            .keep = TRUE) %>%
  bind_rows()

levene_results_col # p >0.05 for ALL YAY! ANOVA IT IS! 

```

All will be ANOVA!
  
```{r stats-final-col}

# MUC2 and ANOVA
# Filter
muc2_data_col <- filter(col_results, Gene == "MUC2")

# ANOVA
muc2_aov_col <- aov(log2_fc ~ Treatment, data = muc2_data_col)
muc2_p_col <- summary(muc2_aov_col)[[1]][["Pr(>F)"]][1]
muc2_p_col

# Tukey HSD
tukey_muc2_col <- TukeyHSD(muc2_aov_col)
muc2_pairs_col <- as.data.frame(tukey_muc2_col$Treatment)
muc2_pairs_col$comparison <- rownames(muc2_pairs_col)
muc2_sig_col <- muc2_pairs_col %>%
  filter(`p adj` < 0.05) %>%
  separate(comparison, into = c("group1", "group2"), sep = "-") %>%
  mutate(
    label = ifelse(`p adj` < 0.01, "**", "*"),
    y.position = max(muc2_data_col$log2_fc, na.rm = TRUE) + 0.2,
    Gene = "MUC2"
  )
muc2_sig_col

# OCLN ANOVA
ocln_data_col <- filter(col_results, Gene == "OCLN")

# ANOVA
ocln_aov_col <- aov(log2_fc ~ Treatment, data = ocln_data_col)
ocln_p_col <- summary(ocln_aov_col)[[1]][["Pr(>F)"]][1]
ocln_p_col

# Tukey HSD
tukey_ocln_col <- TukeyHSD(ocln_aov_col)
ocln_pairs_col <- as.data.frame(tukey_ocln_col$Treatment)
ocln_pairs_col$comparison <- rownames(ocln_pairs_col)
ocln_sig_col <- ocln_pairs_col %>%
  filter(`p adj` < 0.05) %>%
  separate(comparison, into = c("group1", "group2"), sep = "-") %>%
  mutate(
    label = ifelse(`p adj` < 0.01, "**", "*"),
    y.position = max(ocln_data_col$log2_fc, na.rm = TRUE) + 0.2,
    Gene = "OCLN"
  )
ocln_sig_col

# TNFA ANOVA

tnf_data_col <- filter(col_results, Gene == "TNF-alpha")

# ANOVA
tnf_aov_col <- aov(log2_fc ~ Treatment, data = tnf_data_col)
tnf_p_col <- summary(tnf_aov_col)[[1]][["Pr(>F)"]][1]
tnf_p_col


# Tukey HSD
tukey_tnf_col <- TukeyHSD(tnf_aov_col)
tnf_pairs_col <- as.data.frame(tukey_tnf_col$Treatment)
tnf_pairs_col$comparison <- rownames(tnf_pairs_col)
tnf_sig_col <- tnf_pairs_col %>%
  filter(`p adj` < 0.05) %>%
  separate(comparison, into = c("group1", "group2"), sep = "-") %>%
  mutate(
    label = ifelse(`p adj` < 0.01, "**", "*"),
    y.position = max(tnf_data_col$log2_fc, na.rm = TRUE) + 0.2,
    Gene = "MUC2"
  )

tnf_sig_col

```
  
### Manuscript Ready Colon Figure

#### MUC2

```{r muc-2-fig-col}

# Plot for MUC2

# Ensure treatment factor order
muc2_data_col <- muc2_data_col %>%
  mutate(Treatment = factor(Treatment, levels = c("No Injection", "Water", "Mix Dose 1", "Mix Dose 2", "Mix Dose 3")))

# Define comparison pairs (order matters!)
muc2_comparisons_col <- list(
  c("Mix Dose 3", "Mix Dose 2"),
  c("No Injection", "Mix Dose 3")
)

muc2_sig_col <- muc2_sig_col %>%
  mutate(
    group1 = factor(group1, levels = levels(muc2_data_col$Treatment)),
    group2 = factor(group2, levels = levels(muc2_data_col$Treatment))
  )

max_y_col <- max(muc2_data_col$log2_fc, na.rm = TRUE)
muc2_sig_col <- muc2_sig_col %>%
  mutate(y.position = max_y_col + 0.05 + (row_number() * 0.03))  # stagger brackets

levels(muc2_data_col$Treatment)
levels(muc2_sig_col$group1)

muc2_sig_col <- muc2_sig_col %>%
  mutate(across(c(group1, group2), ~factor(.x, levels = levels(muc2_data_col$Treatment))))

muc2_sig_col <- muc2_sig_col %>%
  mutate(
    group1 = factor(group1, levels = levels(muc2_data_col$Treatment)),
    group2 = factor(group2, levels = levels(muc2_data_col$Treatment)),
    y.position = max(muc2_data_col$log2_fc, na.rm = TRUE) + 0.05 + (row_number() * 0.03)
  )

muc2_sig_col <- muc2_sig_col %>%
  mutate(
    y.position = c(0.075, 0.1)  # lower than before
  )

# Plot using your working style
p_muc2_col <- ggplot(muc2_data_col, aes(x = Treatment, y = log2_fc, fill = Treatment)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA, color = "black") +
  geom_jitter(aes(color = Treatment), width = 0.15, size = 1.8, alpha = 0.8) +
  scale_fill_manual(values = treatment_colors) +
  scale_color_manual(values = treatment_colors) +
  theme_minimal(base_size = 14) +
  theme(
    text = element_text(family = "Times New Roman"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    legend.position = "none"
  ) +
  labs(
    title = "MUC2",
    x = NULL,
    y = expression(log[2]*"(Fold Change)")
  ) +
  ggsignif::geom_signif(
    data = muc2_sig_col,
    aes(
      xmin = group1,
      xmax = group2,
      annotations = label,
      y_position = y.position
    ),
    inherit.aes = FALSE,
    manual = TRUE,
    tip_length = 0.01,
    textsize = 4,
    vjust = 0.3
  ) +
  annotate("text", x = 3, y = 0.125,
           label = "ANOVA~italic(p) == 0.002", parse = TRUE, size = 4.5) +
  coord_cartesian(ylim = c(-0.05, 0.125))

# Show the plot

p_muc2_col

```

#### OCLN

```{r ocln-fig-col}

# Plot for OCLN

# Ensure treatment factor levels
ocln_data_col <- col_results %>%
  filter(Gene == "OCLN") %>%
  mutate(Treatment = factor(Treatment, levels = c("No Injection", "Water", "Mix Dose 1", "Mix Dose 2", "Mix Dose 3")))

# Ensure group1 and group2 are factors with the same levels as Treatment
ocln_sig_col <- ocln_sig_col %>%
  mutate(
    group1 = factor(group1, levels = levels(ocln_data_col$Treatment)),
    group2 = factor(group2, levels = levels(ocln_data_col$Treatment))
  )

ocln_sig_col <- ocln_sig_col %>%
  mutate(
    y.position = c(0.05, 0.06, 0.07, 0.08)  # lower than before
  )


# Plot
p_ocln_col <- ggplot(ocln_data_col, aes(x = Treatment, y = log2_fc, fill = Treatment)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA, color = "black") +
  geom_jitter(aes(color = Treatment), width = 0.15, size = 1.8, alpha = 0.8) +
  scale_fill_manual(values = treatment_colors) +
  scale_color_manual(values = treatment_colors) +
  theme_minimal(base_size = 14) +
  theme(
    text = element_text(family = "Times New Roman"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    legend.position = "none"
  ) +
  labs(
    title = "OCLN",
    x = NULL,
    y = expression(log[2]*"(Fold Change)")
  ) + 
  ggsignif::geom_signif(
    data = ocln_sig_col,
    aes(
      xmin = group1,
      xmax = group2,
      annotations = label,
      y_position = y.position
    ),
    inherit.aes = FALSE,
    manual = TRUE,
    tip_length = 0.01,
    textsize = 4,
    vjust = 0.3
  ) +
  annotate("text", x = 3, y = 0.1,
           label = "ANOVA~italic(p) == 0.0004", parse = TRUE, size = 4.5) +
  coord_cartesian(ylim = c(-0.07, 0.1))

# Show plot
p_ocln_col

```


#### TNFa

```{r tnfa-fig-col}

# Plot for TNFa

# Ensure treatment factor order
tnf_data_col <- tnf_data_col %>%
  mutate(Treatment = factor(Treatment, levels = c("No Injection", "Water", "Mix Dose 1", "Mix Dose 2", "Mix Dose 3")))

# Define comparison pairs (order matters!)
tnf_comparisons_col <- list(
  c("No Injection", "Mix Dose 1"),
  c("Water","Mix Dose 1"),
  c("No Injection", "Mix Dose 2"),
  c("Water","Mix Dose 2"),
  c("No Injection", "Mix Dose 3"),
  c("Water","Mix Dose 3")
)

tnf_sig_col <- tnf_sig_col %>%
  mutate(
    group1 = factor(group1, levels = levels(tnf_data_col$Treatment)),
    group2 = factor(group2, levels = levels(tnf_data_col$Treatment))
  )

max_y_col_tnf <- max(tnf_data_col$log2_fc, na.rm = TRUE)
tnf_sig_col <- tnf_sig_col %>%
  mutate(y.position = max_y_col_tnf + 0.05 + (row_number() * 0.03))  # stagger brackets

levels(tnf_data_col$Treatment)
levels(tnf_sig_col$group1)

tnf_sig_col <- tnf_sig_col %>%
  mutate(across(c(group1, group2), ~factor(.x, levels = levels(tnf_data_col$Treatment))))

tnf_sig_col <- tnf_sig_col %>%
  mutate(
    group1 = factor(group1, levels = levels(tnf_data_col$Treatment)),
    group2 = factor(group2, levels = levels(tnf_data_col$Treatment)),
    y.position = max(tnf_data_col$log2_fc, na.rm = TRUE) + 0.05 + (row_number() * 0.03)
  )

tnf_sig_col <- tnf_sig_col %>%
  mutate(
    y.position = c(0.15, 0.1, 0.25, 0.2, 0.35, 0.3)  # lower than before
  )

# Remove NA groups
tnf_data_col <- tnf_data_col %>% filter(!is.na(Treatment))
tnf_sig_col <- tnf_sig_col %>% filter(!is.na(group1) & !is.na(group2))

# Plot using your working style
p_tnf_col <- ggplot(tnf_data_col, aes(x = Treatment, y = log2_fc, fill = Treatment)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA, color = "black") +
  geom_jitter(aes(color = Treatment), width = 0.15, size = 1.8, alpha = 0.8) +
  scale_fill_manual(values = treatment_colors) +
  scale_color_manual(values = treatment_colors) +
  theme_minimal(base_size = 14) +
  theme(
    text = element_text(family = "Times New Roman"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    legend.position = "none"
  ) +
  labs(
    title = "TNF-α",
    x = NULL,
    y = expression(log[2]*"(Fold Change)")
  ) +
  ggsignif::geom_signif(
    data = tnf_sig_col,
    aes(
      xmin = group1,
      xmax = group2,
      annotations = label,
      y_position = y.position
    ),
    inherit.aes = FALSE,
    manual = TRUE,
    tip_length = 0.01,
    textsize = 4,
    vjust = 0.3
  ) +
  annotate("text", x = 3, y = 0.45,
           label = "ANOVA~italic(p) == 7.6e-06", parse = TRUE, size = 4.5) +
  coord_cartesian(ylim = c(-0.5, 0.45))

# Show the plot

p_tnf_col

```


#### All Final Figures

```{r final-manuscript-fig-col, fig.width=6.5, fig.height=6, dpi=300}

col_p <- (p_muc2_col | p_ocln_col | p_tnf_col) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Colon Gene Expression",
    theme = theme(
      plot.title = element_text(family = "Times New Roman", size = 16, face = "bold", hjust = 0.5)
    )
  )

col_p

ggsave("final_plot_col_all_mix.png", plot = col_p, 
       width = 10, height = 6, units = "in", dpi = 600)
```


## Liver

### Pfaffl Analysis

```{r liv-pfaffl}

# STEP 1: Preprocess and average replicates
mix_avg <- mix_pcr %>%
  filter(Tissue == "Liver") %>%
  rowwise() %>%
  mutate(Ct = mean(c_across(Rep1:Rep2), na.rm = TRUE)) %>%
  ungroup() %>%
  select(Sample, Treatment, Gene, Ct)

# STEP 2: Join in efficiency values
mix_eff <- tibble(
  Gene = c("GAPDH", "PRAAK1", "PRAAK2", "DGAT2" , "PPARg"),
  Efficiency = c(1.93, 2.15, 1.99, 2.1, 2.1)
)

mix_data <- left_join(mix_avg, mix_eff, by = "Gene")

glimpse(mix_data)

# STEP 3: Split reference and target genes
reference_gene <- "GAPDH"
control_group <- "Water"

genes_to_test <- c("PRAAK1", "PRAAK2", "DGAT2", "PPARg")

# STEP 4: Create fold change calculator function using Pfaffl formula
pfaffl_fc <- function(df, target_gene, ref_gene, control_group) {
  # Subset target and reference data
  target_df <- df %>% filter(Gene == target_gene)
  ref_df    <- df %>% filter(Gene == ref_gene)

  # Join on sample and treatment
  joined <- inner_join(target_df, ref_df, by = c("Sample", "Treatment"), suffix = c("_target", "_ref"))

  # Calculate ΔCt
  joined <- joined %>%
    mutate(delta_Ct_target = Ct_target - Ct_ref)

  # Pre-calculate mean ΔCt of control group
  control_delta_mean <- joined %>%
    filter(Treatment == control_group) %>%
    summarise(mean_delta_Ct = mean(delta_Ct_target, na.rm = TRUE)) %>%
    pull(mean_delta_Ct)

  # Calculate ΔΔCt and fold change
  joined <- joined %>%
    mutate(
      delta_delta_Ct = delta_Ct_target - control_delta_mean,
      fold_change = (Efficiency_target ^ -delta_delta_Ct) / (Efficiency_ref ^ -delta_delta_Ct)
    )

  return(joined)
}

# STEP 5: Apply function for each target gene
mix_results <- map_dfr(genes_to_test, ~pfaffl_fc(mix_data, .x, ref_gene = "GAPDH", control_group = "Water"), .id = "GeneIndex") %>%
  mutate(
    log2_fc = log2(fold_change),
    Gene = factor(Gene_target, levels = c("MUC2", "OCLN", "TNF-alpha"))  # Adjust if needed
  )

# STEP 6: Plot fold change
mix_results <- mix_results %>%
  select(-Gene) %>%                 # remove the existing Gene column
  rename(Gene = Gene_target)       # now safely rename Gene_target to Gene

mix_nosig <- ggplot(mix_results, aes(x = Treatment, y = fold_change, fill = Treatment)) +
  geom_boxplot(alpha = 0.5) +
  geom_jitter(width = 0.2, size = 2) +
  facet_wrap(~ Gene, scales = "free_y") +
  labs(title = "Liver: Pfaffl-Calculated Fold Change",
       y = "Relative Expression (Fold Change)",
       x = "Treatment") +
  theme_minimal(base_family = "Times New Roman")

mix_nosig

```

### Statistics

```{r stat-assumptions-liv}

# Check ANOVA assumptions

# STEP 1: Shapiro-Wilks test for normality (per gene)
shapiro_results_mix <- mix_results %>%
  group_by(Gene, Treatment) %>%
  summarise(
    shapiro_p = shapiro.test(fold_change)$p.value,
    .groups = "drop"
  )

shapiro_results_mix # p > 0.05 normality assumption met for none of them

# STEP 2: Levene's Test for equal variances (per gene)
levene_results_mix <- mix_results %>%
  group_by(Gene) %>%
  group_map(~car::leveneTest(fold_change ~ Treatment, data = .x) %>% 
              tidy() %>%
              mutate(Gene = .y$Gene),
            .keep = TRUE) %>%
  bind_rows()

levene_results_mix # fine for praak1 not praak2


```

Kruskal Wallis needed for all of them

```{r stats-final-liv}

# Helper: subset, coerce, drop unused factor levels
prep_gene_data <- function(df, gene_name) {
  df %>%
    filter(Gene == gene_name) %>%
    mutate(Treatment = as.factor(Treatment),
           Gene = as.factor(Gene)) %>%
    droplevels()
}

# ── PRAAK1 ────────────────────────────────────────────────────────────────────
praak1_data_liv  <- prep_gene_data(mix_results, "PRAAK1")
kw_praak1_liv    <- kruskal.test(log2_fc ~ Treatment, data = praak1_data_liv)
praak1_p_liv     <- kw_praak1_liv$p.value
dunn_praak1_liv  <- dunnTest(log2_fc ~ Treatment, data = praak1_data_liv, method = "holm")
praak1_pairs_liv <- dunn_praak1_liv$res
praak1_sig_liv   <- praak1_pairs_liv %>%
  filter(P.adj < 0.05) %>%
  separate(Comparison, into = c("group1", "group2"), sep = " - ") %>%
  mutate(label      = ifelse(P.adj < 0.01, "**", "*"),
         y.position = max(praak1_data_liv$log2_fc, na.rm = TRUE) + 0.2,
         Gene       = "PRAAK1")

kw_praak1_liv
praak1_sig_liv

# ── PRAAK2 ────────────────────────────────────────────────────────────────────
praak2_data_liv  <- prep_gene_data(mix_results, "PRAAK2")
kw_praak2_liv    <- kruskal.test(log2_fc ~ Treatment, data = praak2_data_liv)
praak2_p_liv     <- kw_praak2_liv$p.value
dunn_praak2_liv  <- dunnTest(log2_fc ~ Treatment, data = praak2_data_liv, method = "holm")
praak2_pairs_liv <- dunn_praak2_liv$res
praak2_sig_liv   <- praak2_pairs_liv %>%
  filter(P.adj < 0.05) %>%
  separate(Comparison, into = c("group1", "group2"), sep = " - ") %>%
  mutate(label      = ifelse(P.adj < 0.01, "**", "*"),
         y.position = max(praak2_data_liv$log2_fc, na.rm = TRUE) + 0.2,
         Gene       = "PRAAK2")

kw_praak2_liv
praak2_sig_liv

# ── DGAT2 ─────────────────────────────────────────────────────────────────────
dgat2_data_liv  <- prep_gene_data(mix_results, "DGAT2")
kw_dgat2_liv    <- kruskal.test(log2_fc ~ Treatment, data = dgat2_data_liv)
dgat2_p_liv     <- kw_dgat2_liv$p.value
dunn_dgat2_liv  <- dunnTest(log2_fc ~ Treatment, data = dgat2_data_liv, method = "holm")
dgat2_pairs_liv <- dunn_dgat2_liv$res
dgat2_sig_liv   <- dgat2_pairs_liv %>%
  filter(P.adj < 0.05) %>%
  separate(Comparison, into = c("group1", "group2"), sep = " - ") %>%
  mutate(label      = ifelse(P.adj < 0.01, "**", "*"),
         y.position = max(dgat2_data_liv$log2_fc, na.rm = TRUE) + 0.2,
         Gene       = "DGAT2")

kw_dgat2_liv
dgat2_sig_liv

# ── PPARg ─────────────────────────────────────────────────────────────────────
pparg_data_liv  <- prep_gene_data(mix_results, "PPARg")
kw_pparg_liv    <- kruskal.test(log2_fc ~ Treatment, data = pparg_data_liv)
pparg_p_liv     <- kw_pparg_liv$p.value
dunn_pparg_liv  <- dunnTest(log2_fc ~ Treatment, data = pparg_data_liv, method = "holm")
pparg_pairs_liv <- dunn_pparg_liv$res
pparg_sig_liv   <- pparg_pairs_liv %>%
  filter(P.adj < 0.05) %>%
  separate(Comparison, into = c("group1", "group2"), sep = " - ") %>%
  mutate(label      = ifelse(P.adj < 0.01, "**", "*"),
         y.position = max(pparg_data_liv$log2_fc, na.rm = TRUE) + 0.2,
         Gene       = "PPARg")

kw_pparg_liv
pparg_sig_liv

```

### Manuscript Ready Duodenum Figure

#### PRAAK1

```{r liv-praak1-fig}

# Ensure treatment factor levels
praak1_data_liv <- mix_results %>%
  filter(Gene == "PRAAK1") %>%
  mutate(Treatment = factor(Treatment, levels = c("No Injection", "Water", "Mix Dose 1", "Mix Dose 2", "Mix Dose 3")))

# Ensure group1 and group2 are factors with the same levels as Treatment
praak1_sig_liv <- praak1_sig_liv %>%
  mutate(
    group1 = factor(group1, levels = levels(praak1_data_liv$Treatment)),
    group2 = factor(group2, levels = levels(praak1_data_liv$Treatment))
  )

# Plot
p_praak1_liv <- ggplot(praak1_data_liv, aes(x = Treatment, y = log2_fc, fill = Treatment)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA, color = "black") +
  geom_jitter(aes(color = Treatment), width = 0.15, size = 1.8, alpha = 0.8) +
  scale_fill_manual(values = treatment_colors) +
  scale_color_manual(values = treatment_colors) +
  theme_minimal(base_size = 14) +
  theme(
    text = element_text(family = "Times New Roman"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    legend.position = "none"
  ) +
  labs(
    title = "PRKAA1",
    x = NULL,
    y = expression(log[2]*"(Fold Change)")
  ) + 
  ggsignif::geom_signif(
    data = praak1_sig_liv,
    aes(
      xmin = group1,
      xmax = group2,
      annotations = label,
      y_position = 0.55
    ),
    inherit.aes = FALSE,
    manual = TRUE,
    tip_length = 0.01,
    textsize = 4,
    vjust = 0.3
  ) +
  annotate("text", x = 3, y = 0.6,
           label = "KW~italic(p) == 0.002", parse = TRUE, size = 4.5) +
  coord_cartesian(ylim = c(-0.3, 0.6))

# Show plot
p_praak1_liv

```

#### PRAAK2

```{r praak2-liv-fig}
# Ensure treatment factor levels
praak2_data_liv <- mix_results %>%
  filter(Gene == "PRAAK2") %>%
  mutate(Treatment = factor(Treatment, levels = c("No Injection", "Water", "Mix Dose 1", "Mix Dose 2", "Mix Dose 3")))

# Ensure group1 and group2 are factors with the same levels as Treatment
praak2_sig_liv <- praak2_sig_liv %>%
  mutate(
    group1 = factor(group1, levels = levels(praak2_data_liv$Treatment)),
    group2 = factor(group2, levels = levels(praak2_data_liv$Treatment))
  )


# Plot
p_praak2_liv <- ggplot(praak2_data_liv, aes(x = Treatment, y = log2_fc, fill = Treatment)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA, color = "black") +
  geom_jitter(aes(color = Treatment), width = 0.15, size = 1.8, alpha = 0.8) +
  scale_fill_manual(values = treatment_colors) +
  scale_color_manual(values = treatment_colors) +
  theme_minimal(base_size = 14) +
  theme(
    text = element_text(family = "Times New Roman"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    legend.position = "none"
  ) +
  labs(
    title = "PRKAA2",
    x = NULL,
    y = expression(log[2]*"(Fold Change)")
  ) + 
  ggsignif::geom_signif(
    data = praak2_sig_liv,
    aes(
      xmin = group1,
      xmax = group2,
      annotations = label,
      y_position = 0.1
    ),
    inherit.aes = FALSE,
    manual = TRUE,
    tip_length = 0.01,
    textsize = 4,
    vjust = 0.3
  ) +
  annotate("text", x = 3, y = 0.15,
           label = "KW~italic(p) == 0.023", parse = TRUE, size = 4.5) +
  coord_cartesian(ylim = c(-0.1, 0.15))

# Show plot
p_praak2_liv

```

#### DGAT2

```{r dgat2-liv-fig}
# Ensure treatment factor levels
dgat2_data_liv <- mix_results %>%
  filter(Gene == "DGAT2") %>%
  mutate(Treatment = factor(Treatment, levels = c("No Injection", "Water", "Mix Dose 1", "Mix Dose 2", "Mix Dose 3")))

# Ensure group1 and group2 are factors with the same levels as Treatment
dgat2_sig_liv <- dgat2_sig_liv %>%
  mutate(
    group1 = factor(group1, levels = levels(dgat2_data_liv$Treatment)),
    group2 = factor(group2, levels = levels(dgat2_data_liv$Treatment))
  )

# Plot
p_dgat2_liv <- ggplot(dgat2_data_liv, aes(x = Treatment, y = log2_fc, fill = Treatment)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA, color = "black") +
  geom_jitter(aes(color = Treatment), width = 0.15, size = 1.8, alpha = 0.8) +
  scale_fill_manual(values = treatment_colors) +
  scale_color_manual(values = treatment_colors) +
  theme_minimal(base_size = 14) +
  theme(
    text = element_text(family = "Times New Roman"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    legend.position = "none"
  ) +
  labs(
    title = "DGAT2",
    x = NULL,
    y = expression(log[2]*"(Fold Change)")
  ) + 
  ggsignif::geom_signif(
    data = dgat2_sig_liv,
    aes(
      xmin = group1,
      xmax = group2,
      annotations = label,
      y_position = y.position
    ),
    inherit.aes = FALSE,
    manual = TRUE,
    tip_length = 0.01,
    textsize = 4,
    vjust = 0.3
  ) +
  annotate("text", x = 3, y = 0.3,
           label = "KW~italic(p) == 0.22", parse = TRUE, size = 4.5) +
  coord_cartesian(ylim = c(-0.4, 0.3))

# Show plot
p_dgat2_liv

```

#### PPARg

```{r pprag-liv-fig}

# Ensure treatment factor levels
pparg_data_liv <- mix_results %>%
  filter(Gene == "PPARg") %>%
  mutate(Treatment = factor(Treatment, levels = c("No Injection", "Water", "Mix Dose 1", "Mix Dose 2", "Mix Dose 3")))

# Ensure group1 and group2 are factors with the same levels as Treatment
pparg_sig_liv <- pparg_sig_liv %>%
  mutate(
    group1 = factor(group1, levels = levels(pparg_data_liv$Treatment)),
    group2 = factor(group2, levels = levels(pparg_data_liv$Treatment))
  )

pparg_sig_liv <- pparg_sig_liv %>%
  mutate(
    y.position = c(1.2, 1.1, 1.3, 1.5)  # lower than before
  )

# Plot
p_pparg_liv <- ggplot(pparg_data_liv, aes(x = Treatment, y = log2_fc, fill = Treatment)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA, color = "black") +
  geom_jitter(aes(color = Treatment), width = 0.15, size = 1.8, alpha = 0.8) +
  scale_fill_manual(values = treatment_colors) +
  scale_color_manual(values = treatment_colors) +
  theme_minimal(base_size = 14) +
  theme(
    text = element_text(family = "Times New Roman"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    legend.position = "none"
  ) +
  labs(
    title = expression(bold(PPAR[gamma])),
    x = NULL,
    y = expression(log[2]*"(Fold Change)")
  ) + 
  ggsignif::geom_signif(
    data = pparg_sig_liv,
    aes(
      xmin = group1,
      xmax = group2,
      annotations = label,
      y_position = y.position
    ),
    inherit.aes = FALSE,
    manual = TRUE,
    tip_length = 0.01,
    textsize = 4,
    vjust = 0.3
  ) +
  annotate("text", x = 3, y = 1.8,
           label = "KW~italic(p) == 9.9e-05", parse = TRUE, size = 4.5) +
  coord_cartesian(ylim = c(-0.2, 1.8))

# Show plot
p_pparg_liv

```

#### All Final Figures Manuscript Ready

```{r all-liv-figs, fig.width=10, fig.height=10, dpi=300}

liv_p <- (p_praak1_liv | p_praak2_liv) / (p_dgat2_liv | p_pparg_liv)  +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Liver Gene Expression Response to Mix Treatment",
    theme = theme(
      plot.title = element_text(family = "Times New Roman", size = 16, face = "bold", hjust = 0.5)
    )
  )

liv_p

ggsave("final_plot_liv_all_mix.png", plot = liv_p, 
       width = 10, height = 10, units = "in", dpi = 600)

```


