---
title: "Functinoal Genonimcs of H7601"
author: "Chloe J McGovern"
date: "`r Sys.Date()`"
output:
  html_document:
    css: "styles.css"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                        fig.path = "../Figures/05_FunctionalGenomics/")
```

# Invesigating HY7601 for bacteriocins or antimcirobial properties

Whole genome sequencing of HY7601 was received from hy Co., Ltd. This sequence was a complete sequence and sent as a fasta file. It was processed in Ubuntu with BAGEL4, antiSMASH 7.0, and HMMER to identify areas of interest (AOI) and identify what those may be. 

Below is the R analysis of an integrated data set with all those results. 

## Set up

```{r organize, include=FALSE}
# Load packages
pacman::p_load(tidyverse, janitor, ggplot2, readr, here, knitr)

# Set theme and options
windowsFonts("Times New Roman" = windowsFont("Times New Roman"))

```

## Load Data

```{r load-data}
library(here)

#here()

# Define path (relative to project root)
file_path <- here("BAGEL4", "myrun", "analysis", "antismash", "integrated_results.tsv")

# Read file
integrated <- read_tsv ("//wsl.localhost/Ubuntu/home/chloe/BAGEL4/myrun/integrated_results.tsv", col_names = FALSE)

# confirm
glimpse(integrated)

```

## Clean Data Set

```{r data-cleaning}
# check from original file

dim(integrated)    
head(integrated)

# inspect df

colnames(integrated)
head(integrated, 5)

colnames(integrated) <- c(
  "AOI_ID",       # from BAGEL / AOI identifier
  "Organism",     # N/A or organism name from BLAST
  "Annotation",   # functional annotation from BLAST
  "Function",     # enzyme or protein name
  "Pfam_Info"     # domain info from HMMER output
)

#check column names
head(integrated)
glimpse(integrated)

# make sure everything looks ok 

integrated %>% slice(1:10)

```

## Split Pfam Domain Name and ID

```{r split-pfam}

integrated <- integrated %>%
  separate(Pfam_Info, into = c("Pfam_Domain", "Pfam_ID"), sep = " ", extra = "merge", fill = "right")

head(integrated)

```

## Summarize Domain Frequencies

```{r summarize-domain-pfam}
pfam_summary <- integrated %>%
  count(Pfam_Domain, sort = TRUE)

head(pfam_summary, 10)

```

### Visualize Top 15 Domains

```{r pfam-top-domain}
pfam_summary %>%
  slice_max(n, n = 15) %>%
  ggplot(aes(x = reorder(Pfam_Domain, n), y = n)) +
  geom_col(fill = "#4575b4") +
  coord_flip() +
  labs(
    title = "Most Abundant Pfam Domains in BAGEL4 AOIs",
    x = "Pfam Domain",
    y = "Count"
  ) +
  theme_minimal(base_family = "Times New Roman") +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5)
  )

```

## Co-occurance bewtween domains and functions

```{r domain-fxn-co-occurance}
library(reshape2)

cooccur <- integrated %>%
  count(Function, Pfam_Domain) %>%
  pivot_wider(names_from = Pfam_Domain, values_from = n, values_fill = 0)

# heatmap view
cooccur_matrix <- as.matrix(cooccur[,-1])
rownames(cooccur_matrix) <- cooccur$Function

heatmap(
  cooccur_matrix,
  scale = "row",
  col = colorRampPalette(c("white", "#4575b4"))(100),
  margins = c(10, 10),
  main = "Pfam–Function Co-occurrence Heatmap"
)

```

# Diveristy metrics per AOI

```{r aoi-diversity}

#install.packages("vegan")
library(vegan)

# calculate Pfam diversity per AOI
domain_diversity <- integrated %>%
  count(AOI_ID, Pfam_Domain) %>% 
  pivot_wider(names_from = Pfam_Domain, values_from = n, values_fill = 0) %>%
  column_to_rownames("AOI_ID")

# compute diversity metrics
domain_diversity$Shannon <- diversity(domain_diversity, index = "shannon")
domain_diversity$Simpson <- diversity(domain_diversity, index = "simpson")
domain_diversity$Richness <- specnumber(domain_diversity)

# tidy for plotting
div_summary <- domain_diversity %>%
  select(Shannon, Simpson, Richness) %>%
  pivot_longer(cols = everything(), names_to = "Metric", values_to = "Value", values_drop_na = TRUE)

# plot
ggplot(div_summary, aes(x = Metric, y = Value, fill = Metric)) +
  geom_boxplot(alpha = 0.7, width = 0.6) +
  labs(
    title = "Functional Domain Diversity Across AOIs",
    x = "",
    y = "Diversity Index"
  ) +
  theme_minimal(base_family = "Times New Roman") +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.position = "none"
  )

```

## Link domain diversity to functional annotation

```{r diversity-function}
# summarize mean diversity metrics per annotated function
div_with_function <- integrated %>%
  left_join(
    domain_diversity %>%
      select(Shannon, Simpson, Richness) %>%
      rownames_to_column("AOI_ID"),
    by = "AOI_ID"
  ) %>%
  group_by(Function) %>%
  summarise(
    Mean_Shannon = mean(Shannon, na.rm = TRUE),
    Mean_Simpson = mean(Simpson, na.rm = TRUE),
    Mean_Richness = mean(Richness, na.rm = TRUE),
    AOI_Count = n()
  ) %>%
  arrange(desc(Mean_Shannon))

head(div_with_function)

```

### Visualize

```{r visualize-div-fxn}
div_with_function %>%
  slice_max(AOI_Count, n = 20) %>%
  pivot_longer(cols = starts_with("Mean_"), names_to = "Metric", values_to = "Value") %>%
  ggplot(aes(x = reorder(Function, Value), y = Value, fill = Metric)) +
  geom_col(position = "dodge", alpha = 0.8) +
  coord_flip() +
  labs(
    title = "Average Domain Diversity by Functional Annotation",
    x = "Function",
    y = "Mean Diversity Index"
  ) +
  theme_minimal(base_family = "Times New Roman") +
  scale_fill_manual(values = c("#4575b4", "#74add1", "#abd9e9")) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.title = element_blank()
  )

```

## Genome AOI Map

```{r aoi-map}

# Parse AOI coordinates from IDs (example assumes AOI_ID has contig and region info)
aoi_map <- integrated %>%
  separate(AOI_ID, into = c("Contig", "AOI"), sep = "_HY", remove = FALSE) %>%
  mutate(AOI_index = as.numeric(factor(AOI_ID))) %>%
  distinct(AOI_ID, Contig, AOI_index, Annotation)

glimpse(aoi_map)

aoi_map <- aoi_map %>%
  mutate(Function_Group = case_when(
    str_detect(Annotation, regex("oxidoreductase|NAD|dehydrogenase", ignore_case = TRUE)) ~ "Oxidoreductase",
    str_detect(Annotation, regex("transferase|kinase", ignore_case = TRUE)) ~ "Transferase",
    str_detect(Annotation, regex("hydrolase|peptidase|lipase", ignore_case = TRUE)) ~ "Hydrolase/Peptidase",
    str_detect(Annotation, regex("transport|permease|export|import", ignore_case = TRUE)) ~ "Transport",
    TRUE ~ "Other/Unknown"
  ))

ggplot(aoi_map, aes(x = Contig, y = AOI_index, color = Function_Group)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal(base_size = 12) +
  labs(
    title = "Genome-wide Distribution of AOIs by Function",
    x = "Contig ID",
    y = "AOI Index",
    color = "Functional Class"
  ) +
  theme(
    axis.text.x = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

ggsave(
  filename = "AOI_map_highres.png",
  plot = last_plot(),
  path = "figures",      
  width = 15,            
  height = 15,          
  dpi = 300,            
  bg = "white"           
)

ggplot(aoi_map, aes(x = Contig, y = AOI_index)) +
  geom_point(color = "#1f78b4", size = 2) +
  facet_wrap(~ Function_Group, scales = "free_y", ncol = 3) +
  theme_minimal(base_size = 10) +
  labs(title = "AOI Distribution by Function Group",
       x = "Contig ID", y = "AOI Index")

# fixing an error 
colnames(integrated)


aoi_map <- integrated %>%
  group_by(AOI_ID) %>%
  summarise(Domain_Count = n_distinct(Pfam_Domain)) %>%
  left_join(aoi_map, by = "AOI_ID")

ggplot(aoi_map, aes(x = Contig, y = AOI_index, color = Domain_Count)) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_viridis_c(option = "plasma") +
  theme_minimal() +
  labs(
    title = "Genome-wide AOI Distribution by Domain Richness",
    x = "Contig ID",
    y = "AOI Index",
    color = "Unique Pfam Domains"
  )


```

## Antimicrobial Focus

```{r focus-setup}

#install.packages("ggthemes")
#install.packages("ggtext")

library(ggthemes)
library(ggtext)

theme_set(theme_classic(base_family = "Times New Roman"))
```

### Antimicrobial hits

```{r antimicrobial-filter}
colnames(integrated)

# remove the empty NA column
integrated <- integrated %>%
  select(-which(is.na(colnames(.))))

# double-check
colnames(integrated)

antimicrobial_hits <- integrated %>%
  filter(
    str_detect(Pfam_Domain, regex("bacteriocin|AMP|antimicrobial|lysozyme|defensin|oxidase|peptidase|ABC|thioredoxin|NRPS|PKS|terpene", ignore_case = TRUE)) |
    str_detect(Annotation, regex("bacteriocin|thioredoxin|oxidoreductase|peptidase|AMP|ribosomally synthesized", ignore_case = TRUE))
  ) %>%
  distinct()

# quick overview
cat("Number of putative antimicrobial features detected:", nrow(antimicrobial_hits), "\n")
```

### Antimicrobial landscape

```{r antimicrobial-domains-plot}

library(dplyr)
library(ggplot2)

# 1) count + rank explicitly (highest first), take top 20
antimicrobial_summary <- antimicrobial_hits %>%
  count(Pfam_Domain, sort = TRUE) %>%
  arrange(desc(n)) %>%                       # explicit order
  mutate(
    rank = row_number(),                     # 1,2,3,...
    alt_color = if_else(rank %% 2 == 1, "a", "b")
  ) %>%
  slice_head(n = 20) %>%
  # 2) freeze plotting order so the top bar appears at the top after coord_flip()
  mutate(Pfam_Domain = factor(Pfam_Domain, levels = rev(Pfam_Domain)))

# 3) plot with alternating fill based on rank
antimicro_pfam_frq <- ggplot(antimicrobial_summary,
                             aes(x = Pfam_Domain, y = n, fill = alt_color)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = paste0(round(n / sum(n) * 100, 1), "%")),
            hjust = -0.2, size = 4, family = "Times New Roman") +
  scale_fill_manual(values = c("a" = "hotpink1", "b" = "deeppink3")) +
  coord_flip() +
  labs(
    title = "Most Frequent Antimicrobial-Associated Pfam Domains",
    x = "Pfam Domain",
    y = "Count"
  ) +
  theme_classic(base_family = "Times New Roman") +
  theme(
    text = element_text(size = 14),
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5,
                              margin = margin(b = 10)),
    axis.title.x = element_text(size = 14, margin = margin(t = 10)),
    axis.title.y = element_text(size = 14, margin = margin(r = 10)),
    axis.text = element_text(size = 12, color = "black"),
    panel.grid.major.x = element_line(color = "gray90", linetype = "dashed"),
    plot.margin = margin(15, 25, 15, 25)
  ) +
  ylim(0, max(antimicrobial_summary$n) * 1.15)

antimicro_pfam_frq

ggsave(
  filename = "most_freq_antimicro_pfams.png",
  plot = last_plot(),
  path = "figures",      
  width = 12,            
  height = 8,          
  dpi = 300,            
  bg = "white"           
)

```

YAY!

### Antimicrobial function

```{r antimicrobial-function-summary}

function_summary <- antimicrobial_hits %>%
  count(Function, sort = TRUE) %>%
  filter(!is.na(Function) & Function != "") %>%  # remove NAs or blanks
  slice_head(n = 20) %>%
  arrange(desc(n)) %>%
  mutate(
    rank = row_number(),
    alt_color = if_else(rank %% 2 == 1, "a", "b"),
    Function = factor(Function, levels = rev(Function))
  )

antimicrobial_hits_clean <- function_summary %>%
  filter(
    !grepl("AltName|chloroplastic|protein", Function, ignore.case = TRUE)
  )


function_plot <- ggplot(antimicrobial_hits_clean, aes(x = Function, y = n, fill = alt_color)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(
    aes(label = n),
    hjust = -0.2,
    size = 4,
    family = "Times New Roman"
  ) +
  scale_fill_manual(values = c("a" = "mediumorchid1", "b" = "mediumorchid4")) +
  coord_flip() +
  labs(
    title = "Top Functional Annotations Among Antimicrobial Hits",
    x = "Predicted Function",
    y = "Count"
  ) +
  theme_classic(base_family = "Times New Roman") +
  theme(
    text = element_text(size = 14),
    plot.title = element_text(
      face = "bold",
      size = 18,
      hjust = 0.5,
      margin = margin(b = 10)
    ),
    axis.title.x = element_text(size = 14, margin = margin(t = 10)),
    axis.title.y = element_text(size = 14, margin = margin(r = 10)),
    axis.text = element_text(size = 12, color = "black"),
    panel.grid.major.x = element_line(color = "gray90", linetype = "dashed"),
    plot.margin = margin(15, 25, 15, 25)
  ) +
  ylim(0, max(antimicrobial_hits_clean$n) * 1.15)

function_plot

ggsave(
  filename = "function_plot.png",
  plot = last_plot(),
  path = "figures",      
  width = 12,            
  height = 8,          
  dpi = 300,            
  bg = "white"           
)

# now by function color coding 

# create category mapping
antimicrobial_hits_categorized <- antimicrobial_hits %>%
  mutate(Category = case_when(
    grepl("Lmo|Lin|Mb|SSP|subunit|A$|B$|1A|1B", Function, ignore.case = TRUE) ~ "Biosynthesis/Structural",
    grepl("system", Function, ignore.case = TRUE) ~ "Transport/Immunity System",
    grepl("p60|CwlO", Function, ignore.case = TRUE) ~ "Cell-Wall Hydrolase/Lytic",
    grepl("Lactobacillus|Staphylococcus", Function, ignore.case = TRUE) ~ "Species-Linked Homology",
    TRUE ~ "Other"
  ))

# summarize for top hits
function_summary <- antimicrobial_hits_categorized %>%
  count(Function, Category, sort = TRUE) %>%
  slice_head(n = 20)

antimicrobial_hits_clean <- function_summary %>%
  filter(
    !grepl("AltName|chloroplastic|protein", Function, ignore.case = TRUE)
  )

# color palette by biological category
category_colors <- c(
  "Biosynthesis/Structural" = "pink",
  "Transport/Immunity System" = "powderblue",
  "Cell-Wall Hydrolase/Lytic" = "palegoldenrod",
  "Species-Linked Homology" = "darkseagreen2",
  "Other" = "gray70"
)

# plot

cluster_function <- ggplot(antimicrobial_hits_clean, aes(x = reorder(Function, n), y = n, fill = Category)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = category_colors) +
  labs(
    title = "Functional Categories of Antimicrobial-Associated Genes",
    x = "Predicted Function",
    y = "Count",
    fill = "Functional Category"
  ) +
  theme_classic(base_family = "Times New Roman") +
  theme(
    text = element_text(size = 14),
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    axis.text = element_text(size = 12, color = "black"),
    legend.position = "top"
  )

cluster_function

ggsave(
  filename = "cluster_function.png",
  plot = last_plot(),
  path = "figures",      
  width = 12,            
  height = 8,          
  dpi = 300,            
  bg = "white"           
)

```

### Anitmicrobial map

```{r antimicrobial-map}
antimicrobial_hits %>%
  mutate(AOI_Index = as.numeric(factor(AOI_ID))) %>%
  ggplot(aes(x = AOI_Index, y = Pfam_Domain)) +
  geom_point(color = "firebrick3", alpha = 0.7, size = 2) +
  theme_minimal(base_family = "Times New Roman") +
  labs(
    title = "Genome-Wide Distribution of Antimicrobial-Associated Pfam Domains",
    x = "AOI Index",
    y = "Pfam Domain"
  )

# reorder Pfam by frequency for clearer vertical grouping
pfam_order <- antimicrobial_hits %>%
  count(Pfam_Domain, sort = TRUE) %>%
  pull(Pfam_Domain)

# join your categorized dataframe for color mapping
antimicrobial_map <- antimicrobial_hits_categorized %>%
  mutate(
    AOI_Index = as.numeric(factor(AOI_ID)),
    Pfam_Info = factor(Pfam_Domain, levels = rev(pfam_order))
  )

# clean, color-coded plot
library(viridis)

antimicro_distr <- ggplot(antimicrobial_map, aes(x = AOI_Index, y = Category)) +
  geom_bin2d(bins = 30) +
  scale_fill_viridis(option = "magma", name = "Domain Count") +
  labs(
    title = "Density of Antimicrobial-Associated Pfam Domains Across Genomic Regions",
    x = "Genomic Region (AOI Index)",
    y = "Functional Category"
  ) +
  theme_minimal(base_family = "Times New Roman") +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    legend.position = "right"
  )

antimicro_distr

ggsave(
  filename = "antimicro_distr.png",
  plot = last_plot(),
  path = "figures",      
  width = 10,            
  height = 15,          
  dpi = 300,            
  bg = "white"           
)

# top pfam
top_pfam <- antimicrobial_map %>%
  count(Pfam_Info, sort = TRUE) %>%
  slice_head(n = 25) %>%
  pull(Pfam_Info)

ggplot(antimicrobial_map %>% filter(Pfam_Info %in% top_pfam),
       aes(x = AOI_Index, y = Pfam_Info, color = Category)) +
  geom_jitter(width = 0.3, height = 0.15, alpha = 0.9, size = 2.5) +
  scale_color_manual(values = c(
    "Biosynthesis/Structural" = "#d63384",
    "Transport/Immunity System" = "#0077b6",
    "Cell-Wall Hydrolase/Lytic" = "#ffa600",
    "Species-Linked Homology" = "#009e73",
    "Other" = "gray70"
  )) +
  labs(
    title = "Genome-Wide Distribution of Antimicrobial-Associated Pfam Domains",
    subtitle = "Top 25 Pfam domains across antimicrobial-associated gene clusters",
    x = "Genomic Region (AOI Index)",
    y = "Pfam Domain",
    color = "Functional Category"
  ) +
  theme_minimal(base_family = "Times New Roman") +
  theme(
    text = element_text(size = 13, color = "black"),
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray30"),
    axis.text.y = element_text(size = 10),
    panel.grid.major.x = element_line(color = "gray90", linetype = "dotted"),
    panel.grid.major.y = element_blank(),
    legend.position = "top"
  )
```


## Bacteriocin Class Identification

```{r class-id}

# Classify bacteriocin AOIs based on known domain/keyword patterns
bacteriocin_classes <- antimicrobial_hits %>%
  mutate(
    Class = case_when(
      str_detect(Pfam_Domain, regex("LanC|LanB|LanM|Lantibiotic|dehydratase|RiPP", ignore_case = TRUE)) ~ "Class I (RiPP, modified)",
      str_detect(Pfam_Domain, regex("double-glycine|ABC_transporter|Pediocin|Plantarcin|Enterocin|Immunity", ignore_case = TRUE)) ~ "Class II (unmodified, small peptides)",
      str_detect(Pfam_Domain, regex("cell_wall|hydrolase|lytic|muramidase|endopeptidase", ignore_case = TRUE)) ~ "Class III (large, lytic)",
      TRUE ~ "Unclassified"
    )
  )

# Summarize counts
class_summary <- bacteriocin_classes %>%
  count(Class, sort = TRUE) %>%
  mutate(pct = round(100 * n / sum(n), 1))

class_summary

# Plot
genome_class <- ggplot(class_summary, aes(x = reorder(Class, -n), y = n, fill = Class)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(n, " (", pct, "%)")), vjust = -0.5, family = "Times New Roman", size = 4) +
  scale_fill_manual(values = c("#cc0000", "#ffcc00", "#66ccff", "gray70")) +
  labs(
    title = "Predicted Bacteriocin Classes in L. curvatus HY7601 Genome",
    x = "Bacteriocin Class",
    y = "Number of Hits"
  ) +
  theme_minimal(base_family = "Times New Roman") +
  theme(
    text = element_text(size = 13),
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    axis.text.x = element_text(angle = 25, hjust = 1)
  )

genome_class

ggsave(
  filename = "genome_class.png",
  plot = last_plot(),
  path = "figures",      
  width = 8,            
  height = 8,          
  dpi = 300,            
  bg = "white"           
)

bacteriocin_classes <- antimicrobial_hits %>%
  mutate(
    Class = case_when(
      str_detect(Pfam_Domain, regex("LanC|LanB|LanM|LanT|dehydratase|Lantibiotic", ignore_case = TRUE)) ~ "Class I (RiPP, modified)",
      str_detect(Pfam_Domain, regex("double-glycine|GG-motif|ABC_tran|Pediocin|Enterocin|Plantarcin|Immunity", ignore_case = TRUE)) ~ "Class II (unmodified, small peptides)",
      str_detect(Pfam_Domain, regex("cell_wall|hydrolase|lytic|muramidase|endopeptidase|peptidase", ignore_case = TRUE)) ~ "Class III (large, lytic)",
      TRUE ~ "Unclassified"
    )
  )

bacteriocin_classes %>% count(Class)

bacteriocin_summary <- bacteriocin_classes %>%
  count(Class) %>%
  mutate(pct = round(n / sum(n) * 100, 1))

update_genome_class <- ggplot(bacteriocin_summary, aes(x = reorder(Class, -n), y = n, fill = Class)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(n, " (", pct, "%)")), vjust = -0.5, family = "Times New Roman", size = 4) +
  scale_fill_manual(values = c("#cc0000", "#ffcc00", "#66ccff", "gray70")) +
  labs(
    title = "Predicted Bacteriocin Classes in L. curvatus HY7601 Genome",
    x = "Bacteriocin Class",
    y = "Number of Hits"
  ) +
  theme_minimal(base_family = "Times New Roman") +
  theme(
    text = element_text(size = 13),
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    axis.text.x = element_text(angle = 25, hjust = 1)
  )

update_genome_class

ggsave(
  filename = "update_genome_class.png",
  plot = last_plot(),
  path = "figures",      
  width = 8,            
  height = 8,          
  dpi = 300,            
  bg = "white"           
)

# table 

col(bacteriocin_classes)

bacteriocin_table <- bacteriocin_classes %>%
  select(AOI_ID, Organism, Annotation, Function, Pfam_Domain, Class) %>%
  arrange(Class)

# view the first few entries nicely
head(bacteriocin_table, 15)

#View(bacteriocin_table)

```

