---
title: "Dose-Response Analysis: Histology & PCR Endpoints"
author: "Chloe J McGovern"
date: "`r Sys.Date()`"
output:
  html_document:
    css: "styles.css"
    toc: true
    toc_float: true
    toc_depth: 3
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  fig.path  = "../Figures/06_DoseResponse/"
)
```

# Overview

This analysis characterizes the linear dose-response relationship between mix
dose (CFU/mL) and intestinal functionality outcomes using linear mixed models
(LMM) with log10(dose) as a continuous fixed effect.

**Rationale for linear model:** Exploratory analysis of group means across dose
levels revealed no sigmoidal inflection point within the tested dose range
(Water = 0 CFU/mL; Dose 1 = 3.2e5; Dose 2 = 3.15e6; Dose 3 = 4.1e7 CFU/mL).
Responses were either monotonically trending or plateauing, indicating the
tested range does not span the full dynamic range required to fit a
4-parameter log-logistic (Hill) model. A linear dose-response on the log10 dose
scale is therefore the most appropriate and defensible model for these data.

**Dose anchoring:** The Water group (vehicle control, 0 CFU/mL) is included and
assigned a nominal dose of 1 CFU/mL solely to permit log10 transformation.
The No Injection group is excluded from model fitting as it is not a
vehicle-matched control.

**Model framework:**

| Endpoint type | Model | Interpretation |
|---|---|---|
| Histology | LMM: response ~ log10(dose) + (1|bio_rep) | beta = change per 10-fold increase in dose |
| PCR log2 fold change | LMM: log2_fc ~ log10(dose) + (1|Sample) | beta = log2FC change per 10-fold dose |
| Cross-endpoint summary | Slope forest plot with 95% CI | Identifies most dose-sensitive endpoints |

---

# 1. Setup

## 1.1 Packages

```{r packages}

#install.packages("performance")
#nstall.packages("DHARMa")
#install.packages("influence.ME")
#install.packages("ggeffects")
#install.packages("ggsignif")

pacman::p_load(
  tidyverse, lme4, lmerTest, emmeans,
  patchwork, scales, performance, DHARMa, influence.ME, ggeffects, ggsignif, ggtext, install = FALSE
)

# Resolve namespace conflicts (MASS/car may be loaded from other Rmds)
library(conflicted)
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
conflict_prefer("rename", "dplyr")
conflict_prefer("lmer",   "lmerTest")

# Times New Roman font (Windows)
windowsFonts("Times New Roman" = windowsFont("Times New Roman"))

```

## 1.2 Dose Mapping

```{r dose-map}
# Water = 1 CFU/mL (log10 offset anchor; biologically = 0 dose)
# No Injection = NA (excluded from model fitting)
dose_map <- tibble(
  treatment = c("No Injection", "Water",
                "Mix Dose 1", "Mix Dose 2", "Mix Dose 3"),
  dose_cfu  = c(NA, 1, 4.0e5, 2.75e6, 3.2e7)
)

treatment_levels <- c("No Injection", "Water",
                      "Mix Dose 1", "Mix Dose 2", "Mix Dose 3")
```

## 1.3 Shared Theme

## 1.3 Shared Theme

```{r theme}
theme_dr <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      text             = element_text(family = "Times New Roman"),
      axis.title       = element_text(face = "bold"),
      plot.title       = element_text(face = "bold", hjust = 0.5),
      plot.subtitle    = element_text(hjust = 0.5, color = "grey40"),
      panel.grid.minor = element_blank(),
      legend.position  = "none"
    )
}

# Mix-specific treatment color palette — used by scale_color_manual() in 2.6
treatment_colors <- c(
  "Water"      = "#666666",
  "Mix Dose 1" = "coral2",
  "Mix Dose 2" = "cyan3",
  "Mix Dose 3" = "mediumseagreen"
)
```

## 1.4 Helper Functions

```{r helpers}

pfaffl_fc <- function(df, target_gene, ref_gene = "GAPDH",
                      control_group = "Water") {
  target_df <- df %>% dplyr::filter(Gene == target_gene)
  ref_df    <- df %>% dplyr::filter(Gene == ref_gene)

  # control-group mean Ct, computed separately for target and reference
  t_ctrl <- target_df %>%
    dplyr::filter(Treatment == control_group) %>%
    summarise(m = mean(Ct, na.rm = TRUE)) %>%
    pull(m)

  r_ctrl <- ref_df %>%
    dplyr::filter(Treatment == control_group) %>%
    summarise(m = mean(Ct, na.rm = TRUE)) %>%
    pull(m)

  inner_join(target_df, ref_df,
            by     = c("Sample", "Treatment"),
            suffix = c("_target", "_ref")) %>%
    mutate(
      dCt_target  = t_ctrl - Ct_target,   # control - sample (target)
      dCt_ref     = r_ctrl - Ct_ref,      # control - sample (reference)
      fold_change = (Efficiency_target ^ dCt_target) /
                    (Efficiency_ref    ^ dCt_ref),
      log2_fc     = log2(fold_change)
    )
}

# ── LMM: response ~ log10(dose) + (1|random_col) ─────────────────────────────
fit_lmm_dose <- function(df, response_col, random_col, endpoint_label) {
  fit_df <- df %>%
    dplyr::filter(!is.na(dose_cfu)) %>%
    mutate(log10_dose = log10(dose_cfu))
  tryCatch({
    f   <- as.formula(paste0(response_col,
                             " ~ log10_dose + (1|", random_col, ")"))
    fit <- lmerTest::lmer(f, data = fit_df, REML = TRUE)

    if (lme4::isSingular(fit)) {
      warning(sprintf("Singular fit for %s (Mix) — random effect variance ~0",
                      endpoint_label))
    }

    # Extract fixed effects via base R — avoids broom.mixed rlang version conflict
    coef_tbl  <- as.data.frame(coef(summary(fit)))
    ci        <- confint(fit, parm = "log10_dose", method = "Wald")
    tidy_out <- tibble(
      term      = "log10_dose",
      estimate  = coef_tbl["log10_dose", "Estimate"],
      conf.low  = ci["log10_dose", "2.5 %"],
      conf.high = ci["log10_dose", "97.5 %"],
      p.value   = coef_tbl["log10_dose", "Pr(>|t|)"],
      endpoint  = endpoint_label,
      singular  = lme4::isSingular(fit),
      converged = TRUE
    )
    list(model = fit, params = tidy_out,
         converged = TRUE, label = endpoint_label)
  }, error = function(e) {
    warning(sprintf("LMM failed for %s (Mix): %s", endpoint_label, e$message))
    list(converged = FALSE, label = endpoint_label)
  })
}

# ── Population-level prediction line ──────────────────────────────────────────
predict_lmm <- function(model_obj, dose_grid) {
  if (!model_obj$converged) return(NULL)
  nd   <- data.frame(log10_dose = log10(dose_grid))
  pred <- predict(model_obj$model, newdata = nd, re.form = NA)
  tibble(dose_cfu = dose_grid, fit = pred, label = model_obj$label)
}

# Dose grid: Water anchor (1 CFU/mL) through top of Mix Dose 3 (3.2e7 CFU/mL)
dose_grid <- 10^seq(0, log10(5e7), length.out = 400)

# ── PRE-FIT: group size sufficiency per treatment × random-effect level ──────
check_group_sizes <- function(df, group_var = "treatment", random_var) {
  df %>%
    dplyr::filter(!is.na(dose_cfu)) %>%
    dplyr::count(.data[[group_var]], .data[[random_var]]) %>%
    dplyr::count(.data[[group_var]], name = "n_levels") %>%
    dplyr::mutate(flag = if_else(n_levels < 3,
                                 "LOW — check before fitting", "ok"))
}

# ── PRE-FIT: LOESS vs. linear divergence, per endpoint ────────────────────────
check_loess_linear <- function(df, dose_col = "dose_cfu", response_col,
                               endpoint_label) {
  df %>%
    dplyr::filter(!is.na(.data[[dose_col]])) %>%
    ggplot(aes(x = log10(.data[[dose_col]]), y = .data[[response_col]])) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "loess", se = FALSE, color = "firebrick") +
    geom_smooth(method = "lm",    se = FALSE, color = "steelblue",
               linetype = "dashed") +
    labs(title = paste0(endpoint_label, " — LOESS (red) vs linear (blue)"),
        x = "log10(dose CFU/mL)", y = response_col) +
    theme_dr()
}

# ── PRE-FIT: response distribution check (flag count/skewed endpoints) ───────
check_response_distribution <- function(df, response_col, endpoint_label) {
  df %>%
    dplyr::filter(!is.na(dose_cfu)) %>%
    ggplot(aes(x = .data[[response_col]])) +
    geom_histogram(bins = 15) +
    labs(title = paste0(endpoint_label, " — response distribution")) +
    theme_dr()
}

# ── POST-FIT: convergence, singularity, collinearity, ICC ────────────────────
check_lmm_fit <- function(fit) {
  list(
    singular    = lme4::isSingular(fit),
    convergence = fit@optinfo$conv$opt,
    vif         = tryCatch(performance::check_collinearity(fit),
                           error = function(e) NA),
    icc         = performance::icc(fit)
  )
}

fit_lmm_dose <- function(df, response_col, random_col, endpoint_label) {
  fit_df <- df %>%
    dplyr::filter(!is.na(dose_cfu)) %>%
    mutate(log10_dose = log10(dose_cfu))

  # influence.ME needs the fitting data addressable by a *plain name* in
  # .GlobalEnv, because it inspects the model's stored call and re-evaluates
  # it verbatim. Assign a unique global object, then build the lmer call so
  # the stored `data =` argument is the bare symbol, not an expression.
  global_name <- paste0("fit_df_", make.names(endpoint_label))
  assign(global_name, fit_df, envir = .GlobalEnv)

  tryCatch({
    f <- as.formula(paste0(response_col,
                           " ~ log10_dose + (1|", random_col, ")"))

    fit <- do.call(
      lmerTest::lmer,
      list(formula = f, data = as.name(global_name), REML = TRUE)
    )

    if (lme4::isSingular(fit)) {
      warning(sprintf("Singular fit for %s (Mix) — random effect variance ~0",
                      endpoint_label))
    }

    coef_tbl  <- as.data.frame(coef(summary(fit)))
    ci        <- confint(fit, parm = "log10_dose", method = "Wald")
    tidy_out <- tibble(
      term      = "log10_dose",
      estimate  = coef_tbl["log10_dose", "Estimate"],
      conf.low  = ci["log10_dose", "2.5 %"],
      conf.high = ci["log10_dose", "97.5 %"],
      p.value   = coef_tbl["log10_dose", "Pr(>|t|)"],
      endpoint  = endpoint_label,
      singular  = lme4::isSingular(fit),
      converged = TRUE
    )
    list(model = fit, params = tidy_out,
         converged = TRUE, label = endpoint_label, data_name = global_name)
  }, error = function(e) {
    warning(sprintf("LMM failed for %s (Mix): %s", endpoint_label, e$message))
    list(converged = FALSE, label = endpoint_label)
  })
}
```

---

# 2. Histology

## 2.1 Load & Prepare Data

```{r histo-load}
# What raw treatment codes actually exist in the Mix CSV?
read.csv("Data/CSVs/mix-histology-duod.csv") %>%
  count(treatment)

mix_histo <- read.csv("Data/CSVs/mix-histology-duod.csv") %>%
  mutate(
    treatment = factor(treatment,
      levels = c("1", "2", "7", "8", "9"),
      labels = treatment_levels
    )
  ) %>%
  rowwise() %>%
  mutate(
    width_avg = mean(c(villis_width1, villis_width2, villis_width3),
                     na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    radius       = width_avg / 2,
    surface_area = pi * radius^2 + 2 * pi * radius * villis_height,
    vh_cd        = villis_height / crypt_depth
  ) %>%
  left_join(dose_map, by = "treatment")

mix_histo <- mix_histo %>%
  mutate(bio_rep_nested = paste(treatment, biological_rep, sep = "_"))

# ── Join diagnostic — confirm dose_cfu populated for all treatment groups ────
join_check <- mix_histo %>% count(treatment, dose_cfu)
join_check

if (any(is.na(join_check$dose_cfu) & join_check$treatment != "No Injection")) {
  warning("dose_cfu is NA for a treatment group other than 'No Injection' — ",
         "check dose_map$treatment matches mix_histo$treatment exactly")
}
```


## 2.2 Pre and Post-Fit Diagnostics

```{r histo-endpoint-table}
histo_endpoints <- tribble(
  ~col,            ~label,                ~unit,
  "villis_height", "Villus Height",       "um",
  "width_avg",     "Villus Width",        "um",
  "surface_area",  "Villus Surface Area", "um2",
  "crypt_depth",   "Crypt Depth",         "um",
  "vh_cd",         "VH:CD Ratio",         "ratio",
  "goblet_count",  "Goblet Cell Count",   "count"
)
```

```{r histo-prefit-groupsizes}
check_group_sizes(mix_histo, group_var = "treatment", random_var = "bio_rep_nested")
```

```{r histo-prefit-loess, fig.width = 5, fig.height = 4}
# Run per endpoint — flag any where LOESS diverges from the linear fit
loess_checks <- pmap(
  histo_endpoints,
  function(col, label, unit) check_loess_linear(mix_histo, response_col = col,
                                                endpoint_label = label)
)
walk(loess_checks, print)
```

```{r histo-prefit-distribution, fig.width = 4, fig.height = 3}
# Especially important for goblet_count (count data — check for skew/zero-inflation
# before assuming a Gaussian LMM is appropriate)
dist_checks <- pmap(
  histo_endpoints,
  function(col, label, unit) check_response_distribution(mix_histo, response_col = col,
                                                          endpoint_label = label)
)
walk(dist_checks, print)
```


## 2.3 Fit LMMs

```{r histo-fit}
histo_models <- pmap(
  histo_endpoints,
  function(col, label, unit) {
    fit_lmm_dose(mix_histo, col, "bio_rep_nested", label)   # FIXED: was "biol_rep_nested" (typo)
  }
) %>%
  set_names(histo_endpoints$label)

histo_params <- histo_models %>%
  keep(~ .x$converged) %>%
  map_dfr(~ .x$params) %>%
  dplyr::select(endpoint, estimate, conf.low, conf.high, p.value) %>%
  mutate(sig = case_when(
    p.value < 0.001 ~ "***",
    p.value < 0.01  ~ "**",
    p.value < 0.05  ~ "*",
    TRUE            ~ "ns"
  ))

histo_params
```

## 2.4 Post-Fit Diagnostics

```{r histo-postfit-checks}
fit_diagnostics <- histo_models %>%
  keep(~ .x$converged) %>%
  imap_dfr(~ {
    d <- check_lmm_fit(.x$model)
    tibble(endpoint = .y, singular = d$singular, convergence = d$convergence,
          icc = d$icc$ICC_adjusted)
  })

fit_diagnostics
```

```{r histo-postfit-residuals, fig.width = 5, fig.height = 4}
dharma_checks <- histo_models %>%
  keep(~ .x$converged) %>%
  imap(~ {
    sim <- DHARMa::simulateResiduals(.x$model)
    list(label = .y, sim = sim)
  })

walk(dharma_checks, ~ {
  plot(.x$sim, title = .x$label)
})

DHARMa::testQuantiles(dharma_checks[["Goblet Cell Count"]]$sim)

plot(dharma_checks[["Goblet Cell Count"]]$sim, 
    form = mix_histo %>% dplyr::filter(!is.na(dose_cfu)) %>% pull(dose_cfu) %>% log10())
```

```{r histo-postfit-influence, fig.width = 5, fig.height = 4}
influence_checks <- histo_models %>%
  keep(~ .x$converged) %>%
  imap(~ {
    infl <- influence.ME::influence(.x$model, group = "bio_rep_nested")
    plot(infl, which = "cook", main = .y)
  })
```

## 2.5 Check Remaining Endpoints for Flagged Deviations

```{r check-other-params}
dharma_summary <- histo_models %>%
  keep(~ .x$converged) %>%
  imap_dfr(~ {
    sim <- DHARMa::simulateResiduals(.x$model)
    tibble(
      endpoint      = .y,
      ks_p          = DHARMa::testUniformity(sim, plot = FALSE)$p.value,
      dispersion_p  = DHARMa::testDispersion(sim, plot = FALSE)$p.value
    )
  })

dharma_summary

check_loess_linear(mix_histo, response_col = "crypt_depth", endpoint_label = "Crypt Depth")

sim_crypt <- DHARMa::simulateResiduals(histo_models[["Crypt Depth"]]$model)
plot(sim_crypt)
DHARMa::testQuantiles(sim_crypt)
```

## 2.6 Dose-Response Curves — Histology

```{r fig-histo-dose}
flagged_endpoints <- c("Goblet Cell Count", "Crypt Depth")

histo_predictions <- histo_models %>%
  keep(~ .x$converged) %>%
  imap_dfr(~ {
    pred <- ggeffects::ggpredict(.x$model, terms = "log10_dose [all]")
    tibble(
      log10_dose = pred$x,
      dose_cfu   = 10^pred$x,
      fit        = pred$predicted,
      conf.low   = pred$conf.low,
      conf.high  = pred$conf.high,
      label      = .y
    )
  })

histo_obs <- mix_histo %>%                          
  dplyr::filter(!is.na(dose_cfu)) %>%
  dplyr::select(treatment, dose_cfu, all_of(histo_endpoints$col)) %>%
  pivot_longer(all_of(histo_endpoints$col), names_to = "col", values_to = "response") %>%
  left_join(histo_endpoints, by = "col") %>%
  mutate(log10_dose = log10(dose_cfu))

# Dose values per treatment group, on the log10 scale — must match dose_map exactly
dose_lookup <- dose_map %>%
  dplyr::filter(!is.na(dose_cfu)) %>%
  mutate(log10_dose = log10(dose_cfu))

pairwise_sig <- histo_models %>%
  keep(~ .x$converged) %>%
  imap_dfr(~ {
    emm <- emmeans::emmeans(.x$model, ~ log10_dose,
                            at = list(log10_dose = dose_lookup$log10_dose))
    as.data.frame(emmeans::contrast(emm, "pairwise")) %>%
      mutate(label = .y)
  }) %>%
  dplyr::filter(p.value < 0.05)

# emmeans labels contrasts by the numeric log10_dose values (e.g. "0 - 5.602"),
# not treatment names — map back to readable group labels
dose_to_treatment <- dose_lookup %>%
  transmute(dose_str = as.character(round(log10_dose, 3)), treatment)

pairwise_sig <- pairwise_sig %>%
  separate(contrast, into = c("dose1", "dose2"), sep = " - ") %>%
  mutate(
    dose1 = str_trim(dose1),
    dose2 = str_trim(dose2)
  ) %>%
  left_join(dose_to_treatment, by = c("dose1" = "dose_str")) %>%
  dplyr::rename(group1 = treatment) %>%
  left_join(dose_to_treatment, by = c("dose2" = "dose_str")) %>%
  dplyr::rename(group2 = treatment)

pairwise_sig

panel_order <- c("Goblet Cell Count", "Crypt Depth", "VH:CD Ratio",
                 "Villus Height", "Villus Surface Area", "Villus Width")

unit_labels <- c(
  "Goblet Cell Count"   = "Goblet Cell Count (n)<sup>a</sup>",
  "Crypt Depth"         = "Crypt Depth (\u00b5m)<sup>a</sup>",
  "VH:CD Ratio"         = "VH:CD Ratio",
  "Villus Height"       = "Villus Height (\u00b5m)",
  "Villus Surface Area" = "Surface Area (\u00b5m\u00b2)",
  "Villus Width"        = "Villus Width (\u00b5m)"
)

histo_predictions <- histo_predictions %>% mutate(label = factor(label, levels = panel_order))
histo_obs         <- histo_obs         %>% mutate(label = factor(label, levels = panel_order))

p_histo_dr <- ggplot() +
  geom_ribbon(data = histo_predictions,
             aes(x = dose_cfu, ymin = conf.low, ymax = conf.high),
             fill = "#3B6FA0", alpha = 0.15) +
  geom_line(data = histo_predictions,
           aes(x = dose_cfu, y = fit),
           color = "#3B6FA0", linewidth = 1.0) +
  geom_smooth(data = histo_obs %>% dplyr::filter(label %in% flagged_endpoints),
             aes(x = dose_cfu, y = response),
             method = "loess", se = FALSE, color = "grey40",
             linetype = "dashed", linewidth = 0.7) +
  geom_jitter(data = histo_obs,
             aes(x = dose_cfu, y = response, color = treatment),
             width = 0.04, size = 0.9, alpha = 0.45) +
  scale_x_log10(labels = label_log(digits = 1),
               breaks = c(1, 1e5, 1e6, 1e7)) +
  scale_color_manual(values = treatment_colors, name = NULL) +
  facet_wrap(~ label, scales = "free_y", ncol = 3,
             labeller = as_labeller(unit_labels)) +
  labs(x = "Mix Dose (CFU/mL, log scale)", y = NULL,
      caption = "Blue line/band = LMM fit \u00b1 95% CI | Dashed grey = LOESS (flagged endpoints only)") +
  theme_dr(base_size = 9) +
  theme(legend.position = "bottom",
       strip.text      = ggtext::element_markdown(face = "bold", size = 8),
       plot.caption    = element_text(size = 7, color = "grey40"))

p_histo_dr
```


## 2.7 Slope Summary Table

```{r histo-slope-table, fig.width = 6.5, fig.height = 4.5, dpi = 300}
histo_params %>%
  mutate(across(c(estimate, conf.low, conf.high), ~ round(.x, 3)),
         p.value = format.pval(p.value, digits = 3, eps = 0.001)) %>%
  knitr::kable(
    caption   = "LMM slope estimates for histology endpoints. Beta = change per 1-unit increase in log10(dose CFU/mL). Random effect: (1|bio_rep_nested), nested within treatment.",
    col.names = c("Endpoint", "Beta", "95% CI low", "95% CI high",
                  "P-value", "Sig.")
  )

histo_forest <- histo_params %>%
  mutate(
    sig_fill   = p.value < 0.05,
    scale_group = if_else(endpoint == "Villus Surface Area",
                           "Surface area", "Linear / count measures")
  )

p_forest_histo <- ggplot(histo_forest,
                   aes(x = estimate, y = endpoint, shape = sig_fill)) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "grey50", linewidth = 0.6) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.3, linewidth = 0.8, color = "grey20") +
  geom_point(size = 3.2, color = "mediumvioletred") +
  facet_wrap(~ scale_group, scales = "free", ncol = 1) +
  scale_shape_manual(
    values = c("TRUE" = 19, "FALSE" = 1),
    labels = c("TRUE" = "p < 0.05", "FALSE" = "p \u2265 0.05"),
    name   = NULL,
    guide  = guide_legend(override.aes = list(color = "grey40", size = 3))
  ) +
  labs(
    x       = expression(beta~"(change per 1-unit increase in log"[10]*" dose CFU/mL)"),
    y       = NULL,
    caption = "Filled = p < 0.05 | Open = p \u2265 0.05 | Error bars = 95% CI | Random effect: (1|bio_rep_nested), nested within treatment"
  ) +
  theme_dr(base_size = 10) +
  theme(
    legend.position  = "right",
    legend.text      = element_text(size = 8),
    legend.title     = element_text(size = 9, face = "bold"),
    legend.key.size  = unit(0.45, "cm"),
    axis.text.y      = element_text(size = 9),
    axis.text.x      = element_text(size = 8),
    axis.title.x     = element_text(size = 9),
    strip.text       = element_text(size = 9, face = "bold"),
    plot.caption     = element_text(size = 7, color = "grey40",
                                    family = "Times New Roman"),
    plot.margin      = margin(4, 6, 4, 4, "pt")
  )

p_forest_histo

```

# 3. PCR for Duodenum

## 3.1 Load & Compute Pfaffl Fold Change

```{r pcr-load}
mix_pcr <- read.csv("Data/CSVs/pcr-data-mix.csv")

pcr_avg <- mix_pcr %>%
  rowwise() %>%
  mutate(Ct = mean(c_across(Rep1:Rep2), na.rm = TRUE)) %>%
  ungroup() %>%
  dplyr::select(Sample, Treatment, Tissue, Gene, Ct)

eff_duo <- tibble(Gene       = c("GAPDH","OCLN","MUC2","TNF-alpha"),
                  Efficiency = c(2.02, 1.93, 1.90, 1.90))
eff_col <- tibble(Gene       = c("GAPDH","OCLN","MUC2","TNF-alpha"),
                  Efficiency = c(1.93, 1.90, 1.96, 2.10))
eff_liv <- tibble(Gene       = c("GAPDH","PRKAA1","PRKAA2","DGAT2","PPARg"),
                  Efficiency = c(1.93, 2.10, 1.99, 2.1, 2.1))

prep_tissue <- function(tissue_name, eff_tbl, genes_to_test) {
  avg <- pcr_avg %>%
    dplyr::filter(Tissue == tissue_name) %>%
    left_join(eff_tbl, by = "Gene")

  map_dfr(genes_to_test,
          ~ pfaffl_fc(avg, .x, ref_gene = "GAPDH", control_group = "Water"),
          .id = "GeneIndex") %>%
    dplyr::rename(Gene = Gene_target) %>%
    dplyr::select(-any_of("GeneIndex")) %>%
    mutate(
      Tissue    = tissue_name,
      Treatment = factor(Treatment, levels = treatment_levels)
    ) %>%
    left_join(dose_map, by = c("Treatment" = "treatment"))
}

duo_results <- prep_tissue("Duodenum", eff_duo, c("MUC2","OCLN","TNF-alpha"))
col_results <- prep_tissue("Colon",    eff_col, c("MUC2","OCLN","TNF-alpha"))
liv_results <- prep_tissue("Liver",    eff_liv, c("PRKAA1","PRKAA2","DGAT2","PPARg"))

pcr_all <- bind_rows(duo_results, col_results, liv_results) %>%
  mutate(
    endpoint_label = paste0(Gene, " (", Tissue, ")"),
    Tissue         = factor(Tissue, levels = c("Duodenum","Colon","Liver"))
  )
```

## 3.2 Fit LMMs — PCR

```{r pcr-fit}
pcr_endpoint_list <- pcr_all %>%
  distinct(endpoint_label) %>%
  pull(endpoint_label)

pcr_models <- map(pcr_endpoint_list, function(ep) {
  df <- pcr_all %>%
    dplyr::filter(endpoint_label == ep) %>%
    dplyr::rename(response = log2_fc)
  fit_lmm_dose(df, "response", "Sample", ep)
}) %>%
  set_names(pcr_endpoint_list)

pcr_params <- pcr_models %>%
  keep(~ .x$converged) %>%
  map_dfr(~ .x$params) %>%
  dplyr::select(endpoint, estimate, conf.low, conf.high, p.value) %>%
  mutate(sig = case_when(
    p.value < 0.001 ~ "***",
    p.value < 0.01  ~ "**",
    p.value < 0.05  ~ "*",
    TRUE            ~ "ns"
  )) %>%
  separate(endpoint, into = c("Gene","Tissue"), sep = " \\(", extra = "drop") %>%
  mutate(Tissue = str_remove(Tissue, "\\)"),
         Tissue = factor(Tissue, levels = c("Duodenum","Colon","Liver")))

pcr_params
```

## 3.3 Dose-Response Plots — PCR

```{r pcr-plot-function}
plot_pcr_tissue <- function(tissue_name, tissue_results, tissue_models) {

  eps   <- tissue_results %>% distinct(endpoint_label) %>% pull()
  preds <- tissue_models[eps] %>%
    keep(~ .x$converged) %>%
    map_dfr(~ predict_lmm(.x, dose_grid))

  obs_fit <- tissue_results %>%
    dplyr::filter(!is.na(dose_cfu)) %>%
    dplyr::select(dose_cfu, log2_fc, Treatment, endpoint_label)

  obs_noinj <- tissue_results %>%
    dplyr::filter(is.na(dose_cfu)) %>%
    mutate(dose_plot = 0.3) %>%
    dplyr::select(dose_plot, log2_fc, Treatment, endpoint_label)

  ggplot() +
    geom_line(
      data = preds,
      aes(x = dose_cfu, y = fit),
      color = "#3B6FA0", linewidth = 1.1
    ) +
    geom_jitter(
      data  = obs_fit,
      aes(x = dose_cfu, y = log2_fc, color = Treatment),
      width = 0.04, size = 2, alpha = 0.80
    ) +
    geom_jitter(
      data  = obs_noinj,
      aes(x = dose_plot, y = log2_fc, color = Treatment),
      width = 0.02, size = 2, alpha = 0.55, shape = 1
    ) +
    geom_hline(yintercept = 0, linetype = "dotted", color = "grey50") +
    scale_x_log10(
      labels = label_log(digits = 1),
      breaks = c(1, 1e5, 1e6, 1e7),
      name   = "MIX Dose (CFU/mL, log scale)"
    ) +
    scale_color_manual(values = treatment_colors, name = NULL) +
    facet_wrap(~ endpoint_label, scales = "free_y", ncol = 3) +
    labs(
      title    = paste(tissue_name, "Gene Expression - Linear Dose-Response (LMM)"),
      subtitle = "Line = population fit | Dotted = log2FC 0 | Open circles = No Injection",
      y        = "log2(Fold Change)"
    ) +
    theme_dr() +
    theme(legend.position = "bottom",
          strip.text      = element_text(face = "bold"))
}
```

```{r pcr-dr-duo, fig.width = 10, fig.height = 4.5, dpi = 300}
p_pcr_duo <- plot_pcr_tissue(
  "Duodenum",
  pcr_all %>% dplyr::filter(Tissue == "Duodenum"),
  pcr_models
)
p_pcr_duo
```

```{r pcr-dr-col, fig.width = 10, fig.height = 4.5, dpi = 300}
p_pcr_col <- plot_pcr_tissue(
  "Colon",
  pcr_all %>% dplyr::filter(Tissue == "Colon"),
  pcr_models
)
p_pcr_col
```

```{r pcr-dr-liv, fig.width = 8, fig.height = 4.5, dpi = 300}
p_pcr_liv <- plot_pcr_tissue(
  "Liver",
  pcr_all %>% dplyr::filter(Tissue == "Liver"),
  pcr_models
)
p_pcr_liv
```

## 3.4 Slope Summary Table — PCR

```{r pcr-slope-table}
pcr_params %>%
  arrange(Tissue, Gene) %>%
  mutate(across(c(estimate, conf.low, conf.high), ~ round(.x, 4)),
         p.value = format.pval(p.value, digits = 3, eps = 0.001)) %>%
  knitr::kable(
    caption   = "LMM slope estimates for PCR endpoints. Beta = change in log2(fold change) per 1-unit increase in log10(dose CFU/mL). Random effect: (1|Sample).",
    col.names = c("Gene", "Tissue", "Beta", "95% CI low",
                  "95% CI high", "P-value", "Sig.")
  )
```

---

# 4. Cross-Endpoint Slope Comparison

```{r slope-forest, fig.width = 6.5, fig.height = 4.5, dpi = 300}

category_colors <- c("Duodenum" = "#E25A2E",
                     "Colon"    = "#3B9AB2",
                     "Liver"    = "#78B455")

pcr_forest_df <- pcr_params %>%
  transmute(
    Endpoint = paste0(Gene, " (", Tissue, ")"),
    Slope    = estimate,
    CI_lo    = conf.low,
    CI_hi    = conf.high,
    P        = p.value,
    sig_fill = p.value < 0.05,
    Category = as.character(Tissue)
  ) %>%
  mutate(Endpoint = factor(Endpoint, levels = Endpoint[order(Slope)]))

p_forest <- ggplot(pcr_forest_df,
                   aes(x = Slope, y = Endpoint,
                       color = Category, shape = sig_fill)) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "grey50", linewidth = 0.6) +
  geom_errorbarh(aes(xmin = CI_lo, xmax = CI_hi),
                 height = 0.3, linewidth = 0.8) +
  geom_point(size = 3.2) +
  scale_color_manual(
    values = category_colors,
    name   = "Tissue",
    guide  = guide_legend(override.aes = list(shape = 19, size = 3))
  ) +
  scale_shape_manual(
    values = c("TRUE" = 19, "FALSE" = 1),
    labels = c("TRUE" = "p < 0.05", "FALSE" = "p \u2265 0.05"),
    name   = NULL,
    guide  = guide_legend(override.aes = list(color = "grey40", size = 3))
  ) +
  labs(
    x       = expression(beta~"(log"[2]*"FC per log"[10]*" CFU/mL)"),
    y       = NULL,
    caption = "Filled = p < 0.05 | Open = p \u2265 0.05 | Error bars = 95% CI"
  ) +
  theme_dr(base_size = 10) +
  theme(
    legend.position  = "right",
    legend.text      = element_text(size = 8),
    legend.title     = element_text(size = 9, face = "bold"),
    legend.key.size  = unit(0.45, "cm"),
    axis.text.y      = element_text(size = 9),
    axis.text.x      = element_text(size = 8),
    axis.title.x     = element_text(size = 9),
    plot.caption     = element_text(size = 7, color = "grey40",
                                    family = "Times New Roman"),
    plot.margin      = margin(4, 6, 4, 4, "pt")
  )

p_forest
```

```{r slope-forest-liver, fig.width = 6.5, fig.height = 4.5, dpi = 300}
category_colors <- c("Duodenum" = "deeppink4",
                     "Colon"    = "mediumorchid4",
                     "Liver"    = "steelblue3")

pcr_forest_liver <- pcr_params %>%
  filter(Tissue == "Liver") %>%
  transmute(
    Endpoint = paste0(Gene, " (", Tissue, ")"),
    Slope    = estimate,
    CI_lo    = conf.low,
    CI_hi    = conf.high,
    P        = p.value,
    sig_fill = p.value < 0.05,
    Category = as.character(Tissue)
  ) %>%
  mutate(Endpoint = factor(Endpoint, levels = Endpoint[order(Slope)]))

p_forest_liver <- ggplot(pcr_forest_liver,
                   aes(x = Slope, y = Endpoint,
                       color = Category, shape = sig_fill)) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "grey50", linewidth = 0.6) +
  geom_errorbarh(aes(xmin = CI_lo, xmax = CI_hi),
                 height = 0.3, linewidth = 0.8) +
  geom_point(size = 3.2) +
  scale_color_manual(
    values = category_colors,
    name   = "Tissue",
    guide  = guide_legend(override.aes = list(shape = 19, size = 3))
  ) +
  scale_shape_manual(
    values = c("TRUE" = 19, "FALSE" = 1),
    labels = c("TRUE" = "p < 0.05", "FALSE" = "p \u2265 0.05"),
    name   = NULL,
    guide  = guide_legend(override.aes = list(color = "grey40", size = 3))
  ) +
  labs(
    x       = expression(beta~"(log"[2]*"FC per log"[10]*" CFU/mL)"),
    y       = NULL,
    caption = "Filled = p < 0.05 | Open = p \u2265 0.05 | Error bars = 95% CI"
  ) +
  theme_dr(base_size = 10) +
  theme(
    legend.position  = "right",
    legend.text      = element_text(size = 8),
    legend.title     = element_text(size = 9, face = "bold"),
    legend.key.size  = unit(0.45, "cm"),
    axis.text.y      = element_text(size = 9),
    axis.text.x      = element_text(size = 8),
    axis.title.x     = element_text(size = 9),
    plot.caption     = element_text(size = 7, color = "grey40",
                                    family = "Times New Roman"),
    plot.margin      = margin(4, 6, 4, 4, "pt")
  )
p_forest_liver
```

```{r slope-forest-colon, fig.width = 6.5, fig.height = 4.5, dpi = 300}
pcr_forest_colon <- pcr_params %>%
  filter(Tissue == "Colon") %>%
  transmute(
    Endpoint = paste0(Gene, " (", Tissue, ")"),
    Slope    = estimate,
    CI_lo    = conf.low,
    CI_hi    = conf.high,
    P        = p.value,
    sig_fill = p.value < 0.05,
    Category = as.character(Tissue)
  ) %>%
  mutate(Endpoint = factor(Endpoint, levels = Endpoint[order(Slope)]))

p_forest_colon <- ggplot(pcr_forest_colon,
                   aes(x = Slope, y = Endpoint,
                       color = Category, shape = sig_fill)) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "grey50", linewidth = 0.6) +
  geom_errorbarh(aes(xmin = CI_lo, xmax = CI_hi),
                 height = 0.3, linewidth = 0.8) +
  geom_point(size = 3.2) +
  scale_color_manual(
    values = category_colors,
    name   = "Tissue",
    guide  = guide_legend(override.aes = list(shape = 19, size = 3))
  ) +
  scale_shape_manual(
    values = c("TRUE" = 19, "FALSE" = 1),
    labels = c("TRUE" = "p < 0.05", "FALSE" = "p \u2265 0.05"),
    name   = NULL,
    guide  = guide_legend(override.aes = list(color = "grey40", size = 3))
  ) +
  labs(
    x       = expression(beta~"(log"[2]*"FC per log"[10]*" CFU/mL)"),
    y       = NULL,
    caption = "Filled = p < 0.05 | Open = p \u2265 0.05 | Error bars = 95% CI"
  ) +
  theme_dr(base_size = 10) +
  theme(
    legend.position  = "right",
    legend.text      = element_text(size = 8),
    legend.title     = element_text(size = 9, face = "bold"),
    legend.key.size  = unit(0.45, "cm"),
    axis.text.y      = element_text(size = 9),
    axis.text.x      = element_text(size = 8),
    axis.title.x     = element_text(size = 9),
    plot.caption     = element_text(size = 7, color = "grey40",
                                    family = "Times New Roman"),
    plot.margin      = margin(4, 6, 4, 4, "pt")
  )
p_forest_colon
```

```{r slope-forest-duodenum, fig.width = 6.5, fig.height = 4.5, dpi = 300}
pcr_forest_duodenum <- pcr_params %>%
  filter(Tissue == "Duodenum") %>%
  transmute(
    Endpoint = paste0(Gene, " (", Tissue, ")"),
    Slope    = estimate,
    CI_lo    = conf.low,
    CI_hi    = conf.high,
    P        = p.value,
    sig_fill = p.value < 0.05,
    Category = as.character(Tissue)
  ) %>%
  mutate(Endpoint = factor(Endpoint, levels = Endpoint[order(Slope)]))

p_forest_duodenum <- ggplot(pcr_forest_duodenum,
                   aes(x = Slope, y = Endpoint,
                       color = Category, shape = sig_fill)) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "grey50", linewidth = 0.6) +
  geom_errorbarh(aes(xmin = CI_lo, xmax = CI_hi),
                 height = 0.3, linewidth = 0.8) +
  geom_point(size = 3.2) +
  scale_color_manual(
    values = category_colors,
    name   = "Tissue",
    guide  = guide_legend(override.aes = list(shape = 19, size = 3))
  ) +
  scale_shape_manual(
    values = c("TRUE" = 19, "FALSE" = 1),
    labels = c("TRUE" = "p < 0.05", "FALSE" = "p \u2265 0.05"),
    name   = NULL,
    guide  = guide_legend(override.aes = list(color = "grey40", size = 3))
  ) +
  labs(
    x       = expression(beta~"(log"[2]*"FC per log"[10]*" CFU/mL)"),
    y       = NULL,
    caption = "Filled = p < 0.05 | Open = p \u2265 0.05 | Error bars = 95% CI"
  ) +
  theme_dr(base_size = 10) +
  theme(
    legend.position  = "right",
    legend.text      = element_text(size = 8),
    legend.title     = element_text(size = 9, face = "bold"),
    legend.key.size  = unit(0.45, "cm"),
    axis.text.y      = element_text(size = 9),
    axis.text.x      = element_text(size = 8),
    axis.title.x     = element_text(size = 9),
    plot.caption     = element_text(size = 7, color = "grey40",
                                    family = "Times New Roman"),
    plot.margin      = margin(4, 6, 4, 4, "pt")
  )
p_forest_duodenum
```


